---
title: "Bayesian Models Built in R's Rethinking Package"
output: html_document
date: "2024-08-20"
name: Charlotte Uden
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Lightning Models


```{r, echo=FALSE}
library(ggplot2)
library(dplyr)
library(rethinking)
library(Metrics)
library(ramify)
library(cowplot)
library(gridExtra)
library(ggpointdensity)
library(viridis)
```

## Chen et al 2021
Begin by parameterizing the CAPE x precipitation models from Chen et al 2021. 


```{r, echo=FALSE}
df <- read.csv("/raid/cuden/data/era5_vaisalaLightning_monthlySummaries_2005-2010_NEclip.csv")[,2:16]
cxp <- df$cxp
y <- df$mean_strike_rate
```

```{r}

```

```{r}
ggplot(data = df, mapping = aes(x = cxp_monthly_mean, y = mean_strike_rate)) +
  geom_pointdensity(adjust = .001, alpha=0.5) +
  scale_color_viridis() + 
  xlab(expression(CAPE~x~Precip~(W~m^-2))) +
  ylab(expression(Lightning~flash~rate~(number~per~km^2~per~month))) + 
  ggtitle("May-August 2005-2010 n=3246")+
  guides(alpha="none", color="none") + 
  theme_minimal() +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) 
```

### Power Law: FR = a(CAPE × Precip)^b
With the parameters a and b derived from the optimization at the log-log scale (that is, a linear regression between logarithmic values of flash rate and CAPE × Precip were performed in practice). Similar to Romps et al.36, this model assumed that the lightning flash rate depends on the product of CAPE and Precip. The coefficient b allows nonlinear impacts to be represented in this form. 

```{r}
x <- df$cxp_monthly_mean
y <- df$mean_strike_rate
# Transform data
log_x <- log(x)
log_y <- log(y)

plot(log_x, log_y)

#fit the model
pl <- nls(log_y ~ a*log_x^b, 
          start = list(a = -1, 
                       b = -1))
summary(pl)

curve.nlslrc = nlsLM(photolrc ~ Am*(1-((1-(Rd/Am))^(1-(PARlrc/LCP)))),
                     start=list(Am=(max(photolrc)-min(photolrc)),
                                Rd=-min(photolrc),
                                LCP= (max(photolrc)-1)),
                     data = curvelrc)

#parameter values
a_pl <- summary(pl)$coefficients[1,1]
b_pl <- summary(pl)$coefficients[2,1]

#predict flash rate
y_pred_pl <- a_pl*log_x^b_pl

plot(x, y_pred_pl)

```

### Power law (linear opt) 
A variation of the power law model, but with the least squares optimization obtained without a logarithmic transformation. 

```{r}
pl_op <- nls(y ~ a*x^b, start = list(a = 1, b = 1))
summary(pl_op)

a_pl_op <- summary(pl_op)$coefficients[1,1]
b_pl_op <- summary(pl_op)$coefficients[2,1]

y_pred_pl_op <- a_pl_op*x^b_pl_op

plot(x, y_pred_pl_op)

```

### Scale: FR = a(CAPE × Precip) 
The original scaling approach proposed by Romps et al. 2014

```{r}
sc <- lm(y ~ x)
summary(sc)

a_sc <- summary(sc)$coefficients[2,1]
b_sc <- summary(sc)$coefficients[1,1]

y_pred_sc <- a_sc*x

plot(x, y_pred_sc)
```

### Linear: FR = a(CAPE × Precip) + b
we slightly modified the scaling approach by allowing a negative intercept: FR = a(CAPE × Precip) + b -that is, positive flash rate was simulated only when CAPE × Precip was greater than a threshold.

```{r}
li <- lm(y ~ x)
summary(li)

a_li <- summary(li)$coefficients[2,1]
b_li <- summary(li)$coefficients[1,1]

y_pred_li <- clip(a_li*x+b_li, 0, NA)

plot(x, y_pred_li)
```

### Non-parametric 
Modelled flash rate value in each bin was derived using the arithmetic means of the observed flash rate in all grid cells associated with the corresponding ranges of CAPE × Precip. Based on the script Chen shared, there are 50 bins. 

```{r}
max <- max(df$cxp_monthly_mean)
min <- min(df$cxp_monthly_mean)
nbins <- 50
range <- max-min
bin_size = range / nbins

cxp_bin_min <- seq(min, max, by=bin_size)

strike_bin_means <- rep(0, length(cxp_bin_min))

for (i in seq(1: length(cxp_bin_min))){
  bin <- cxp_bin_min[i]
  
  cxp <- df %>% 
    filter(between(cxp_monthly_mean, bin, bin+bin_size))
  
  mean <- mean(cxp$mean_strike_rate)
  strike_bin_means[i] <- mean
}

np_table <- as.data.frame(cbind(cxp_bin_min, strike_bin_means))
np_table <- mutate(np_table, cxp_bin_max = cxp_bin_min + bin_size)

#apply the model
y_pred_np <- rep(0, length(x))

for (i in seq(1:length(x))){
  cxp = x[i]
  
  bin <- np_table %>% filter(cxp_bin_min <= cxp, 
                      cxp_bin_max > cxp)
  
  y_pred_i <- bin$strike_bin_means
  y_pred_np[i] <- y_pred_i
}

plot(x, y_pred_np)
```

```{r}
df <- cbind(df, y_pred_pl_op, y_pred_sc, y_pred_li, y_pred_np)
df <- mutate(df, mean = (y_pred_pl_op+y_pred_sc+y_pred_li+y_pred_np)/4)
```

```{r}
p1 <- ggplot(data = df, mapping = aes(x = cxp_monthly_mean, y = mean_strike_rate)) +
  geom_pointdensity(adjust = .001, alpha=0.5) +
  scale_color_viridis() + 
  xlab(expression(CAPE~x~Precip~(W~m^-2))) +
  ylab(expression(Lightning~flash~rate~(number~per~km^2~per~month))) + 
  ggtitle("May-August 2005-2010 n=3246")+
  guides(alpha="none", color="none") + 
  geom_line(data=df, aes(y=y_pred_pl_op), color='orange', linetype = "dashed",linewidth=0.7) + 
  geom_line(data=df, aes(y=y_pred_sc), color='yellowgreen', linetype = "dashed",linewidth=0.7) + 
  geom_line(data=df, aes(y=y_pred_li), color='lightblue', linetype = "dashed",linewidth=0.7) + 
  geom_line(data=df, aes(y=y_pred_np), color='darkblue', linetype = "dashed",linewidth=0.7) + 
  geom_line(data=df, aes(y=mean), color='black',linewidth=0.7) +
  theme_minimal() +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) 

legend_colors <- c("Power law (linear opt.)" = "orange", "Scale" = "yellowgreen", "Linear" = "lightblue", "Non-parametric" ="darkblue", "Mean"= "black")

legend_plot <- ggplot(df, aes(x = cxp_monthly_mean)) +
  geom_line(aes(y = y_pred_pl_op, color = 'Power law (linear opt.)'), linewidth = 0.7, linetype = "dashed") +
  geom_line(aes(y = y_pred_sc, color = 'Scale'), linewidth = 0.7, linetype = "dashed") +
  geom_line(aes(y = y_pred_li, color = 'Linear'), linewidth = 0.7, linetype = "dashed") +
  geom_line(aes(y = y_pred_np, color = 'Non-parametric'), linewidth = 0.7, linetype = "dashed") +
  geom_line(aes(y = mean, color = 'Mean'), linewidth = 0.7, linetype = "dashed") +
  labs(color = "Models") +
  scale_color_manual(values = legend_colors) +
  theme_minimal() 

cowplot::get_legend(legend_plot)
legend <- get_legend(legend_plot)
#grid.draw(legend)  

p2 <- ggplot(data=df, aes(x=mean_strike_rate, y=after_stat(density))) +
  coord_flip() + 
  scale_y_reverse() +
  geom_density(color="#6DBCC3", linetype = "dashed") +
  geom_histogram(aes(y = ..density..), color = "white", fill = alpha("#6DBCC3", .2)) + 
  theme_minimal() +
  guides(y="none", y.sec=guide_axis(title = expression(Lightning~flash~rate~(number~per~km^2~per~month)))) +
  scale_x_continuous(breaks=seq(0, 2, by=1)) +
  xlab("")

p3 <- ggplot(data=df, aes(x=cxp_monthly_mean, y=after_stat(density))) +
  geom_density(color="#6DBCC3", linetype = "dashed") +
  geom_histogram(aes(y = ..density..), color = "white", fill = alpha("#6DBCC3", .2)) + 
  theme_minimal() +
  xlab(expression(CAPE~x~Precip~(W~m^-2))) +
  scale_x_continuous(breaks=seq(0, 0.015, by=0.005)) 

blankPlot <- ggplot()+geom_blank(aes(1,1))+
  theme(
    plot.background = element_blank(), 
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(), 
    panel.border = element_blank(),
    panel.background = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.text.x = element_blank(), 
    axis.text.y = element_blank(),
    axis.ticks = element_blank(),
    axis.line = element_blank()
  )

grid.arrange(p1, p2, p3, cowplot::get_legend(legend_plot), 
             ncol=2, nrow=2, widths=c(4, 1), heights=c(4, 1))

```

```{r, echo=FALSE}
#Root Mean Squared Error - lower values are better
rmse_pl_op <- rmse(y, y_pred_pl_op) #best score
rmse_sc <- rmse(y, y_pred_sc)
rmse_li <- rmse(y, y_pred_li)
rmse_np <- rmse(y, y_pred_np)
rmse_table <- as.data.frame(cbind(rmse_pl_op, rmse_sc, rmse_li, rmse_np))
print(rmse_table)
```

### Rethiking 

The Paleo climate data used to drive LPJ-GUESS does not include convective available potential energy. 
Goal: to build a lightning prediction model that uses climate variables included in the PalEON dataset.
Compare model performance with that of models from Chen et al. 2021 

```{r, echo=FALSE}
dat <- list(
  S = df$mean_strike_rate,
  T = df$t2m_monthly_mean)

m1 <- ulam(
  alist(
    S ~ dnorm(mu, sigma), 
    mu <- a + bT*T,
    a ~ dnorm(0, 10), 
    bT ~ dnorm(0, 10), 
    sigma ~ dexp(1)
  ), data=dat, chains=2)

precis(m1)
traceplot(m1)
```


```{r}
post <- extract.samples( m1 , n=1e4 )
precis(post)
```

```{r}
plot( S ~ T , data=dat , col=rangi2 )
#find mean posterior prediction for each parameter
aT_map <- mean(post$a)
bT_map <- mean(post$bT)
sigma_map <- mean(post$sigma) #don't use sigma for the line, only to sample from a normal distribution if we were to apply the model
curve( aT_map + bT_map*(x) , add=TRUE )

y_pred_m1 <- aT_map + bT_map*dat$T
```
### Short Wave Radiation

```{r, echo=FALSE}

dat2 <- list(
  S = df$mean_strike_rate,
  R = df$msdwswrf_monthly_mean)

m2 <- ulam(
  alist(
    S ~ dnorm(mu, sigma), 
    mu <- a + bR*R,
    a ~ dnorm(0, 10), 
    bR ~ dnorm(0, 10), 
    sigma ~ dexp(1)
  ), data=dat2, chains=2)

precis(m2)
traceplot(m2)
```

```{r, echo=FLASE}
post2 <- extract.samples( m2 , n=1e4 )

plot( S ~ R , data=dat2 , col=rangi2 )
#find mean posterior prediction for each parameter
aR_map <- mean(post2$a)
bR_map <- mean(post2$bR)
sigma_map <- mean(post$sigma) #don't use sigma for the line, only to sample from a normal distribution if we were to apply the model
curve( aR_map + bR_map*(x) , add=TRUE )

y_pred_m2 <- aR_map + bR_map*dat2$R
```

### Linear Model from Chen 2021 in Ulam
FR = a(CAPE × Precip) + b

```{r, echo=FLASE}
dat3 <- list(
  S = df$mean_strike_rate,
  C = df$cxp_monthly_mean)

m3 <- ulam(
  alist(
    S ~ dnorm(mu, sigma), 
    mu <- a + bC*C,
    a ~ dnorm(0, 10), 
    bC ~ dnorm(0, 10), 
    sigma ~ dexp(1)
  ), data=dat3, chains=2)

precis(m3)
traceplot(m3)
```


```{r, echo=FLASE}
post3 <- extract.samples( m3 , n=1e4 )

plot( S ~ C , data=dat3 , col=rangi2 )
#find mean posterior prediction for each parameter
aC_map <- mean(post3$a)
bC_map <- mean(post3$bC)
curve( aC_map + bC_map*(x) , add=TRUE )

y_pred_m3 <- aC_map + bC_map*dat3$C
```
```{r}
dat4 <- list(
  S = df$mean_strike_rate,
  P = df$mtpr_monthly_mean)

m4 <- ulam(
  alist(
    S ~ dnorm(mu, sigma), 
    mu <- a + bP*P,
    a ~ dnorm(0, 10), 
    bP ~ dnorm(0, 10), 
    sigma ~ dexp(1)
  ), data=dat4, chains=2)

precis(m4)
traceplot(m4)

post4 <- extract.samples( m4 , n=1e4 )

plot( S ~ P , data=dat4 , col=rangi2 )
#find mean posterior prediction for each parameter
aP_map <- mean(post4$a)
bP_map <- mean(post4$bP)
curve( aP_map + bP_map*(x) , add=TRUE )

y_pred_m4 <- aP_map + bP_map*dat4$P
```

```{r}
dat5 <- list(
  S = df$mean_strike_rate,
  W = df$i10fg_monthly_mean)

m5 <- ulam(
  alist(
    S ~ dnorm(mu, sigma), 
    mu <- a + bW*W,
    a ~ dnorm(0, 10), 
    bW ~ dnorm(0, 10), 
    sigma ~ dexp(1)
  ), data=dat5, chains=2)

precis(m5)
traceplot(m5)

post5 <- extract.samples( m5 , n=1e4 )

plot( S ~ W , data=dat5 , col=rangi2 )
#find mean posterior prediction for each parameter
aW_map <- mean(post5$a)
bW_map <- mean(post5$bW)
curve( aW_map + bW_map*(x) , add=TRUE )

y_pred_m5 <- aW_map + bW_map*dat5$W
```
```{r}
dat6 <- list(
  S = df$mean_strike_rate,
  SP = df$sp_monthly_mean)

m6 <- ulam(
  alist(
    S ~ dnorm(mu, sigma), 
    mu <- a + bSP*SP,
    a ~ dnorm(0, 10), 
    bSP ~ dnorm(0, 10), 
    sigma ~ dexp(1)
  ), data=dat6, chains=2)

precis(m6)
traceplot(m6)

post6 <- extract.samples( m6 , n=1e4 )

plot( S ~ SP , data=dat6 , col=rangi2 )
#find mean posterior prediction for each parameter
aSP_map <- mean(post6$a)
bSP_map <- mean(post6$bSP)
curve( aSP_map + bSP_map*(x) , add=TRUE )

y_pred_m6 <- aSP_map + bSP_map*dat6$SP
```

```{r}
dat7 <- list(
  S = df$mean_strike_rate,
  H = df$rh_monthly_mean)

m7 <- ulam(
  alist(
    S ~ dnorm(mu, sigma), 
    mu <- a + bH*H,
    a ~ dnorm(0, 10), 
    bH ~ dnorm(0, 10), 
    sigma ~ dexp(1)
  ), data=dat7, chains=2)

precis(m7)
traceplot(m7)

post7 <- extract.samples( m7 , n=1e4 )

plot( S ~ H , data=dat7 , col=rangi2 )
#find mean posterior prediction for each parameter
aH_map <- mean(post7$a)
bH_map <- mean(post7$bH)
curve( aH_map + bH_map*(x) , add=TRUE )

y_pred_m7 <- aH_map + bH_map*dat7$H
```

```{r, echo=FALSE}

rmse_pl_op <- rmse(y, y_pred_pl_op) 
rmse_sc <- rmse(y, y_pred_sc)
rmse_li <- rmse(y, y_pred_li)
rmse_np <- rmse(y, y_pred_np)

rmse_m1 <- rmse(y, y_pred_m1)
rmse_m2 <- rmse(y, y_pred_m2)
rmse_m3 <- rmse(y, y_pred_m3)
rmse_m4 <- rmse(y, y_pred_m4)
rmse_m5 <- rmse(y, y_pred_m5)
rmse_m6 <- rmse(y, y_pred_m6)
rmse_m7 <- rmse(y, y_pred_m7)

rmse_table <- as.data.frame(cbind(rmse_pl_op, rmse_sc, rmse_li, rmse_np, rmse_m1, rmse_m2, rmse_m3, rmse_m4, rmse_m5, rmse_m6, rmse_m7))
print(rmse_table)

```
Calculate WAIC
```{r}
#need the log-likelihood of each observation i at each sample s from the posterior:
n_samples <- 1000
logprob <- sapply( 1:n_samples ,
                   function(s) {
                     mu <- post$a[s] + post$bT[s]*dat$T
                     dnorm( dat$S , mu , post$sigma[s] , log=TRUE )
                     } )
#compute lppd (bayesian deviance) average samples in each row, take log liklihood, add all logs together. do this on the log scale
n_cases <- nrow(df)
lppd <- sapply( 1:n_cases , function(i) log_sum_exp(logprob[i,]) - log(n_samples) )
#lppd: 
sum(lppd)

#pWAIC: 
#compute variance across samples for each observation and add together
pWAIC <- sapply( 1:n_cases , function(i) var(logprob[i,]) )
sum(pWAIC)

#WAIC
-2*( sum(lppd) - sum(pWAIC) )

#turn into function

waic <- function(data, mu, post){
  n_samples <- 1000
  logprob <- sapply( 1:n_samples ,
                   function(s) {
                    mu = eval(parse(text = mu))
                    dnorm( data$S, mu , post$sigma[s] , log=TRUE )
                     } )
  #compute lppd (bayesian deviance) average samples in each row, take log liklihood, add all logs together. do this on the log scale
  n_cases <- length(data[[1]])
  lppd <- sapply( 1:n_cases , function(i) log_sum_exp(logprob[i,]) - log(n_samples) )
  
  #pWAIC: 
  #compute variance across samples for each observation and add together
  pWAIC <- sapply( 1:n_cases , function(i) var(logprob[i,]) )

  #WAIC
  WAIC <- -2*( sum(lppd) - sum(pWAIC) )
  
  #calculate standard error
  waic_vec <- -2*( lppd - pWAIC )
  standardError <-sqrt( n_cases*var(waic_vec) )
  
  return(c(WAIC=WAIC, standardError=standardError))
}
mu <- "post$a[s] + post$bT[s]*data$T"
WAIC_m1 <- waic(data=dat, mu, post)

mu <- "post$a[s] + post$bR[s]*data$R"
WAIC_m2 <- waic(data=dat2, mu, post2)

mu <- "post$a[s] + post$bC[s]*data$C"
WAIC_m3 <- waic(data=dat3, mu, post=post3)

mu <- "post$a[s] + post$bP[s]*data$P"
WAIC_m4 <- waic(dat4, mu, post4)

mu <- "post$a[s] + post$bW[s]*data$W"
WAIC_m5 <- waic(dat5, mu, post5)

mu <- "post$a[s] + post$bSP[s]*data$SP"
WAIC_m6 <- waic(dat6, mu, post6)

mu <- "post$a[s] + post$bH[s]*data$H"
WAIC_m7 <- waic(dat7, mu, post7)

WAIC_table <- as.data.frame(rbind(m1=WAIC_m1, m2=WAIC_m2, m3=WAIC_m3, m4=WAIC_m4, m5=WAIC_m5, m6=WAIC_m6, m7=WAIC_m7))
print(WAIC_table)

```

Of the five variables available in the PalEON dataset, short wave radiation (m2) has lowest RMSE and WAIC score, indicating good fit and predictiv permormance. However, cape x precipitation (m3) performed best. 

```{r, echo=FLASE}
ggplot() + geom_point(data = df, aes(x=msdwswrf_monthly_mean, y=mean_strike_rate, col=t2m_monthly_mean))
```


```{r, echo=FLASE}
ggplot() + geom_point(data = df, aes(x=msdwswrf_monthly_mean, y=t2m_monthly_mean, col = mean_strike_rate))
ggplot() + geom_point(data = df, aes(x=msdwswrf_monthly_mean, y=mtpr_monthly_mean, col = mean_strike_rate))
ggplot() + geom_point(data = df, aes(x=msdwswrf_monthly_mean, col=mtpr_monthly_mean, y = mean_strike_rate))
```
```{r}
ggplot() + geom_point(data = df, aes(x=rxp,y = mean_strike_rate))
```

```{r, echo=FLASE}
dat8 <- list(
  S = df$mean_strike_rate,
  P = df$mtpr_monthly_mean, 
  R = df$msdwswrf_monthly_mean)

m8 <- ulam(
  alist(
    S ~ dnorm(mu, sigma), 
    mu <- a + bP*P + bR*R,
    a ~ dnorm(0, 10), 
    bP ~ dnorm(0, 10), 
    bR ~ dnorm(0, 10),
    sigma ~ dexp(1)
  ), data=dat8, chains=2)

precis(m8)
traceplot(m8)

post8 <- extract.samples( m8 , n=1e4 )

plot( S ~ R , data=dat8 , col=P )
#find mean posterior prediction for each parameter
a_map <- mean(post8$a)
bP_map <- mean(post8$bP)
bR_map <- mean(post8$bR)
mu <-  a_map + bP_map*dat8$P + bR_map*dat8$R
y_pred_m8 <- dnorm( dat8$S, mu , post8$sigma)
#curve( a_map + bP_map*(x) + bR_map*(x) , add=TRUE )
plot(dat8$R, y_pred_m8)
y_pred_m8 <- a_map + bP_map*dat8$P + bR_map*dat8$R
plot(dat8$R, y_pred_m8)

mu <- "post$a[s] + post$bP[s]*data$P+ post$bR[s]*data$R"
WAIC_m8 <- waic(dat8, mu, post8)
WAIC_m8
```

```{r, echo=FLASE}









```
