---
title: "Lightning Models"
output: html_document
date: "2024-08-31"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, echo=FALSE}
library(ggplot2)
library(dplyr)
library(rethinking)
library(Metrics)
library(ramify)
library(cowplot)
library(gridExtra)
library(ggpointdensity)
library(viridis)
library(stats)
```


Begin by standardizing all variables: Z = (X − µ)/σ

```{r, echo=FALSE}
data <- read.csv("/raid/cuden/data/era5_vaisalaLightning_monthlySummaries_2005-2010_NEclip.csv")[,2:16]
data <- data %>%
  mutate(strikes = (mean_strike_rate - mean(mean_strike_rate))/sd(mean_strike_rate)) %>% #standardized 
  mutate(strikes_norm = (mean_strike_rate-min(mean_strike_rate))/(max(mean_strike_rate)-min(mean_strike_rate))) %>% #normalized 
  mutate(cape = (cape_monthly_mean - mean(cape_monthly_mean))/sd(cape_monthly_mean)) %>%
  mutate(precip = (mtpr_monthly_mean - mean(mtpr_monthly_mean))/sd(mtpr_monthly_mean)) %>%
  mutate(cxp = (cxp_monthly_mean - mean(cxp_monthly_mean))/sd(cxp_monthly_mean)) %>%
  mutate(tair = (t2m_monthly_mean - mean(t2m_monthly_mean))/sd(t2m_monthly_mean)) %>%
  mutate(wind = (i10fg_monthly_mean - mean(i10fg_monthly_mean))/sd(i10fg_monthly_mean)) %>%
  mutate(swr = (msdwswrf_monthly_mean - mean(msdwswrf_monthly_mean))/sd(msdwswrf_monthly_mean)) %>%
  mutate(sp = (sp_monthly_mean - mean(sp_monthly_mean))/sd(sp_monthly_mean)) %>%
  mutate(rh = (rh_monthly_mean - mean(rh_monthly_mean))/sd(rh_monthly_mean)) 


  
df <- data[,c("strikes", "strikes_norm", "cape", "precip", "cxp", "tair", "wind", "swr", "sp", "rh")]
head(df)

plot(df)
```

Random Forest for measures of variable importance

Mean Decrease Accuracy (%IncMSE): 
 - remove variables from the model and asses which have the greatest impact on model accuracy.
 - Higher the value of mean decrease accuracy or mean decrease gini score , 
       higher the importance of the variable in the model. 
 - tells you effect of randomly shuffling values of a variable on increasing mse of model predictions
 - a higher number indicates that the variable is more important

 Mean Decrease Gini or Increase in Node Purity (IncNodePurity):
 - expresses the change in the homogeneity of the of the groups created by the trees 
     (using the Gini coefficient as a measure). 
 - What is expressed is the decrease in said purity if a particular variable 
      has no information. If a variable has no information to begin with, 
      the decrase would be zero. (from: https://www.reagro.org/methods/statistical/randomforest.html)


```{r}
#apply random forest:
rf <-randomForest(strikes~.,data=df, ntree=500) 
print(rf)

#mtry: Number of random variables used in each tree
# - reducing mtry reduces correlation and strength. 
# - the oob error rate tells you what value of mtry is optimal

mtry <- tuneRF(df[-1],df$strikes, ntreeTry=500,
               stepFactor=1.5,improve=0.01, trace=TRUE, plot=TRUE)
best.m <- mtry[mtry[, 2] == min(mtry[, 2]), 1]
#looking for mtry for the lowest oob error rate 
print(mtry)
print(best.m) 

#apply best mtry to model
rf <-randomForest(strikes~.,data=df, mtry=best.m, importance=TRUE,ntree=500)
print(rf)

#Evaluate variable importance
importance(rf)
varImpPlot(rf)
```
# Using Stan to reparameterize CAPE x Precipitation models from Chen et al. 2021 

```{r}
ggplot(data = df, mapping = aes(x = cxp, y = strikes)) +
  geom_pointdensity() +
  scale_color_viridis() + 
  xlab(expression(Standardized~CAPE~x~Precip~(W~m^-2))) +
  ylab(expression(Standardized~Lightning~flash~rate~(number~per~km^2~per~month))) + 
  ggtitle("May-August 2005-2010 n=3246")+
  guides(alpha="none", color="none") + 
  theme_minimal()

```


### Power Law: FR = a(CAPE × Precip)^b
With the parameters a and b derived from the optimization at the log-log scale (that is, a linear regression between logarithmic values of flash rate and CAPE × Precip were performed in practice). Similar to Romps et al.36, this model assumed that the lightning flash rate depends on the product of CAPE and Precip. The coefficient b allows nonlinear impacts to be represented in this form. 

---could not get this model to work---

### Power law (linear opt) 
A variation of the power law model, but with the least squares optimization obtained without a logarithmic transformation. 

```{r}
dat <- list(
  S = df$strikes,
  CAPE = df$cape)

m1 <- ulam(
  alist(
    S ~ dnorm(mu, sigma), 
    mu <- a * pow(CAPE, b),
    a ~ dnorm(0, 5), 
    b ~ dnorm(0, 5), 
    sigma ~ dexp(1)
  ), data=dat, chains=2)

precis(m1)
traceplot(m1)

post <- extract.samples( m1 , n=1e4 )
precis(post)

plot( S ~ CAPE , data=dat , col=rangi2 )

#find mean posterior prediction for each parameter
a_map <- mean(post$a)
b_map <- mean(post$b)
sigma_map <- mean(post$sigma) #don't use sigma for the line, only to sample from a normal distribution if we were to apply the model

plot( S ~ CAPE , data=dat , col=rangi2 )
a_map <- 1
b_map <- 2
curve( a_map*((x)^b_map) , add=TRUE )

y_pred_mu <- a_map + dat$CAPE^b_map
y_pred_m1 <- rnorm(dat$CAPE, y_pred_mu, sigma_map)

plot(dat$CAPE, y_pred_m1)
```

###Linear: FR = a(CAPE × Precip) + b

```{r}
dat <- list(
  S = df$strikes,
  CXP = df$cxp)

m2 <- ulam(
  alist(
    S ~ dnorm(mu, sigma), 
    mu <- a * CXP + b,
    a ~ dnorm(0, 5), 
    b ~ dnorm(0, 5), 
    sigma ~ dexp(1)
  ), data=dat, chains=2)

precis(m2)
traceplot(m2)

post <- extract.samples( m2 , n=1e4 )
precis(post)

#find mean posterior prediction for each parameter
a_map <- mean(post$a)
b_map <- mean(post$b)
sigma_map <- mean(post$sigma) #don't use sigma for the line, only to sample from a normal distribution if we were to apply the model
#use posterior prediction to calculate mean (mu) strike rate from CAPE
y_pred_mu <- a_map * dat$CXP + b_map
#use mu and sigma to sample from normal distribution
y_pred_m1 <- rnorm(dat$CXP, y_pred_mu, sigma_map)

d <- as.data.frame(cbind(S=dat$S, CXP=dat$CXP, y_pred_mu, y_pred_m1))

ggplot(data=d, aes(x=CXP, y=S)) + geom_point(aes(alpha=0.1)) + 
  geom_line(aes(y=y_pred_mu)) +
  geom_pointdensity() +
  scale_color_viridis() 

ggplot(data=d, aes(x=CXP, y=y_pred_m1)) + geom_point(aes(alpha=0.1)) + 
  geom_line(aes(y=y_pred_mu)) +
  geom_pointdensity() +
  scale_color_viridis() 

rmse_m2 <- rmse(d$S, d$y_pred_m1)
```


```{r}
m2.2 <- lm(strikes ~ cxp, data=df)
summary(m2.2)

a <- summary(m2.2)$coefficients[2,1]
b <- summary(m2.2)$coefficients[1,1]

y_pred_m2.2 <- a*df$cxp + b

d <- as.data.frame(cbind(S=df$strikes, CXP=df$cxp, y_pred_m2.2))

ggplot(data=d, aes(x=CXP, y=S)) + geom_point(aes(alpha=0.1)) + 
  geom_line(aes(y=y_pred_m2.2)) +
  geom_pointdensity() +
  scale_color_viridis() 

rmse_m2.2 <- rmse(d$S, d$y_pred_m2.2)
```

###WAIC and Standard Error Calculations
WAIC: Measure of predictive power. lower values are better. This functioin will only work on models built in stan using the rethinking package. 


```{r}
waic <- function(data, mu, sigma, post, include_sigma){
  
  n_samples <- 1000
  
  if (include_sigma == TRUE) {
    logprob <- sapply( 1:n_samples ,
                   function(s) {
                    mu = eval(parse(text = mu))
                    sigma = eval(parse(text = sigma))
                    dnorm( data$S, mu , sigma , log=TRUE )
                     } )
  } else {
    logprob <- sapply( 1:n_samples ,
                   function(s) {
                    mu = eval(parse(text = mu))
                    dnorm( data$S, mu , post$sigma[s] , log=TRUE )
                     } )
  }

  #compute lppd (bayesian deviance) average samples in each row, take log liklihood, add all logs together. do this on the log scale
  n_cases <- length(data[[1]])
  lppd <- sapply( 1:n_cases , function(i) log_sum_exp(logprob[i,]) - log(n_samples) )
  
  #pWAIC: 
  #compute variance across samples for each observation and add together
  pWAIC <- sapply( 1:n_cases , function(i) var(logprob[i,]) )

  #WAIC
  WAIC <- -2*( sum(lppd) - sum(pWAIC) )
  
  #calculate standard error
  waic_vec <- -2*( lppd - pWAIC )
  standardError <-sqrt( n_cases*var(waic_vec) )
  
  return(c(WAIC=WAIC, standardError=standardError))
}

mu <- "post$a[s]*data$CXP + post$b[s]"
WAIC_m2 <- waic(data=dat, mu, sigma=NA, post, include_sigma=FALSE)
```

Since the random forest identified CAPE as the most important variable, let's try applying a stan model to a model that predicts lightning strike rate from cape. 

```{r, echo=FALSE}
dat <- list(
  S = df$strikes,
  CAPE = df$cape)

m3 <- ulam(
  alist(
    S ~ dnorm(mu, sigma), 
    mu <- a + b*CAPE,
    a ~ dnorm(0, 10), 
    b ~ dnorm(0, 10), 
    sigma ~ dexp(1)
  ), data=dat, chains=2)

precis(m3)
traceplot(m3)

post <- extract.samples( m3 , n=1e4 )
precis(post)
```

Make predictions from posterior:
```{r}
a_map <- mean(post$a)
b_map <- mean(post$b)
sigma_map <- mean(post$sigma) 

y_pred_mu <- a_map * dat$CAPE + b_map
y_pred_m3 <- rnorm(dat$CAPE, y_pred_mu, sigma_map)

d <- as.data.frame(cbind(S=dat$S, CAPE=dat$CAPE, y_pred_mu, y_pred_m3))

ggplot(data=d, aes(x=CAPE, y=S)) + geom_point(aes(alpha=0.1)) + 
  geom_line(aes(y=y_pred_mu)) +
  geom_pointdensity() +
  scale_color_viridis() 

ggplot(data=d, aes(x=CAPE, y=y_pred_m3)) + geom_point(aes(alpha=0.1)) + 
  geom_line(aes(y=y_pred_mu)) +
  geom_pointdensity() +
  scale_color_viridis() 

rmse_m3 <- rmse(d$S, d$y_pred_m3)

mu <- "post$a[s] + post$b[s]*data$CAPE"
WAIC_m3 <- waic(data=dat, mu, post)

```

This looks awful! Mu does not appear to change at all with CAPE! Let's see if this is true - does the observed mean lightning rate change with CAPE?

```{r}
#function to calculate mean flash rate at binned intervals of a variable (x)
calculate_mean_strikes <- function(data, x, n_bins){
  #data=df
  #x="cape"
  #y=strikes
  
  
  bin_size <- (max(data[[x]])-min(data[[x]]))/n_bins

  bins <- seq(min(data[[x]]), max(data[[x]]), bin_size)
  
  strike_means <- rep(0, length(bins))
  
  for (i in seq(1:length(bins))){
    low <- bins[i]
    high <- low+bin_size
    x_i <- data %>% filter(between(get(x, data), low, high))
    if (nrow(x_i) >= 1) {
      mean <- mean(x_i$strikes)
      } else {
        mean <-  NA
        }
    strike_means[i] <- mean
  }
  
  d <- as.data.frame(cbind(bins, strike_means))
  return(d) 
}

d <- calculate_mean_strikes(data=df, x="cape", n_bins=60)

plot(d$bins, d$strike_means)
```

What about standard deviation - how does that change with cape?

```{r}
#function to calculate standard deviation of flash rate at binned intervals of a variable (x)
calculate_sd_strikes <- function(data, x, n_bins){

  bin_size <- (max(data[[x]])-min(data[[x]]))/n_bins

  bins <- seq(min(data[[x]]), max(data[[x]]), bin_size)
  
  strike_sd <- rep(0, length(cape_bins))
  
  for (i in seq(1:length(bins))){
    low <- bins[i]
    high <- low+bin_size
    x_i <- data %>% filter(between(get(x, data), low, high))
    if (nrow(x_i) >= 1) {
      sd <- sd(x_i$strikes)
      } else {
        sd <-  NA
        }
    strike_sd[i] <- sd
  }
  
  d <- as.data.frame(cbind(bins, strike_sd))
  return(d) 
}

d <- calculate_sd_strikes(data=df, x="cape", n_bins=60)

plot(d$bins, d$strike_sd)

```
Let's try to model the change in mu AND sigma with CAPE:

```{r}
dat <- list(
  S = df$strikes,
  CAPE = df$cape)

m3.2 <- ulam(
  alist(
    S ~ dnorm(mu, sigma), 
    mu <- a + b*CAPE,
    sigma <- c + d*CAPE,
    a ~ dnorm(0, 10), 
    b ~ dnorm(0, 10), 
    c ~ dnorm(0, 10), 
    d ~ dnorm(0, 10)
  ), data=dat, chains=2)

precis(m3.2)
traceplot(m3.2)

post <- extract.samples( m3.2 , n=1e4 )
precis(post)

a_map <- mean(post$a)
b_map <- mean(post$b)
c_map <- mean(post$c)
d_map <- mean(post$d)

y_pred_mu <- a_map + b_map*dat$CAPE
y_pred_sigma <- c_map + d_map*dat$CAPE
y_pred_m3.2 <- rnorm(dat$CAPE, y_pred_mu, y_pred_sigma)

d <- as.data.frame(cbind(S=dat$S, CAPE=dat$CAPE, y_pred_mu, y_pred_sigma, y_pred_m3.2))

ggplot(data=d, aes(x=CAPE, y=S)) + geom_point(aes(alpha=0.1)) + 
  geom_line(aes(y=y_pred_mu)) +
  geom_pointdensity() +
  scale_color_viridis() 

ggplot(data=d, aes(x=CAPE, y=S)) + geom_point(aes(alpha=0.1)) + 
  geom_line(aes(y=y_pred_sigma)) +
  geom_pointdensity() +
  scale_color_viridis() 

rmse_m3.2 <- rmse(d$S, d$y_pred_m3.2)

mu <- "post$a[s] + post$b[s]*data$CAPE"
sigma <- "post$c[s] + post$d[s]*data$CAPE"
WAIC_m3.2 <- waic(data=dat, mu, sigma, post, include_sigma = TRUE)
```

Next, let's try short wave radiation, by first looking at changes in the mean and sd flash rate across short wave radiation bins:

```{r}
d <- calculate_mean_strikes(data=df, x="swr", n_bins=60)
d <- cbind(d, strike_sd=calculate_sd_strikes(data=df, x="swr", n_bins=60)$strike_sd)
plot(d$bins, d$strike_means)
plot(d$bins, d$strike_sd)
```
Like before, there will a model that includes changes in mu over short wave radiation (m4), and another that includes changes in mu and sigma over short wave (m4.2) radiation. 
```{r}
dat <- list(
  S = df$strikes,
  SWR = df$swr)

m4 <- ulam(
  alist(
    S ~ dnorm(mu, sigma), 
    mu <- a + b*SWR,
    a ~ dnorm(0, 10), 
    b ~ dnorm(0, 10), 
    sigma ~ dexp(1)
  ), data=dat, chains=2)

precis(m4)
traceplot(m4)

post <- extract.samples( m4 , n=1e4 )
precis(post)

a_map <- mean(post$a)
b_map <- mean(post$b)
sigma_map <- mean(post$sigma)

y_pred_mu <- a_map + b_map*dat$SWR
y_pred_m4 <- rnorm(dat$SWR, y_pred_mu, sigma_map)

d <- as.data.frame(cbind(S=dat$S, SWR=dat$SWR, y_pred_mu, y_pred_m4))

ggplot(data=d, aes(x=SWR, y=S)) + geom_point(aes(alpha=0.1)) + 
  geom_line(aes(y=y_pred_mu)) +
  geom_pointdensity() +
  scale_color_viridis() 

ggplot(data=d, aes(x=SWR, y=y_pred_m4)) + geom_point(aes(alpha=0.1)) + 
  geom_line(aes(y=y_pred_mu)) +
  geom_pointdensity() +
  scale_color_viridis() 

rmse_m4 <- rmse(d$S, d$y_pred_m4)

mu <- "post$a[s] + post$b[s]*data$SWR"
WAIC_m4 <- waic(data=dat, mu, sigma=NA, post, include_sigma = FALSE)
```


```{r}
dat <- list(
  S = df$strikes,
  SWR = df$swr)

m4.2 <- ulam(
  alist(
    S ~ dnorm(mu, sigma), 
    mu <- a + b*SWR,
    sigma <- c + d*SWR,
    a ~ dnorm(0, 10), 
    b ~ dnorm(0, 10), 
    c ~ dnorm(0, 10), 
    d ~ dnorm(0, 10)
  ), data=dat, chains=2)

precis(m4.2)
traceplot(m4.2)

post <- extract.samples( m4.2 , n=1e4 )
precis(post)

a_map <- mean(post$a)
b_map <- mean(post$b)
c_map <- mean(post$c)
d_map <- mean(post$d)

y_pred_mu <- a_map + b_map*dat$SWR
y_pred_sigma <- c_map + d_map*dat$SWR
y_pred_m4.2 <- rnorm(dat$SWR, y_pred_mu, y_pred_sigma)

d <- as.data.frame(cbind(S=dat$S, SWR=dat$SWR, y_pred_mu, y_pred_sigma, y_pred_m4.2))

ggplot(data=d, aes(x=SWR, y=S)) + geom_point(aes(alpha=0.1)) + 
  geom_line(aes(y=y_pred_mu)) +
  geom_pointdensity() +
  scale_color_viridis() 

ggplot(data=d, aes(x=SWR, y=S)) + geom_point(aes(alpha=0.1)) + 
  geom_line(aes(y=y_pred_sigma)) +
  geom_pointdensity() +
  scale_color_viridis() 

ggplot(data=d, aes(x=SWR, y=y_pred_m4.2)) + geom_point(aes(alpha=0.1)) + 
  geom_line(aes(y=y_pred_mu)) +
  geom_pointdensity() +
  scale_color_viridis() 


rmse_m4.2 <- rmse(d$S, d$y_pred_m4.2)

mu <- "post$a[s] + post$b[s]*data$SWR"
sigma <- "post$c[s] + post$d[s]*data$SWR"
WAIC_m4.2 <- waic(data=dat, mu, sigma, post, include_sigma = TRUE)
```

Next, relative humidity:
```{r}
d <- calculate_mean_strikes(data=df, x="rh", n_bins=60)
d <- cbind(d, strike_sd=calculate_sd_strikes(data=df, x="rh", n_bins=60)$strike_sd)
plot(d$bins, d$strike_means)
plot(d$bins, d$strike_sd)
```

```{r}
dat <- list(
  S = df$strikes,
  RH = df$rh)

m5 <- ulam(
  alist(
    S ~ dnorm(mu, sigma), 
    mu <- a + b*RH,
    a ~ dnorm(0, 10), 
    b ~ dnorm(0, 10), 
    sigma ~ dexp(1)
  ), data=dat, chains=2)

precis(m5)
traceplot(m5)

post <- extract.samples( m5 , n=1e4 )
precis(post)

a_map <- mean(post$a)
b_map <- mean(post$b)
sigma_map <- mean(post$sigma)

y_pred_mu <- a_map + b_map*dat$RH
y_pred_m5 <- rnorm(dat$RH, y_pred_mu, sigma_map)

d <- as.data.frame(cbind(S=dat$S, RH=dat$RH, y_pred_mu, y_pred_m5))

ggplot(data=d, aes(x=RH, y=S)) + geom_point(aes(alpha=0.1)) + 
  geom_line(aes(y=y_pred_mu)) +
  geom_pointdensity() +
  scale_color_viridis() 

ggplot(data=d, aes(x=RH, y=y_pred_m5)) + geom_point(aes(alpha=0.1)) + 
  geom_line(aes(y=y_pred_mu)) +
  geom_pointdensity() +
  scale_color_viridis() 

rmse_m5 <- rmse(d$S, d$y_pred_m5)

mu <- "post$a[s] + post$b[s]*data$RH"
WAIC_m5 <- waic(data=dat, mu, sigma=NA, post, include_sigma = FALSE)
```

```{r}
dat <- list(
  S = df$strikes,
  RH = df$rh)

m5.2 <- ulam(
  alist(
    S ~ dnorm(mu, sigma), 
    mu <- a + b*RH,
    sigma <- c + d*RH,
    a ~ dnorm(0, 10), 
    b ~ dnorm(0, 10), 
    c ~ dnorm(0, 10), 
    d ~ dnorm(0, 10)
  ), data=dat, chains=2)

precis(m5.2)
traceplot(m5.2)

post <- extract.samples( m5.2 , n=1e4 )
precis(post)

a_map <- mean(post$a)
b_map <- mean(post$b)
c_map <- mean(post$c)
d_map <- mean(post$d)

y_pred_mu <- a_map + b_map*dat$RH
y_pred_sigma <- c_map + d_map*dat$RH
y_pred_m5.2 <- rnorm(dat$RH, y_pred_mu, y_pred_sigma)

d <- as.data.frame(cbind(S=dat$S, RH=dat$RH, y_pred_mu, y_pred_sigma, y_pred_m5.2))

ggplot(data=d, aes(x=RH, y=S)) + geom_point(aes(alpha=0.1)) + 
  geom_line(aes(y=y_pred_mu)) +
  geom_pointdensity() +
  scale_color_viridis() 

ggplot(data=d, aes(x=RH, y=S)) + geom_point(aes(alpha=0.1)) + 
  geom_line(aes(y=y_pred_sigma)) +
  geom_pointdensity() +
  scale_color_viridis() 

ggplot(data=d, aes(x=RH, y=y_pred_m5.2)) + geom_point(aes(alpha=0.1)) + 
  geom_line(aes(y=y_pred_mu)) +
  geom_pointdensity() +
  scale_color_viridis() 


rmse_m5.2 <- rmse(d$S, d$y_pred_m5.2)

mu <- "post$a[s] + post$b[s]*data$RH"
sigma <- "post$c[s] + post$d[s]*data$RH"
WAIC_m5.2 <- waic(data=dat, mu, sigma, post, include_sigma = TRUE)
```


```{r}
summary1 <- as.data.frame(rbind(WAIC_m1, WAIC_m2, WAIC_m3, WAIC_m3.2, WAIC_m4, WAIC_m4.2, WAIC_m5, WAIC_m5.2))
summary2 <- as.data.frame(rbind(rmse_m1, rmse_m2, rmse_m3, rmse_m3.2, rmse_m4, rmse_m4.2, rmse_m5, rmse_m5.2))

cbind(summary1, summary2)

```

