---
title: "Probability Distributions: Normal"
author: "Charlotte Uden"
date: "2024-11-27"
output: html_document
---

```{r}
library(rethinking)
library(dplyr)
library(stats)
library(caret)
```

```{r}
data <- read.csv("/raid/cuden/data/era5_vaisalaLightning_monthlySummerSummaries_2005-2010_NEclip.csv")[,2:16]
colnames(data) <- c("lat", "lon", "year", "strikes", "cape", "precip", "cxp", "d2m", "tair", "wind", "swr", "sp", "rh", "tmin", "tmax")
hist(data$strikes, breaks = 30)
```

```{r}
## 80% of the sample size
sample_size <- floor(0.8 * nrow(data))

## set the seed to make partition reproducible
set.seed(123)
train_index <- sample(seq_len(nrow(data)), size = sample_size)

train <- data[train_index, ]
test <- data[-train_index, ]

```


Standardize predictor variables 
```{r}
#save min and max (normalization parameters?):
#-----
mu_cxp <- mean(train$cxp)
sigma_cxp <- sd(train$cxp)
mu_cape <- mean(train$cape)
sigma_cape <- sd(train$cape)
mu_precip <- mean(train$precip)
sigma_precip <- sd(train$precip)
mu_tair <- mean(train$tair)
sigma_tair <- sd(train$tair)
mu_wind <- mean(train$wind)
sigma_wind <- sd(train$wind)
mu_swr <- mean(train$swr)
sigma_swr <- sd(train$swr)
mu_sp <- mean(train$sp)
sigma_sp <- sd(train$sp)
mu_rh <- mean(train$rh)
sigma_rh <- sd(train$rh)
#-----

train <- train %>%
  mutate(cape = (cape-mu_cape)/sigma_cape) %>%
  mutate(precip = (precip-mu_precip)/sigma_precip) %>%
  mutate(cxp = (cxp-mu_cxp)/sigma_cxp) %>%
  mutate(tair = (tair-mu_tair)/sigma_tair) %>%
  mutate(wind = (wind-mu_wind)/sigma_wind) %>%
  mutate(swr = (swr-mu_swr)/sigma_swr) %>%
  mutate(sp = (sp-mu_sp)/sigma_sp) %>%
  mutate(rh = (rh-mu_rh)/sigma_rh) 

train <- train[,c("strikes", "cape", "precip", "cxp", "tair", "wind", "swr", "sp", "rh")]
head(train)

plot(train)

#use the min and max values from the train set to normalize the test set 
test <- test %>%
  mutate(cape = (cape-mu_cape)/sigma_cape) %>%
  mutate(precip = (precip-mu_precip)/sigma_precip) %>%
  mutate(cxp = (cxp-mu_cxp)/sigma_cxp) %>%
  mutate(tair = (tair-mu_tair)/sigma_tair) %>%
  mutate(wind = (wind-mu_wind)/sigma_wind) %>%
  mutate(swr = (swr-mu_swr)/sigma_swr) %>%
  mutate(sp = (sp-mu_sp)/sigma_sp) %>%
  mutate(rh = (rh-mu_rh)/sigma_rh) 

test <- test[,c("strikes", "cape", "precip", "cxp", "tair", "wind", "swr", "sp", "rh")]

```

```{r, echo=FALSE}

dat <- list(
  S = train$strikes)

m0 <- ulam(
  alist(
    S ~ dnorm( mu , sigma ),
    mu ~ dnorm(0,10),
    sigma ~ dnorm(0,10)
  ), data=dat , chains = 4, cores = 4, log_lik=TRUE)

```



```{r}
rethinking::traceplot(m0)

summary_table <- precis(  m0 )
plot( coeftab( m0 ) , pars=c("mu","sigma") )


#predictions 

mean_mu <- summary_table["mu", "mean"]
mean_sigma <- summary_table["sigma", "mean"]

simStrikes <- rnorm(length(test$strikes), mean_mu, mean_sigma)

hist(simStrikes, breaks=30, main = "predictions")

hist(test$strikes, breaks=30, main = "observations")

```


### Truncated normal - the normal distribution predicts negative lightning strike rates. I can:
1. truncate the normal distribution so that no negative values are predicted - I dont know how to write  a model like this. and after reading a little, it seems that the gamma distrubution is a better choice for positive. see Rstan_models.Rmd for these models. 
2. use a gamma distribution

#Gamma Distribution
```{r, echo=FALSE}

dat <- list(
  S = train$strikes)

m1 <- ulam(
  alist(
    S ~ dgamma2( k , beta ),
    k ~ dnorm(0,10),
    beta ~ dnorm(0,10)
  ), data=dat , chains = 4, cores = 4, log_lik=TRUE)

```


```{r}

rethinking::traceplot(m1)
summary_table <- precis(  m1 )
plot( coeftab( m1 ) , pars=c("k","beta") )

mean_k <- summary_table["k", "mean"]
mean_beta <- summary_table["beta", "mean"]

simStrikes <- rgamma(length(test$strikes), mean_k, mean_beta)

hist(simStrikes, breaks=30, main = "predictions")
hist(test$strikes, breaks=30, main = "observations")

```

```{r}
compare(m0, m1)
```


```{r, echo=FALSE}

dat <- list(
  S = train$strikes)

m2 <- ulam(
  alist(
    S ~ dgamma2( k , beta ),
    k <- mu^2 / sigma^2,
    beta <- mu / sigma^2,
    mu ~ dnorm(0,5), 
    sigma ~ dnorm(0,5)
  ), data=dat , chains = 4, cores = 4, log_lik=TRUE)

```


```{r}

rethinking::traceplot(m2)
summary_table <- precis(  m2 )
plot( coeftab( m2 ) , pars=c("mu","sigma") )

mean_mu <- summary_table["mu", "mean"]
mean_sigma <- summary_table["sigma", "mean"]

k <- mean_mu^2 / mean_sigma^2
beta <- mean_mu / mean_sigma^2

simStrikes <- rgamma(length(data$strikes), k, beta)

hist(simStrikes, breaks=30, main = "predictions")
hist(test$strikes, breaks=30, main = "observations")

```

```{r}
compare(m0, m1, m2)
```


```{r}

dat <- list(
  S = train$strikes,
  CAPE = train$cape)


m3 <- ulam(alist(
    S ~ dgamma2( k , beta ),
    k <- a_k + b_k * CAPE,
    beta <- a_beta + b_beta * CAPE,
    a_k ~ dnorm(0,10),
    b_k ~ dnorm(0,10),
    a_beta ~ dnorm(0,10),
    b_beta ~ dnorm(0,10)
  ), data=dat, chains = 4, cores = 4, log_lik=TRUE)
```

```{r}

rethinking::traceplot(m3)
summary_table <- precis(  m3 )
plot( coeftab( m3 ) , pars=c("a_mu","b_mu", "a_sigma", "b_sigma") )

#predictions 

a_mu <- summary_table["a_mu", "mean"]
b_mu <- summary_table["b_mu", "mean"]
a_sigma <- summary_table["a_sigma", "mean"]
b_sigma <- summary_table["b_sigma", "mean"]

mu <- a_mu + b_mu * test$cape
sigma <- a_sigma + b_sigma * test$cape

simStrikes <- rnorm(length(test$cape), mu, sigma)

hist(simStrikes, breaks=30)

plot(test$cape, simStrikes, main = "predictions")
plot(test$cape, test$strikes, main = "observations" )
plot(simStrikes, test$strikes)


```


```{r}

dat <- list(
  S = train$strikes,
  CXP = train$cxp)

m4 <- ulam(alist(
    S ~ dnorm( mu , sigma ),
    mu <- a_mu + b_mu * CXP,
    sigma <- a_sigma + b_sigma * CXP,
    a_mu ~ dnorm(0,10),
    b_mu ~ dnorm(0,10),
    a_sigma ~ dnorm(0,10),
    b_sigma ~ dnorm(0,10)
  ), data=dat, chains = 4, cores = 4, log_lik=TRUE)

rethinking::
summary_table <- precis(  m4 )
plot( coeftab( m4 ) , pars=c("a_mu","b_mu", "a_sigma", "b_sigma") )

#predictions 

a_mu <- summary_table["a_mu", "mean"]
b_mu <- summary_table["b_mu", "mean"]
a_sigma <- summary_table["a_sigma", "mean"]
b_sigma <- summary_table["b_sigma", "mean"]

mu <- a_mu + b_mu * test$cxp
sigma <- a_sigma + b_sigma * test$cxp

simStrikes <- rnorm(length(test$cxp), mu, sigma)

hist(simStrikes, breaks=30)

plot(test$cxp, simStrikes, main = "predictions")
plot(test$cxp, test$strikes, main = "observations" )
plot(simStrikes, test$strikes)

```


```{r}
dat <- list(
  S = train$strikes,
  CXP = train$cxp)

m6 <- ulam(alist(
    log(S) ~ dnorm( mu , sigma ),
    mu <- a_mu + b_mu * CXP,
    sigma <- a_sigma + b_sigma * CXP,
    a_mu ~ dnorm(0,10),
    b_mu ~ dnorm(0,10),
    a_sigma ~ dnorm(0,10),
    b_sigma ~ dnorm(0,10)
  ), data=dat, chains = 4, cores = 4, log_lik=TRUE)


summary_table <- precis(  m6 )
plot( coeftab( m6 ) , pars=c("a_mu","b_mu", "a_sigma", "b_sigma") )

#predictions 

a_mu <- summary_table["a_mu", "mean"]
b_mu <- summary_table["b_mu", "mean"]
a_sigma <- summary_table["a_sigma", "mean"]
b_sigma <- summary_table["b_sigma", "mean"]

mu <- a_mu + b_mu * test$cxp
sigma <- a_sigma + b_sigma * test$cxp

#simulate log(S) values
simLogStrikes <- rnorm(length(test$cxp), mu, sigma)
#transform:
simStrikes <- exp(simLogStrikes)

hist(simStrikes, breaks=30)

plot(test$cxp, simStrikes, main = "predictions")
plot(test$cxp, test$strikes, main = "observations" )
plot(simStrikes, test$strikes)
abline(a = 0, b = 1, col = "red", lty = 2, lwd = 2)
```

#How do i make sure predictions are greater than 0?

```{r}

dat <- list(
  S = train$strikes,
  CXP = train$cxp)

m7 <- ulam(alist(
    S ~ dnorm( mu , sigma )  T[0, ], #truncate at 0
    mu <- a_mu + b_mu * CXP,
    sigma <- a_sigma + b_sigma * CXP,
    a_mu ~ dnorm(0,10),
    b_mu ~ dnorm(0,10),
    a_sigma ~ dnorm(0,10),
    b_sigma ~ dnorm(0,10)
  ), data=dat, chains = 4, cores = 4, log_lik=TRUE)


summary_table <- precis(  m7 )
plot( coeftab( m7 ) , pars=c("a_mu","b_mu", "a_sigma", "b_sigma") )

#predictions 

a_mu <- summary_table["a_mu", "mean"]
b_mu <- summary_table["b_mu", "mean"]
a_sigma <- summary_table["a_sigma", "mean"]
b_sigma <- summary_table["b_sigma", "mean"]

mu <- a_mu + b_mu * test$cxp
sigma <- a_sigma + b_sigma * test$cxp

simStrikes <- rnorm(length(test$cxp), mu, sigma)

hist(simStrikes, breaks=30)

plot(test$cxp, simStrikes, main = "predictions")
plot(test$cxp, test$strikes, main = "observations" )
plot(simStrikes, test$strikes)


```



```{r}
compare(m0, m1, m2, m3, m5, m6)
```

```{r}

```

```{r}

```
```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```
```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```
```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```
```{r}

```