---
title: "Modeling Lightning Strike Rate"
author: "Charlotte Uden"
date: "2024-11-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This script uses the rethinking package to build Bayesian models of lighting strike rate from eight climate variables. Two of these variables are derived from information at various levels in the atmosphere:  CAPE (convective available potential energy) and CAPE x Precipitation. These variables have been used before to model lighting strike rate and provide a baseline for comparison with six surface level climate variables: short wave radiation, relative humidity, temperature, surface pressure, wind, and precipitation. The first model tested is a null model, where lighting strike rate is modeled from a normal distribution with no dependence on climate: 

M0:
Lightning Strike Rate ~ dnorm(mu, sigma)
mu ~ dnorm(0, 10)
sigma ~ dexp(1) 

The following eight models (M1 - M8) allow mu and sigma to change with climate. Each climate variable is independently tested:

M1 - M8:
Lightning Strike Rate ~ dnorm(mu, sigma)
mu <- a_mu + b_mu * climate_variable_x
sigma <- a_sigma + b_sigma * climate_variable_x
a_mu ~ dnorm(0, 10)
b_mu ~ dnorm(0, 10)
a_sigma ~ dexp(1) 
b_sigma ~ dexp(1) 

The compare() function is used to compare model performance, specifically WAIC scores. The final ten models (M9 - M17) use the performance of each of these single variable models to explore additive affects and interaction terms. 


```{r, echo=FALSE, include=FALSE}
library(rethinking)
library(dplyr)
library(stats)
library(caret)
```

# Preprocessing

1. Read in the data
```{r, echo=FALSE}
data <- read.csv("/raid/cuden/data/era5_vaisalaLightning_monthlySummerSummaries_2005-2010_NEclip.csv")[,2:16]
colnames(data) <- c("lat", "lon", "year", "strikes", "cape", "precip", "cxp", "d2m", "tair", "wind", "swr", "sp", "rh", "tmin", "tmax")
hist(data$strikes, breaks = 30)
```

2. split data into training and testing
```{r, include=FALSE}
## 80% of the sample size
sample_size <- floor(0.8 * nrow(data))

## set the seed to make partition reproducible
set.seed(123)
train_index <- sample(seq_len(nrow(data)), size = sample_size)

train <- data[train_index, ]
test <- data[-train_index, ]

```


3. Standardize predictor variables 
```{r, echo=FALSE}
#save min and max (normalization parameters?):
#-----
mu_cxp <- mean(train$cxp)
sigma_cxp <- sd(train$cxp)
mu_cape <- mean(train$cape)
sigma_cape <- sd(train$cape)
mu_precip <- mean(train$precip)
sigma_precip <- sd(train$precip)
mu_tair <- mean(train$tair)
sigma_tair <- sd(train$tair)
mu_wind <- mean(train$wind)
sigma_wind <- sd(train$wind)
mu_swr <- mean(train$swr)
sigma_swr <- sd(train$swr)
mu_sp <- mean(train$sp)
sigma_sp <- sd(train$sp)
mu_rh <- mean(train$rh)
sigma_rh <- sd(train$rh)
#-----

train <- train %>%
  mutate(cape = (cape-mu_cape)/sigma_cape) %>%
  mutate(precip = (precip-mu_precip)/sigma_precip) %>%
  mutate(cxp = (cxp-mu_cxp)/sigma_cxp) %>%
  mutate(tair = (tair-mu_tair)/sigma_tair) %>%
  mutate(wind = (wind-mu_wind)/sigma_wind) %>%
  mutate(swr = (swr-mu_swr)/sigma_swr) %>%
  mutate(sp = (sp-mu_sp)/sigma_sp) %>%
  mutate(rh = (rh-mu_rh)/sigma_rh) 

train <- train[,c("strikes", "cape", "precip", "cxp", "tair", "wind", "swr", "sp", "rh")]
head(train)

plot(train)

#use the min and max values from the train set to normalize the test set 
test <- test %>%
  mutate(cape = (cape-mu_cape)/sigma_cape) %>%
  mutate(precip = (precip-mu_precip)/sigma_precip) %>%
  mutate(cxp = (cxp-mu_cxp)/sigma_cxp) %>%
  mutate(tair = (tair-mu_tair)/sigma_tair) %>%
  mutate(wind = (wind-mu_wind)/sigma_wind) %>%
  mutate(swr = (swr-mu_swr)/sigma_swr) %>%
  mutate(sp = (sp-mu_sp)/sigma_sp) %>%
  mutate(rh = (rh-mu_rh)/sigma_rh) 

test <- test[,c("strikes", "cape", "precip", "cxp", "tair", "wind", "swr", "sp", "rh")]

```
# M0: normal model, no dependance on climate

First, model lightning strike rate using the normal distribution with no dependence on climate. Other models will be compared with this model. 

    S ~ dnorm( mu , sigma ),
    mu ~ dnorm(0,10),
    sigma ~ dexp(0,10)

```{r, echo=FALSE, results='hide', include=FALSE}

dat <- list(
  S = train$strikes)

m0 <- ulam(
  alist(
    S ~ dnorm( mu , sigma ),
    mu ~ dnorm(0,10),
    sigma ~ dnorm(0,10)
  ), data=dat , chains = 4, cores = 4, log_lik=TRUE)

```

```{r, echo=FALSE}
rethinking::traceplot(m0)

summary_table <- precis(  m0 )
plot( coeftab( m0 ) , pars=c("mu","sigma") )


#predictions 

mean_mu <- summary_table["mu", "mean"]
mean_sigma <- summary_table["sigma", "mean"]

simStrikes <- rnorm(length(test$strikes), mean_mu, mean_sigma)
simStrikes[simStrikes < 0] <- 0

hist(simStrikes, breaks=30, main = "predictions")

hist(test$strikes, breaks=30, main = "observations")

rmse_m0 <- RMSE(simStrikes, test$strikes)

```

# M1: CAPE

Lightning strike rate modeled from a normal distribution where mu and sigma are dependant on climate: CAPE
    
    S ~ dnorm( mu , sigma ),
    mu <- a_mu + b_mu * CAPE,
    sigma <- a_sigma + b_sigma * CAPE,
    
```{r, echo=FALSE, results='hide', include=FALSE}

dat <- list(
  S = train$strikes,
  CAPE = train$cape)

m1 <- ulam(alist(
    S ~ dnorm( mu , sigma ),
    mu <- a_mu + b_mu * CAPE,
    sigma <- a_sigma + b_sigma * CAPE,
    a_mu ~ dnorm(0,10),
    b_mu ~ dnorm(0,10),
    a_sigma ~ dnorm(0,10),
    b_sigma ~ dnorm(0,10)
  ), data=dat, chains = 4, cores = 4, log_lik=TRUE)
```

```{r, echo=FALSE}
#rethinking::traceplot(m1)
summary_table <- precis(  m1 )
plot( coeftab( m1 ) , pars=c("a_mu","b_mu", "a_sigma", "b_sigma") )

#predictions 

a_mu <- summary_table["a_mu", "mean"]
b_mu <- summary_table["b_mu", "mean"]
a_sigma <- summary_table["a_sigma", "mean"]
b_sigma <- summary_table["b_sigma", "mean"]

mu <- a_mu + b_mu * test$cape
sigma <- a_sigma + b_sigma * test$cape

simStrikes <- rnorm(length(test$cape), mu, sigma)
simStrikes[simStrikes < 0] <- 0

hist(simStrikes, breaks=30)

plot(test$cxp, simStrikes, main = "predictions")
plot(test$cxp, test$strikes, main = "observations" )
plot(simStrikes, test$strikes)

rmse_m1 <- RMSE(simStrikes, test$strikes)

```
# M2: CAPE x Precipitation

Lighning strike rate modeled from a normal distribution where mu and sigma are dependant on climate: CAPE x Precipitation

    S ~ dnorm( mu , sigma ),
    mu <- a_mu + b_mu * CXP,
    sigma <- a_sigma + b_sigma * CXP,
    
```{r, echo=FALSE, results='hide', include=FALSE}

dat <- list(
  S = train$strikes,
  CXP = train$cxp)

m2 <- ulam(alist(
    S ~ dnorm( mu , sigma ),
    mu <- a_mu + b_mu * CXP,
    sigma <- a_sigma + b_sigma * CXP,
    a_mu ~ dnorm(0,10),
    b_mu ~ dnorm(0,10),
    a_sigma ~ dnorm(0,10),
    b_sigma ~ dnorm(0,10)
  ), data=dat, chains = 4, cores = 4, log_lik=TRUE)

```

```{r, echo=FALSE, results='hide', include=FALSE}
#rethinking::traceplot(m2)
summary_table <- precis(  m2 )
plot( coeftab( m2 ) , pars=c("a_mu","b_mu", "a_sigma", "b_sigma") )

#predictions 

a_mu <- summary_table["a_mu", "mean"]
b_mu <- summary_table["b_mu", "mean"]
a_sigma <- summary_table["a_sigma", "mean"]
b_sigma <- summary_table["b_sigma", "mean"]

mu <- a_mu + b_mu * test$cxp
sigma <- a_sigma + b_sigma * test$cxp

simStrikes <- rnorm(length(test$cxp), mu, sigma)
simStrikes[simStrikes < 0] <- 0

hist(simStrikes, breaks=30)

plot(test$cxp, simStrikes, main = "predictions")
plot(test$cxp, test$strikes, main = "observations" )
plot(simStrikes, test$strikes)
abline(a = 0, b = 1, col = "red", lty = 2, lwd = 2)

rmse_m2 <- RMSE(simStrikes, test$strikes)

```
# M3: Relative Humidity

Lighning strike rate modeled from a normal distribution where mu and sigma are dependant on climate: Relative Humidity

    S ~ dnorm( mu , sigma ),
    mu <- a_mu + b_mu * RH,
    sigma <- a_sigma + b_sigma * RH,
    
```{r, echo=FALSE, results='hide', include=FALSE}

dat <- list(
  S = train$strikes,
  RH = train$rh)

m3 <- ulam(alist(
    S ~ dnorm( mu , sigma ),
    mu <- a_mu + b_mu * RH,
    sigma <- a_sigma + b_sigma * RH,
    a_mu ~ dnorm(0,10),
    b_mu ~ dnorm(0,10),
    a_sigma ~ dnorm(0,10),
    b_sigma ~ dnorm(0,10)
  ), data=dat, chains = 4, cores = 4, log_lik=TRUE)

```

```{r, echo=FALSE}
#rethinking::traceplot(m3)
summary_table <- precis(  m3 )
plot( coeftab( m3 ) , pars=c("a_mu","b_mu", "a_sigma", "b_sigma") )

#predictions 

a_mu <- summary_table["a_mu", "mean"]
b_mu <- summary_table["b_mu", "mean"]
a_sigma <- summary_table["a_sigma", "mean"]
b_sigma <- summary_table["b_sigma", "mean"]

mu <- a_mu + b_mu * test$rh
sigma <- a_sigma + b_sigma * test$rh

simStrikes <- rnorm(length(test$rh), mu, sigma)
simStrikes[simStrikes < 0] <- 0

hist(simStrikes, breaks=30)

plot(test$rh, simStrikes, main = "predictions")
plot(test$rh, test$strikes, main = "observations" )
plot(simStrikes, test$strikes)
abline(a = 0, b = 1, col = "red", lty = 2, lwd = 2)

rmse_m3 <- RMSE(simStrikes, test$strikes)
```
# M4: Short Wave Radiation

Lighning strike rate modeled from a normal distribution where mu and sigma are dependant on climate: Short wave radiation

    S ~ dnorm( mu , sigma ),
    mu <- a_mu + b_mu * SWR,
    sigma <- a_sigma + b_sigma * SWR,
    
```{r, echo=FALSE, results='hide', include=FALSE}

dat <- list(
  S = train$strikes,
  SWR = train$swr)

m4 <- ulam(alist(
    S ~ dnorm( mu , sigma ),
    mu <- a_mu + b_mu * SWR,
    sigma <- a_sigma + b_sigma * SWR,
    a_mu ~ dnorm(0,10),
    b_mu ~ dnorm(0,10),
    a_sigma ~ dnorm(0,10),
    b_sigma ~ dnorm(0,10)
  ), data=dat, chains = 4, cores = 4, log_lik=TRUE)

```

```{r, echo=FALSE}
#rethinking::traceplot(m4)
summary_table <- precis(  m4 )
plot( coeftab( m4 ) , pars=c("a_mu","b_mu", "a_sigma", "b_sigma") )

#predictions 

a_mu <- summary_table["a_mu", "mean"]
b_mu <- summary_table["b_mu", "mean"]
a_sigma <- summary_table["a_sigma", "mean"]
b_sigma <- summary_table["b_sigma", "mean"]

mu <- a_mu + b_mu * test$swr
sigma <- a_sigma + b_sigma * test$swr

simStrikes <- rnorm(length(test$swr), mu, sigma)
simStrikes[simStrikes < 0] <- 0

hist(simStrikes, breaks=30)

plot(test$swr, simStrikes, main = "predictions")
plot(test$swr, test$strikes, main = "observations" )
plot(simStrikes, test$strikes)

rmse_m4 <- RMSE(simStrikes, test$strikes)
```
# M5: Temperature

Lightning strike rate modeled from a normal distribution where mu and sigma are dependant on climate: Temperature

    S ~ dnorm( mu , sigma ),
    mu <- a_mu + b_mu * T,
    sigma <- a_sigma + b_sigma * T,
    
```{r, echo=FALSE, results='hide', include=FALSE}

dat <- list(
  S = train$strikes,
  T = train$tair)

m5 <- ulam(alist(
    S ~ dnorm( mu , sigma ),
    mu <- a_mu + b_mu * T,
    sigma <- a_sigma + b_sigma * T,
    a_mu ~ dnorm(0,10),
    b_mu ~ dnorm(0,10),
    a_sigma ~ dnorm(0,10),
    b_sigma ~ dnorm(0,10)
  ), data=dat, chains = 4, cores = 4, log_lik=TRUE)

```

```{r, echo=FALSE}
#rethinking::traceplot(m5)
summary_table <- precis(  m5 )
plot( coeftab( m5 ) , pars=c("a_mu","b_mu", "a_sigma", "b_sigma") )

#predictions 

a_mu <- summary_table["a_mu", "mean"]
b_mu <- summary_table["b_mu", "mean"]
a_sigma <- summary_table["a_sigma", "mean"]
b_sigma <- summary_table["b_sigma", "mean"]

mu <- a_mu + b_mu * test$tair
sigma <- a_sigma + b_sigma * test$tair

simStrikes <- rnorm(length(test$tair), mu, sigma)
simStrikes[simStrikes < 0] <- 0

hist(simStrikes, breaks=30)

plot(test$tair, simStrikes, main = "predictions")
plot(test$tair, test$strikes, main = "observations" )
plot(simStrikes, test$strikes)
abline(a = 0, b = 1, col = "red", lty = 2, lwd = 2)

rmse_m5 <- RMSE(simStrikes, test$strikes)
```

# M6: Surface Pressure

Lightning strike rate modeled from a normal distribution where mu and sigma are dependant on climate: Surface Pressure

    S ~ dnorm( mu , sigma ),
    mu <- a_mu + b_mu * SP,
    sigma <- a_sigma + b_sigma * SP,
    
```{r, echo=FALSE, results='hide', include=FALSE}

dat <- list(
  S = train$strikes,
  SP = train$sp)

m6 <- ulam(alist(
    S ~ dnorm( mu , sigma ),
    mu <- a_mu + b_mu * SP,
    sigma <- a_sigma + b_sigma * SP,
    a_mu ~ dnorm(0,10),
    b_mu ~ dnorm(0,10),
    a_sigma ~ dnorm(0,10),
    b_sigma ~ dnorm(0,10)
  ), data=dat, chains = 4, cores = 4, log_lik=TRUE)

```

```{r, echo=FALSE}
#rethinking::traceplot(m6)
summary_table <- precis(  m6 )
plot( coeftab( m6 ) , pars=c("a_mu","b_mu", "a_sigma", "b_sigma") )

#predictions 

a_mu <- summary_table["a_mu", "mean"]
b_mu <- summary_table["b_mu", "mean"]
a_sigma <- summary_table["a_sigma", "mean"]
b_sigma <- summary_table["b_sigma", "mean"]

mu <- a_mu + b_mu * test$sp
sigma <- a_sigma + b_sigma * test$sp

simStrikes <- rnorm(length(test$sp), mu, sigma)
simStrikes[simStrikes < 0] <- 0

hist(simStrikes, breaks=30)

plot(test$sp, simStrikes, main = "predictions")
plot(test$sp, test$strikes, main = "observations" )
plot(simStrikes, test$strikes)
abline(a = 0, b = 1, col = "red", lty = 2, lwd = 2)

rmse_m6 <- RMSE(simStrikes, test$strikes)
```

# M7: Precipitation

Lightning strike rate modeled from a normal distribution where mu and sigma are dependant on climate: Precipitation

    S ~ dnorm( mu , sigma ),
    mu <- a_mu + b_mu * P,
    sigma <- a_sigma + b_sigma * P,
    
```{r, echo=FALSE, results='hide', include=FALSE}

dat <- list(
  S = train$strikes,
  P = train$precip)

m7 <- ulam(alist(
    S ~ dnorm( mu , sigma ),
    mu <- a_mu + b_mu * P,
    sigma <- a_sigma + b_sigma * P,
    a_mu ~ dnorm(0,10),
    b_mu ~ dnorm(0,10),
    a_sigma ~ dnorm(0,10),
    b_sigma ~ dnorm(0,10)
  ), data=dat, chains = 4, cores = 4, log_lik=TRUE)

```

```{r, echo=FALSE}
#rethinking::traceplot(m7)
summary_table <- precis(  m7 )
plot( coeftab( m7 ) , pars=c("a_mu","b_mu", "a_sigma", "b_sigma") )

#predictions 

a_mu <- summary_table["a_mu", "mean"]
b_mu <- summary_table["b_mu", "mean"]
a_sigma <- summary_table["a_sigma", "mean"]
b_sigma <- summary_table["b_sigma", "mean"]

mu <- a_mu + b_mu * test$precip
sigma <- a_sigma + b_sigma * test$precip

simStrikes <- rnorm(length(test$precip), mu, sigma)
simStrikes[simStrikes < 0] <- 0

hist(simStrikes, breaks=30)

plot(test$precip, simStrikes, main = "predictions")
plot(test$precip, test$strikes, main = "observations" )
plot(simStrikes, test$strikes)
abline(a = 0, b = 1, col = "red", lty = 2, lwd = 2)

rmse_m7 <- RMSE(simStrikes, test$strikes)
```
# M8: Wind

Lightning strike rate modeled from a normal distribution where mu and sigma are dependant on climate: Wind

    S ~ dnorm( mu , sigma ),
    mu <- a_mu + b_mu * W,
    sigma <- a_sigma + b_sigma * W,
    
```{r, echo=FALSE, results='hide', include=FALSE}

dat <- list(
  S = train$strikes,
  W = train$wind)

m8 <- ulam(alist(
    S ~ dnorm( mu , sigma ),
    mu <- a_mu + b_mu * W,
    sigma <- a_sigma + b_sigma * W,
    a_mu ~ dnorm(0,10),
    b_mu ~ dnorm(0,10),
    a_sigma ~ dnorm(0,10),
    b_sigma ~ dnorm(0,10)
  ), data=dat, chains = 4, cores = 4, log_lik=TRUE)

```

```{r, echo=FALSE}
#rethinking::traceplot(m8)
summary_table <- precis(  m8 )
plot( coeftab( m8 ) , pars=c("a_mu","b_mu", "a_sigma", "b_sigma") )

#predictions 

a_mu <- summary_table["a_mu", "mean"]
b_mu <- summary_table["b_mu", "mean"]
a_sigma <- summary_table["a_sigma", "mean"]
b_sigma <- summary_table["b_sigma", "mean"]

mu <- a_mu + b_mu * test$wind
sigma <- a_sigma + b_sigma * test$wind

simStrikes <- rnorm(length(test$wind), mu, sigma)
simStrikes[simStrikes < 0] <- 0

hist(simStrikes, breaks=30)

plot(test$wind, simStrikes, main = "predictions")
plot(test$wind, test$strikes, main = "observations" )
plot(simStrikes, test$strikes)
abline(a = 0, b = 1, col = "red", lty = 2, lwd = 2)

rmse_m8 <- RMSE(simStrikes, test$strikes)
```


```{r, echo=FALSE}
compare(m0, m1, m2, m3, m4, m5, m6, m7, m8)
```
WAIC scores: best score is m1, CAPE. followed by shortwave radiation and temperature. (then CAPExPrecip, then relative humidity, wind, precip, surface pressure, and the null model). 
# M9: Short Wave Radiation + Temperature

Lightning strike rate modeled from a normal distribution where mu and sigma are dependant on climate: shortwave radiation and temperature, where swr and temp have addative affects. They contribute independently to the mean lightning strike rate and standard deviation. swr and temp do not intereact; ie. the effect of swr on lighting is teh same regardless of the temperature outside. 

    S ~ dnorm( mu , sigma ),
    mu <- a_mu + b_mu * SWR + c_mu * T,
    sigma <- a_sigma + b_sigma * SWR + c_sigma * T,
    
```{r, echo=FALSE, results='hide', include=FALSE}

dat <- list(
  S = train$strikes,
  SWR = train$swr,
  T = train$tair)

m9 <- ulam(alist(
    S ~ dnorm( mu , sigma ),
    mu <- a_mu + b_mu * SWR + c_mu * T,
    sigma <- a_sigma + b_sigma * SWR + c_sigma * T,
    a_mu ~ dnorm(0,10),
    b_mu ~ dnorm(0,10),
    c_mu ~ dnorm(0, 10),
    a_sigma ~ dnorm(0,10),
    b_sigma ~ dnorm(0,10),
    c_sigma ~ dnorm(0,10)
  ), data=dat, chains = 4, cores = 4, log_lik=TRUE)

```

```{r, echo=FALSE}
#rethinking::traceplot(m9)
summary_table <- precis(  m9 )
plot( coeftab( m9 ) , pars=c("a_mu", "b_mu", "c_mu", "a_sigma", "b_sigma", "c_sigma") )

#predictions 

a_mu <- summary_table["a_mu", "mean"]
b_mu <- summary_table["b_mu", "mean"]
c_mu <- summary_table["c_mu", "mean"]
a_sigma <- summary_table["a_sigma", "mean"]
b_sigma <- summary_table["b_sigma", "mean"]
c_sigma <- summary_table["c_sigma", "mean"]

mu <- a_mu + b_mu * test$swr + c_mu * test$tair
sigma <- a_sigma + b_sigma * test$swr + c_sigma * test$tair

simStrikes <- rnorm(length(test$swr), mu, sigma)
simStrikes[simStrikes < 0] <- 0

hist(simStrikes, breaks=30)

ggplot(data = data.frame(observed = test$strikes, predicted = simStrikes, T=test$tair, SWR = test$swr)) + geom_point(aes(x=SWR, y=predicted, color=T))
ggplot(data = data.frame(observed = test$strikes, predicted = simStrikes, T=test$tair, SWR = test$swr)) + geom_point(aes(x=SWR, y=observed, color=T))

plot(simStrikes, test$strikes)
abline(a = 0, b = 1, col = "red", lty = 2, lwd = 2)

rmse_m9 <- RMSE(simStrikes, test$strikes)
```


```{r, echo=FALSE}
compare(m0, m1, m2, m3, m4, m5, m6, m7, m8, m9)
```
# M10: Short Wave Radiation x Temperature

Lightning strike rate modeled from a normal distribution where mu and sigma are dependent on climate: shortwave radiation and temperature, which interact. The independent effects of short wave radiation and temperature are not accounted for in this model. 

    S ~ dnorm( mu , sigma ),
    mu <- a_mu + b_mu * SWR * T,
    sigma <- a_sigma + b_sigma * SWR * T,
    
```{r, echo=FALSE, results='hide', include=FALSE}

dat <- list(
  S = train$strikes,
  SWR = train$swr,
  T = train$tair)

m10 <- ulam(alist(
    S ~ dnorm( mu , sigma ),
    mu <- a_mu + b_mu * SWR * T,
    sigma <- a_sigma + b_sigma * SWR * T,
    a_mu ~ dnorm(0,10),
    b_mu ~ dnorm(0,10),
    a_sigma ~ dnorm(0,10),
    b_sigma ~ dnorm(0,10)
  ), data=dat, chains = 4, cores = 4, log_lik=TRUE)

```

```{r, echo=FALSE}
#rethinking::traceplot(m10)
summary_table <- precis(  m10 )
plot( coeftab( m10 ) , pars=c("a_mu", "b_mu", "a_sigma", "b_sigma") )

#predictions 

a_mu <- summary_table["a_mu", "mean"]
b_mu <- summary_table["b_mu", "mean"]
a_sigma <- summary_table["a_sigma", "mean"]
b_sigma <- summary_table["b_sigma", "mean"]

mu <- a_mu + b_mu * test$swr * test$tair
sigma <- a_sigma + b_sigma * test$swr * test$tair

simStrikes <- rnorm(length(test$swr), mu, sigma)
simStrikes[simStrikes < 0] <- 0

hist(simStrikes, breaks=30)

ggplot(data = data.frame(observed = test$strikes, predicted = simStrikes, T=test$tair, SWR = test$swr)) + geom_point(aes(x=SWR, y=predicted, color=T))
ggplot(data = data.frame(observed = test$strikes, predicted = simStrikes, T=test$tair, SWR = test$swr)) + geom_point(aes(x=SWR, y=observed, color=T))

plot(simStrikes, test$strikes)
abline(a = 0, b = 1, col = "red", lty = 2, lwd = 2)

rmse_m10 <- RMSE(simStrikes, test$strikes)
```

```{r, echo=FALSE}
compare(m0, m1, m2, m3, m4, m5, m6, m7, m8, m9, m10)
```
# M11: Short Wave Radiation + Temperature + (Short Wave Radiation x Temperature)

Lightning strike rate modeled from a normal distribution where mu and sigma are dependant on climate: shortwave radiation and temperature, where swr and temp have addative affects and an interaction term. 

    S ~ dnorm( mu , sigma ),
    mu <- a_mu + b_mu*SWR + c_mu*T + d_mu*SWR*T,
    sigma <- a_sigma + b_sigma*SWR + c_sigma*T + d_sigma*SWR*T,
    
```{r, echo=FALSE, results='hide', include=FALSE}

dat <- list(
  S = train$strikes,
  SWR = train$swr,
  T = train$tair)

m11 <- ulam(alist(
    S ~ dnorm( mu , sigma ),
    mu <- a_mu + b_mu*SWR + c_mu*T + d_mu*SWR*T,
    sigma <- a_sigma + b_sigma*SWR + c_sigma*T + d_sigma*SWR*T,
    a_mu ~ dnorm(0,10),
    b_mu ~ dnorm(0,10),
    c_mu ~ dnorm(0, 10),
    d_mu ~ dnorm(0, 10),
    a_sigma ~ dexp(1),
    b_sigma ~ dexp(1),
    c_sigma ~ dexp(1),
    d_sigma ~ dexp(1)
  ), data=dat, chains = 4, cores = 4, log_lik=TRUE)

```

```{r, echo=FALSE}
#rethinking::traceplot(m11)
summary_table <- precis(  m11 )
plot( coeftab( m11 ) , pars=c("a_mu", "b_mu", "c_mu", "d_mu", "a_sigma", "b_sigma", "c_sigma", "d_sigma") )

#predictions 

a_mu <- summary_table["a_mu", "mean"]
b_mu <- summary_table["b_mu", "mean"]
c_mu <- summary_table["c_mu", "mean"]
d_mu <- summary_table["d_mu", "mean"]
a_sigma <- summary_table["a_sigma", "mean"]
b_sigma <- summary_table["b_sigma", "mean"]
c_sigma <- summary_table["c_sigma", "mean"]
d_sigma <- summary_table["d_sigma", "mean"]

mu <- a_mu + b_mu * test$swr + c_mu * test$tair + d_mu * test$swr*test$tair
sigma <- a_sigma + b_sigma * test$swr + c_sigma * test$tair + d_sigma * test$swr*test$tair

simStrikes <- rnorm(length(test$swr), mu, sigma)
simStrikes[simStrikes < 0] <- 0

hist(simStrikes, breaks=30)

ggplot(data = data.frame(observed = test$strikes, predicted = simStrikes, T=test$tair, SWR = test$swr)) + geom_point(aes(x=SWR, y=predicted, color=T))
ggplot(data = data.frame(observed = test$strikes, predicted = simStrikes, T=test$tair, SWR = test$swr)) + geom_point(aes(x=SWR, y=observed, color=T))

plot(simStrikes, test$strikes)
abline(a = 0, b = 1, col = "red", lty = 2, lwd = 2)

rmse_m11 <- RMSE(simStrikes, test$strikes)
```

Note that d_sigma is very close to 0, which make the interaction term between swr and tair also almost 0 - so interactions between tair and swr are minimal. This means that m11 is essentially the same as m9. Add, in fact, has a similar WAIC score as m9.  

```{r, echo=FALSE}
compare(m0, m1, m2, m3, m4, m5, m6, m7, m8, m9, m10, m11)
```

# M12: Short Wave Radiation + Temperature + Relative Humidity

Lightning strike rate modeled from a normal distribution where mu and sigma are dependant on climate: shortwave radiation, temperature, and relative humidity, where each predictor variable has an additive affect, but do not interact. Relative humidity has been added since it has the next best WAIC score when comparing simpler models that predict lighting from just one climate variable. 

    S ~ dnorm( mu , sigma ),
    mu <- a_mu + b_mu*SWR + c_mu*T + d_mu*RH,
    sigma <- a_sigma + b_sigma*SWR + c_sigma*T + d_sigma*RH,
    
```{r, echo=FALSE, results='hide', include=FALSE}

dat <- list(
  S = train$strikes,
  SWR = train$swr,
  T = train$tair, 
  RH = train$rh)

m12 <- ulam(alist(
    S ~ dnorm( mu , sigma ),
    mu <- a_mu + b_mu*SWR + c_mu*T + d_mu*RH,
    sigma <- a_sigma + b_sigma*SWR + c_sigma*T + d_sigma*RH,
    a_mu ~ dnorm(0,10),
    b_mu ~ dnorm(0,10),
    c_mu ~ dnorm(0, 10),
    d_mu ~ dnorm(0, 10),
    a_sigma ~ dexp(1),
    b_sigma ~ dexp(1),
    c_sigma ~ dexp(1),
    d_sigma ~ dexp(1)
  ), data=dat, chains = 4, cores = 4, log_lik=TRUE)

```

```{r, echo=FALSE}
#rethinking::traceplot(m12)
summary_table <- precis(  m12 )
plot( coeftab( m12 ) , pars=c("a_mu", "b_mu", "c_mu", "d_mu", "a_sigma", "b_sigma", "c_sigma", "d_sigma") )

#predictions 

a_mu <- summary_table["a_mu", "mean"]
b_mu <- summary_table["b_mu", "mean"]
c_mu <- summary_table["c_mu", "mean"]
d_mu <- summary_table["d_mu", "mean"]
a_sigma <- summary_table["a_sigma", "mean"]
b_sigma <- summary_table["b_sigma", "mean"]
c_sigma <- summary_table["c_sigma", "mean"]
d_sigma <- summary_table["d_sigma", "mean"]

mu <- a_mu + b_mu * test$swr + c_mu * test$tair + d_mu * test$rh
sigma <- a_sigma + b_sigma * test$swr + c_sigma * test$tair + d_sigma * test$rh

simStrikes <- rnorm(length(test$swr), mu, sigma)
simStrikes[simStrikes < 0] <- 0

hist(simStrikes, breaks=30)

ggplot(data = data.frame(observed = test$strikes, predicted = simStrikes, T=test$tair, SWR = test$swr, RH = test$rh)) + geom_point(aes(x=SWR, y=predicted, color=RH))
ggplot(data = data.frame(observed = test$strikes, predicted = simStrikes, T=test$tair, SWR = test$swr, RH = test$rh)) + geom_point(aes(x=SWR, y=observed, color=RH))

plot(simStrikes, test$strikes)
abline(a = 0, b = 1, col = "red", lty = 2, lwd = 2)

rmse_m12 <- RMSE(simStrikes, test$strikes)
```

```{r, echo=FALSE}
compare(m0, m1, m2, m3, m4, m5, m6, m7, m8, m9, m10, m11, m12)
```
# M13: Short Wave Radiation + Temperature + Relative Humidity + Wind

Lightning strike rate modeled from a normal distribution where mu and sigma are dependent on climate: shortwave radiation, temperature, relative humidity, and wind, where each predictor variable has an additive affect, but do not interact. The wind model has the next best WAIC score.  

    S ~ dnorm( mu , sigma ),
    mu <- a_mu + b_mu*SWR + c_mu*T + d_mu*RH + e_mu*W,
    sigma <- a_sigma + b_sigma*SWR + c_sigma*T + d_sigma*RH +e_sigma*W,
    
```{r, echo=FALSE, results='hide', include=FALSE}

dat <- list(
  S = train$strikes,
  SWR = train$swr,
  T = train$tair, 
  RH = train$rh,
  W = train$wind)

m13 <- ulam(alist(
    S ~ dnorm( mu , sigma ),
    mu <- a_mu + b_mu*SWR + c_mu*T + d_mu*RH + e_mu*W,
    sigma <- a_sigma + b_sigma*SWR + c_sigma*T + d_sigma*RH +e_sigma*W,
    a_mu ~ dnorm(0,10),
    b_mu ~ dnorm(0,10),
    c_mu ~ dnorm(0, 10),
    d_mu ~ dnorm(0, 10),
    e_mu ~ dnorm(0, 10),
    a_sigma ~ dexp(1),
    b_sigma ~ dexp(1),
    c_sigma ~ dexp(1),
    d_sigma ~ dexp(1),
    e_sigma ~ dexp(1)
  ), data=dat, chains = 4, cores = 4, log_lik=TRUE)

```

```{r, echo=FALSE}
#rethinking::traceplot(m13)
summary_table <- precis(  m13 )
plot( coeftab( m13 ) , pars=c("a_mu", "b_mu", "c_mu", "d_mu", "e_mu", "a_sigma", "b_sigma", "c_sigma", "d_sigma", "e_sigma") )

#predictions 

a_mu <- summary_table["a_mu", "mean"]
b_mu <- summary_table["b_mu", "mean"]
c_mu <- summary_table["c_mu", "mean"]
d_mu <- summary_table["d_mu", "mean"]
e_mu <- summary_table["e_mu", "mean"]
a_sigma <- summary_table["a_sigma", "mean"]
b_sigma <- summary_table["b_sigma", "mean"]
c_sigma <- summary_table["c_sigma", "mean"]
d_sigma <- summary_table["d_sigma", "mean"]
e_sigma <- summary_table["e_sigma", "mean"]

mu <- a_mu + b_mu * test$swr + c_mu * test$tair + d_mu * test$rh + e_mu * test$wind
sigma <- a_sigma + b_sigma * test$swr + c_sigma * test$tair + d_sigma * test$rh + e_sigma * test$wind

simStrikes <- rnorm(length(test$swr), mu, sigma)
simStrikes[simStrikes < 0] <- 0

hist(simStrikes, breaks=30)

ggplot(data = data.frame(observed = test$strikes, predicted = simStrikes, T=test$tair, SWR = test$swr, RH = test$rh, W = test$wind)) + geom_point(aes(x=SWR, y=predicted, color=W))
ggplot(data = data.frame(observed = test$strikes, predicted = simStrikes, T=test$tair, SWR = test$swr, RH = test$rh, W = test$wind)) + geom_point(aes(x=SWR, y=observed, color=W))

plot(simStrikes, test$strikes)
abline(a = 0, b = 1, col = "red", lty = 2, lwd = 2)

rmse_m13 <- RMSE(simStrikes, test$strikes)
```

```{r, echo=FALSE}
compare(m0, m1, m2, m3, m4, m5, m6, m7, m8, m9, m10, m11, m12, m13)
```
# M14: Short Wave Radiation + Temperature + Relative Humidity + Wind + Precipitation 

Lightning strike rate modeled from a normal distribution where mu and sigma are dependent on climate: shortwave radiation, temperature, relative humidity, wind, and precipitation, where each predictor variable has an additive affect, but do not interact. The precipitation model has the next best WAIC score.  

    S ~ dnorm( mu , sigma ),
    mu <- a_mu + b_mu*SWR + c_mu*T + d_mu*RH + e_mu*W + f_mu*P,
    sigma <- a_sigma + b_sigma*SWR + c_sigma*T + d_sigma*RH + e_sigma*W + f_sigma*P,

```{r, echo=FALSE, results='hide', include=FALSE}

dat <- list(
  S = train$strikes,
  SWR = train$swr,
  T = train$tair, 
  RH = train$rh,
  W = train$wind,
  P= train$precip)

m14 <- ulam(alist(
    S ~ dnorm( mu , sigma ),
    mu <- a_mu + b_mu*SWR + c_mu*T + d_mu*RH + e_mu*W + f_mu*P,
    sigma <- a_sigma + b_sigma*SWR + c_sigma*T + d_sigma*RH + e_sigma*W + f_sigma*P,
    a_mu ~ dnorm(0,10),
    b_mu ~ dnorm(0,10),
    c_mu ~ dnorm(0, 10),
    d_mu ~ dnorm(0, 10),
    e_mu ~ dnorm(0, 10),
    f_mu ~ dnorm(0, 10),
    a_sigma ~ dexp(1),
    b_sigma ~ dexp(1),
    c_sigma ~ dexp(1),
    d_sigma ~ dexp(1),
    e_sigma ~ dexp(1),
    f_sigma ~ dexp(1)
  ), data=dat, chains = 4, cores = 4, log_lik=TRUE)

```

```{r, echo=FALSE}
#rethinking::traceplot(m14)
summary_table <- precis(  m14 )
plot( coeftab( m14 ) , pars=c("a_mu", "b_mu", "c_mu", "d_mu", "e_mu", "f_mu", "a_sigma", "b_sigma", "c_sigma", "d_sigma", "e_sigma", "f_sigma") )

#predictions 

a_mu <- summary_table["a_mu", "mean"]
b_mu <- summary_table["b_mu", "mean"]
c_mu <- summary_table["c_mu", "mean"]
d_mu <- summary_table["d_mu", "mean"]
e_mu <- summary_table["e_mu", "mean"]
f_mu <- summary_table["f_mu", "mean"]
a_sigma <- summary_table["a_sigma", "mean"]
b_sigma <- summary_table["b_sigma", "mean"]
c_sigma <- summary_table["c_sigma", "mean"]
d_sigma <- summary_table["d_sigma", "mean"]
e_sigma <- summary_table["e_sigma", "mean"]
f_sigma <- summary_table["f_sigma", "mean"]

mu <- a_mu + b_mu * test$swr + c_mu * test$tair + d_mu * test$rh + e_mu * test$wind + f_mu * test$precip
sigma <- a_sigma + b_sigma * test$swr + c_sigma * test$tair + d_sigma * test$rh + e_sigma * test$wind + f_sigma * test$precip

simStrikes <- rnorm(length(test$swr), mu, sigma)
simStrikes[simStrikes < 0] <- 0

hist(simStrikes, breaks=30)

ggplot(data = data.frame(observed = test$strikes, predicted = simStrikes, T=test$tair, SWR = test$swr, RH = test$rh, W = test$wind, P = test$precip)) + geom_point(aes(x=SWR, y=predicted, color=P))
ggplot(data = data.frame(observed = test$strikes, predicted = simStrikes, T=test$tair, SWR = test$swr, RH = test$rh, W = test$wind, P = test$precip)) + geom_point(aes(x=SWR, y=observed, color=P))

plot(simStrikes, test$strikes)
abline(a = 0, b = 1, col = "red", lty = 2, lwd = 2)

rmse_m14 <- RMSE(simStrikes, test$strikes)
```

```{r, echo=FALSE}
compare(m0, m1, m2, m3, m4, m5, m6, m7, m8, m9, m10, m11, m12, m13, m14)
```
# M15: Short Wave Radiation + Temperature + Relative Humidity + Wind + Precipitation + Surface Pressure

Lightning strike rate modeled from a normal distribution where mu and sigma are dependent on climate: shortwave radiation, temperature, relative humidity, wind, precipitation, and surface pressure, where each predictor variable has an additive affect, but do not interact. The surface pressure model has the next best WAIC score, but also the lowest after the null model, so it may not improve predictions by much. 

    S ~ dnorm( mu , sigma ),
    mu <- a_mu + b_mu*SWR + c_mu*T + d_mu*RH + e_mu*W + f_mu*P + g_mu*SP,
    sigma <- a_sigma + b_sigma*SWR + c_sigma*T + d_sigma*RH + e_sigma*W + f_sigma*P + g_sigma*SP,

```{r, echo=FALSE, results='hide', include=FALSE}

dat <- list(
  S = train$strikes,
  SWR = train$swr,
  T = train$tair, 
  RH = train$rh,
  W = train$wind,
  P= train$precip, 
  SP = train$sp)

m15 <- ulam(alist(
    S ~ dnorm( mu , sigma ),
    mu <- a_mu + b_mu*SWR + c_mu*T + d_mu*RH + e_mu*W + f_mu*P + g_mu*SP,
    sigma <- a_sigma + b_sigma*SWR + c_sigma*T + d_sigma*RH + e_sigma*W + f_sigma*P + g_sigma*SP,
    a_mu ~ dnorm(0,10),
    b_mu ~ dnorm(0,10),
    c_mu ~ dnorm(0, 10),
    d_mu ~ dnorm(0, 10),
    e_mu ~ dnorm(0, 10),
    f_mu ~ dnorm(0, 10),
    g_mu ~ dnorm(0, 10),
    a_sigma ~ dexp(1),
    b_sigma ~ dexp(1),
    c_sigma ~ dexp(1),
    d_sigma ~ dexp(1),
    e_sigma ~ dexp(1),
    f_sigma ~ dexp(1),
    g_sigma ~ dexp(1)
  ), data=dat, chains = 4, cores = 4, log_lik=TRUE)

```

```{r, echo=FALSE}
#rethinking::traceplot(m15)
summary_table <- precis(  m15 )
plot( coeftab( m15 ) , pars=c("a_mu", "b_mu", "c_mu", "d_mu", "e_mu", "f_mu", "g_mu", "a_sigma", "b_sigma", "c_sigma", "d_sigma", "e_sigma", "f_sigma", "g_sigma") )

#predictions 

a_mu <- summary_table["a_mu", "mean"]
b_mu <- summary_table["b_mu", "mean"]
c_mu <- summary_table["c_mu", "mean"]
d_mu <- summary_table["d_mu", "mean"]
e_mu <- summary_table["e_mu", "mean"]
f_mu <- summary_table["f_mu", "mean"]
g_mu <- summary_table["g_mu", "mean"]
a_sigma <- summary_table["a_sigma", "mean"]
b_sigma <- summary_table["b_sigma", "mean"]
c_sigma <- summary_table["c_sigma", "mean"]
d_sigma <- summary_table["d_sigma", "mean"]
e_sigma <- summary_table["e_sigma", "mean"]
f_sigma <- summary_table["f_sigma", "mean"]
g_sigma <- summary_table["g_sigma", "mean"]

mu <- a_mu + b_mu * test$swr + c_mu * test$tair + d_mu * test$rh + e_mu * test$wind + f_mu * test$precip + g_mu * test$sp
sigma <- a_sigma + b_sigma * test$swr + c_sigma * test$tair + d_sigma * test$rh + e_sigma * test$wind + f_sigma * test$precip + g_sigma * test$sp

simStrikes <- rnorm(length(test$swr), mu, sigma)
simStrikes[simStrikes < 0] <- 0

hist(simStrikes, breaks=30)

ggplot(data = data.frame(observed = test$strikes, predicted = simStrikes, T=test$tair, SWR = test$swr, RH = test$rh, W = test$wind, P = test$precip, SP = test$sp)) + geom_point(aes(x=SWR, y=predicted, color=SP))
ggplot(data = data.frame(observed = test$strikes, predicted = simStrikes, T=test$tair, SWR = test$swr, RH = test$rh, W = test$wind, P = test$precip, SP = test$sp)) + geom_point(aes(x=SWR, y=observed, color=SP))

plot(simStrikes, test$strikes)
abline(a = 0, b = 1, col = "red", lty = 2, lwd = 2)

rmse_m15 <- RMSE(simStrikes, test$strikes)
```

```{r, echo=FALSE}
compare(m0, m1, m2, m3, m4, m5, m6, m7, m8, m9, m10, m11, m12, m13, m14, m15)
```
# M16: Short Wave Radiation + Temperature + Relative Humidity + Precipitation + Surface Pressure

Adding surface pressure to the model produces a lower WAIC score, and wind also reduces the WAIC score. The next two models remove wind, then surface pressure and wind. 

    S ~ dnorm( mu , sigma ),
    mu <- a_mu + b_mu*SWR + c_mu*T + d_mu*RH + f_mu*P + g_mu*SP,
    sigma <- a_sigma + b_sigma*SWR + c_sigma*T + d_sigma*RH + f_sigma*P + g_sigma*SP,
    
```{r, echo=FALSE, results='hide', include=FALSE}

dat <- list(
  S = train$strikes,
  SWR = train$swr,
  T = train$tair, 
  RH = train$rh,
  P= train$precip, 
  SP = train$sp)

m16 <- ulam(alist(
    S ~ dnorm( mu , sigma ),
    mu <- a_mu + b_mu*SWR + c_mu*T + d_mu*RH + f_mu*P + g_mu*SP,
    sigma <- a_sigma + b_sigma*SWR + c_sigma*T + d_sigma*RH + f_sigma*P + g_sigma*SP,
    a_mu ~ dnorm(0,10),
    b_mu ~ dnorm(0,10),
    c_mu ~ dnorm(0, 10),
    d_mu ~ dnorm(0, 10),
    f_mu ~ dnorm(0, 10),
    g_mu ~ dnorm(0, 10),
    a_sigma ~ dexp(1),
    b_sigma ~ dexp(1),
    c_sigma ~ dexp(1),
    d_sigma ~ dexp(1),
    f_sigma ~ dexp(1),
    g_sigma ~ dexp(1)
  ), data=dat, chains = 4, cores = 4, log_lik=TRUE)

```

```{r, echo=FALSE}
#rethinking::traceplot(m16)
summary_table <- precis(  m16 )
plot( coeftab( m16 ) , pars=c("a_mu", "b_mu", "c_mu", "d_mu", "f_mu", "g_mu", "a_sigma", "b_sigma", "c_sigma", "d_sigma", "f_sigma", "g_sigma") )

#predictions 

a_mu <- summary_table["a_mu", "mean"]
b_mu <- summary_table["b_mu", "mean"]
c_mu <- summary_table["c_mu", "mean"]
d_mu <- summary_table["d_mu", "mean"]
f_mu <- summary_table["f_mu", "mean"]
g_mu <- summary_table["g_mu", "mean"]
a_sigma <- summary_table["a_sigma", "mean"]
b_sigma <- summary_table["b_sigma", "mean"]
c_sigma <- summary_table["c_sigma", "mean"]
d_sigma <- summary_table["d_sigma", "mean"]
f_sigma <- summary_table["f_sigma", "mean"]
g_sigma <- summary_table["g_sigma", "mean"]

mu <- a_mu + b_mu * test$swr + c_mu * test$tair + d_mu * test$rh + f_mu * test$precip + g_mu * test$sp
sigma <- a_sigma + b_sigma * test$swr + c_sigma * test$tair + d_sigma * test$rh + f_sigma * test$precip + g_sigma * test$sp

simStrikes <- rnorm(length(test$swr), mu, sigma)
simStrikes[simStrikes < 0] <- 0

hist(simStrikes, breaks=30)

ggplot(data = data.frame(observed = test$strikes, predicted = simStrikes, T=test$tair, SWR = test$swr, RH = test$rh, P = test$precip, SP = test$sp)) + geom_point(aes(x=SWR, y=predicted, color=SP))
ggplot(data = data.frame(observed = test$strikes, predicted = simStrikes, T=test$tair, SWR = test$swr, RH = test$rh, P = test$precip, SP = test$sp)) + geom_point(aes(x=SWR, y=observed, color=SP))

plot(simStrikes, test$strikes)
abline(a = 0, b = 1, col = "red", lty = 2, lwd = 2)

rmse_m16 <- RMSE(simStrikes, test$strikes)
```
# M17: Short Wave Radiation + Temperature + Relative Humidity + Precipitation 

    S ~ dnorm( mu , sigma ),
    mu <- a_mu + b_mu*SWR + c_mu*T + d_mu*RH + f_mu*P,
    sigma <- a_sigma + b_sigma*SWR + c_sigma*T + d_sigma*RH + f_sigma*P,

```{r, echo=FALSE, results='hide', include=FALSE}

dat <- list(
  S = train$strikes,
  SWR = train$swr,
  T = train$tair, 
  RH = train$rh,
  P= train$precip)

m17 <- ulam(alist(
    S ~ dnorm( mu , sigma ),
    mu <- a_mu + b_mu*SWR + c_mu*T + d_mu*RH + f_mu*P,
    sigma <- a_sigma + b_sigma*SWR + c_sigma*T + d_sigma*RH + f_sigma*P,
    a_mu ~ dnorm(0,10),
    b_mu ~ dnorm(0,10),
    c_mu ~ dnorm(0, 10),
    d_mu ~ dnorm(0, 10),
    f_mu ~ dnorm(0, 10),
    a_sigma ~ dexp(1),
    b_sigma ~ dexp(1),
    c_sigma ~ dexp(1),
    d_sigma ~ dexp(1),
    f_sigma ~ dexp(1)
  ), data=dat, chains = 4, cores = 4, log_lik=TRUE)

```

```{r, echo=FALSE}
#rethinking::traceplot(m17)
summary_table <- precis(  m17 )
plot( coeftab( m17 ) , pars=c("a_mu", "b_mu", "c_mu", "d_mu", "f_mu", "a_sigma", "b_sigma", "c_sigma", "d_sigma", "f_sigma") )

#predictions 

a_mu <- summary_table["a_mu", "mean"]
b_mu <- summary_table["b_mu", "mean"]
c_mu <- summary_table["c_mu", "mean"]
d_mu <- summary_table["d_mu", "mean"]
f_mu <- summary_table["f_mu", "mean"]
a_sigma <- summary_table["a_sigma", "mean"]
b_sigma <- summary_table["b_sigma", "mean"]
c_sigma <- summary_table["c_sigma", "mean"]
d_sigma <- summary_table["d_sigma", "mean"]
f_sigma <- summary_table["f_sigma", "mean"]

mu <- a_mu + b_mu * test$swr + c_mu * test$tair + d_mu * test$rh + f_mu * test$precip 
sigma <- a_sigma + b_sigma * test$swr + c_sigma * test$tair + d_sigma * test$rh + f_sigma * test$precip

simStrikes <- rnorm(length(test$swr), mu, sigma)
simStrikes[simStrikes < 0] <- 0

hist(simStrikes, breaks=30)

ggplot(data = data.frame(observed = test$strikes, predicted = simStrikes, T=test$tair, SWR = test$swr, RH = test$rh, P = test$precip)) + geom_point(aes(x=SWR, y=predicted, color=P))
ggplot(data = data.frame(observed = test$strikes, predicted = simStrikes, T=test$tair, SWR = test$swr, RH = test$rh, P = test$precip)) + geom_point(aes(x=SWR, y=observed, color=P))

plot(simStrikes, test$strikes)
abline(a = 0, b = 1, col = "red", lty = 2, lwd = 2)

rmse_m17 <- RMSE(simStrikes, test$strikes)
```



```{r}
compare(m0, m1, m2, m3, m4, m5, m6, m7, m8, m9, m10, m11, m12, m13, m14, m15, m16, m17)
```
M16 has the lowest WAIC score! This model excludes wind: 

    S ~ dnorm( mu , sigma )
    mu = a_mu + b_mu*SWR + c_mu*T + d_mu*RH + f_mu*P + g_mu*SP
    sigma = a_sigma + b_sigma*SWR + c_sigma*T + d_sigma*RH + f_sigma*P + g_sigma*SP

    
Route Mean Square Error:
```{r}
rmse_table <- data.frame(cbind(
  model = c("m0", "m1", "m2", "m3", "m4", "m5", "m6", "m7", "m8", "m9", "m10", "m11", "m12", "m13", "m14", "m15", "m16", "m17"),
  RMSE = c(rmse_m0, rmse_m1, rmse_m2, rmse_m3, rmse_m4, rmse_m5, rmse_m6, rmse_m7, rmse_m8, rmse_m9, rmse_m10, rmse_m11, rmse_m12, rmse_m13, rmse_m14, rmse_m15, rmse_m16, rmse_m17)))

rmse_table <- rmse_table %>%
  arrange(RMSE)

print(rmse_table)

#write.csv(rmse_table, "/raid/cuden/LightningModels/Uden/rmse_rethinking_normal_models_negativesToZeros.csv")
#write.csv(rmse_table, "/raid/cuden/LightningModels/Uden/rmse_rethinking_normal_models.csv")
```
