---
title: "Rstan Models"
author: "Charlotte Uden"
date: "2024-11-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, include=FALSE}
library(rstan)
library(loo)
library(dplyr)
library(stats)
library(caret)
```
1. Read in the data


```{r, echo=FALSE}
data <- read.csv("/raid/cuden/data/era5_vaisalaLightning_monthlySummerSummaries_2005-2010_NEclip.csv")[,2:16]
colnames(data) <- c("lat", "lon", "year", "strikes", "cape", "precip", "cxp", "d2m", "tair", "wind", "swr", "sp", "rh", "tmin", "tmax")
hist(data$strikes, breaks = 30)
```

2. Split into training and testing data
```{r, echo=FALSE}
## 80% of the sample size
sample_size <- floor(0.8 * nrow(data))

## set the seed to make partition reproducible
set.seed(123)
train_index <- sample(seq_len(nrow(data)), size = sample_size)

train <- data[train_index, ]
test <- data[-train_index, ]

```


3. Standardize predictor variables 
```{r, echo=FALSE}
#save min and max (normalization parameters?):
#-----
mu_cxp <- mean(train$cxp)
sigma_cxp <- sd(train$cxp)
mu_cape <- mean(train$cape)
sigma_cape <- sd(train$cape)
mu_precip <- mean(train$precip)
sigma_precip <- sd(train$precip)
mu_tair <- mean(train$tair)
sigma_tair <- sd(train$tair)
mu_wind <- mean(train$wind)
sigma_wind <- sd(train$wind)
mu_swr <- mean(train$swr)
sigma_swr <- sd(train$swr)
mu_sp <- mean(train$sp)
sigma_sp <- sd(train$sp)
mu_rh <- mean(train$rh)
sigma_rh <- sd(train$rh)
#-----

train <- train %>%
  mutate(cape = (cape-mu_cape)/sigma_cape) %>%
  mutate(precip = (precip-mu_precip)/sigma_precip) %>%
  mutate(cxp = (cxp-mu_cxp)/sigma_cxp) %>%
  mutate(tair = (tair-mu_tair)/sigma_tair) %>%
  mutate(wind = (wind-mu_wind)/sigma_wind) %>%
  mutate(swr = (swr-mu_swr)/sigma_swr) %>%
  mutate(sp = (sp-mu_sp)/sigma_sp) %>%
  mutate(rh = (rh-mu_rh)/sigma_rh) 

train <- train[,c("strikes", "cape", "precip", "cxp", "tair", "wind", "swr", "sp", "rh")]
head(train)

plot(train)

#use the min and max values from the train set to normalize the test set 
test <- test %>%
  mutate(cape = (cape-mu_cape)/sigma_cape) %>%
  mutate(precip = (precip-mu_precip)/sigma_precip) %>%
  mutate(cxp = (cxp-mu_cxp)/sigma_cxp) %>%
  mutate(tair = (tair-mu_tair)/sigma_tair) %>%
  mutate(wind = (wind-mu_wind)/sigma_wind) %>%
  mutate(swr = (swr-mu_swr)/sigma_swr) %>%
  mutate(sp = (sp-mu_sp)/sigma_sp) %>%
  mutate(rh = (rh-mu_rh)/sigma_rh) 

test <- test[,c("strikes", "cape", "precip", "cxp", "tair", "wind", "swr", "sp", "rh")]

```

```{r, echo=FALSE, results='hide', include=FALSE}

#library(rstan)

dat <- list(
  N = nrow(train),
  S = train$strikes
)


# Compile and fit the Stan model
m0.1 <- stan(file = "normal.stan", data = dat, 
           iter = 2000, 
           chains = 4, 
           cores = min(parallel::detectCores(), 4))

```

```{r}
print(m0.1, pars = c("mu", "sigma"))

traceplot(m0.1)

parameter_stats <- summary(m0.1)$summary[, c("mean", "sd")][-3,]
print(parameter_stats[-3,])

summary_table <- data.frame(
  parameter = rownames(parameter_stats),
  mean = parameter_stats[, "mean"],
  sd = parameter_stats[, "sd"]
)

ggplot(summary_table, aes(x = parameter, y = mean)) +
  geom_point(color = "blue", size = 3) +  # Point for the mean
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), width = 0.2) +  # Error bars for SD
  labs(y = "Parameter Estimate", x = "Parameter") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Parameter Means and SDs from Stan Model")

#predictions 
mean_mu <- summary_table["mu", "mean"]
mean_sigma <- summary_table["sigma", "mean"]

simStrikes <- rnorm(length(test$strikes), mean_mu, mean_sigma)

hist(simStrikes, breaks=30, main = "predictions")

hist(test$strikes, breaks=30, main = "observations")

```
Model Comparison in Rstan:
Loo: Leave one out cross validation
WAIC: widely applicable information criterion

```{r}
# Extract the log-likelihood
log_lik <- extract_log_lik(m0.1, parameter_name = "log_lik")

# Calculate WAIC
waic_result_m0.1 <- waic(log_lik)
# Calculate LOO
loo_result_m0.1 <- loo(log_lik)

# Print WAIC and Loo result
print(waic_result_m0.1)
print(loo_result_m0.1)

```



```{r, echo=FALSE, results='hide', include=FALSE}
# 
# #library(rstan)
# 
# dat <- list(
#   N = nrow(train),
#   S = train$strikes
# )
# 
# 
# # Compile and fit the Stan model
# m0.2 <- stan(file = "normal_truncated.stan", data = dat, 
#            iter = 2000, 
#            chains = 4, 
#            cores = min(parallel::detectCores(), 4))
# 
# print(m2, pars = c("mu", "sigma"))
# 
# traceplot(m2)
# 
# parameter_stats <- summary(m2)$summary[, c("mean", "sd")][-3,]
# print(parameter_stats[-3,])
# 
# summary_table <- data.frame(
#   parameter = rownames(parameter_stats),
#   mean = parameter_stats[, "mean"],
#   sd = parameter_stats[, "sd"]
# )
# 
# ggplot(summary_table, aes(x = parameter, y = mean)) +
#   geom_point(color = "blue", size = 3) +  # Point for the mean
#   geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), width = 0.2) +  # Error bars for SD
#   labs(y = "Parameter Estimate", x = "Parameter") +
#   theme_minimal() +
#   theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
#   ggtitle("Parameter Means and SDs from Stan Model")
# 
# #predictions 
# mean_mu <- summary_table["mu", "mean"]
# mean_sigma <- summary_table["sigma", "mean"]
# 
# simStrikes <- rnorm(length(test$strikes), mean_mu, mean_sigma)
# 
# simStrikes <- pmax(0, simStrikes)
# 
# hist(simStrikes, breaks=30, main = "predictions")
# 
# hist(test$strikes, breaks=30, main = "observations")

```

```{r, echo=FALSE, results='hide', include=FALSE}

#library(rstan)

dat <- list(
  N = nrow(train),
  S = train$strikes
)


# Compile and fit the Stan model
m0.3 <- stan(file = "gamma.stan", data = dat, 
           iter = 2000, 
           chains = 4, 
           cores = min(parallel::detectCores(), 4))

```


```{r}

print(m0.3, pars = c("alpha", "beta"))

traceplot(m0.3)

parameter_stats <- summary(m0.3)$summary[, c("mean", "sd")][1:2, 1:2]
print(parameter_stats)

summary_table <- data.frame(
  parameter = rownames(parameter_stats),
  mean = parameter_stats[, "mean"],
  sd = parameter_stats[, "sd"]
)

ggplot(summary_table, aes(x = parameter, y = mean)) +
  geom_point(color = "blue", size = 3) +  # Point for the mean
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), width = 0.2) +  # Error bars for SD
  labs(y = "Parameter Estimate", x = "Parameter") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Parameter Means and SDs from Stan Model")

#predictions 
mean_alpha <- summary_table["alpha", "mean"]
mean_beta <- summary_table["beta", "mean"]

simStrikes <- rgamma(length(test$strikes), mean_alpha, mean_beta)

hist(simStrikes, breaks=30, main = "predictions")

hist(test$strikes, breaks=30, main = "observations")

```


```{r}
# Extract the log-likelihood
log_lik <- extract_log_lik(m0.3, parameter_name = "log_lik")

# Calculate WAIC
waic_result_m0.3 <- waic(log_lik)
# Calculate LOO
loo_result_m0.3 <- loo(log_lik)

# Print WAIC and Loo result
print(waic_result_m0.3)
print(loo_result_m0.3)

```


# M1: CAPE

S ~ gamma(alpha, beta);
alpha = a_alpha + b_alpha * CAPE;
beta = a_beta + b_beta * CAPE;

```{r, echo=FALSE, results='hide', include=FALSE}

#library(rstan)

dat <- list(
  N = nrow(train),
  S = train$strikes,
  CAPE = train$cape
)


# Compile and fit the Stan model
m1 <- stan(file = "gamma_cape.stan", data = dat, 
           iter = 2000, 
           chains = 4, 
           cores = min(parallel::detectCores(), 4))
```


```{r}

print(m1, pars = c("a_alpha", "b_alpha", "a_beta", "b_beta"))

traceplot(m1)

parameter_stats <- summary(m1)$summary[, c("mean", "sd")][1:4, 1:2]
print(parameter_stats)

summary_table <- data.frame(
  parameter = rownames(parameter_stats),
  mean = parameter_stats[, "mean"],
  sd = parameter_stats[, "sd"]
)

ggplot(summary_table, aes(x = parameter, y = mean)) +
  geom_point(color = "blue", size = 3) +  # Point for the mean
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), width = 0.2) +  # Error bars for SD
  labs(y = "Parameter Estimate", x = "Parameter") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Parameter Means and SDs from Stan Model")

#predictions 
a_alpha <- summary_table["a_alpha", "mean"]
b_alpha <- summary_table["b_alpha", "mean"]
a_beta <- summary_table["a_beta", "mean"]
b_beta  <- summary_table["b_beta ", "mean"]
a_beta <- summary_table[3,2]
b_beta  <- summary_table[4,2]

alpha <-  a_alpha + b_alpha * test$cape
beta <-  a_beta + b_beta * test$cape

simStrikes <- rgamma(length(test$strikes), alpha, beta)

hist(simStrikes, breaks=30, main = "predictions")

hist(test$strikes, breaks=30, main = "observations")

plot(test$cape, simStrikes)
plot(test$cape, test$strikes)
plot(simStrikes, test$strikes)
abline(a = 0, b = 1, col = "red", lty = 2, lwd = 2)

```


```{r}
# Extract the log-likelihood
log_lik <- extract_log_lik(m1, parameter_name = "log_lik")

# Calculate WAIC
waic_result_m1 <- waic(log_lik)
# Calculate LOO
loo_result_m1 <- loo(log_lik)

# Print WAIC and Loo result
print(waic_result_m1)
print(loo_result_m1)

```

# M2: CAPE x Precipitation

S ~ gamma(alpha, beta);
alpha = a_alpha + b_alpha * CXP;
beta = a_beta + b_beta * CXP;

```{r, echo=FALSE, results='hide', include=FALSE}

#library(rstan)

dat <- list(
  N = nrow(train),
  S = train$strikes,
  CXP = train$cxp
)


# Compile and fit the Stan model
m2 <- stan(file = "gamma_cxp.stan", data = dat, 
           iter = 2000, 
           chains = 4, 
           cores = min(parallel::detectCores(), 4))
```


```{r}

print(m2, pars = c("a_alpha", "b_alpha", "a_beta", "b_beta"))

traceplot(m2)

parameter_stats <- summary(m2)$summary[, c("mean", "sd")][1:4, 1:2]
print(parameter_stats)

summary_table <- data.frame(
  parameter = rownames(parameter_stats),
  mean = parameter_stats[, "mean"],
  sd = parameter_stats[, "sd"]
)

ggplot(summary_table, aes(x = parameter, y = mean)) +
  geom_point(color = "blue", size = 3) +  # Point for the mean
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), width = 0.2) +  # Error bars for SD
  labs(y = "Parameter Estimate", x = "Parameter") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Parameter Means and SDs from Stan Model")

#predictions 
a_alpha <- summary_table["a_alpha", "mean"]
b_alpha <- summary_table["b_alpha", "mean"]
a_beta <- summary_table["a_beta", "mean"]
b_beta  <- summary_table["b_beta ", "mean"]
a_beta <- summary_table[3,2]
b_beta  <- summary_table[4,2]

alpha <-  a_alpha + b_alpha * test$cxp
beta <-  a_beta + b_beta * test$cxp

simStrikes <- rgamma(length(test$strikes), alpha, beta)

hist(simStrikes, breaks=30, main = "predictions")

hist(test$strikes, breaks=30, main = "observations")

plot(test$cxp, simStrikes)
plot(test$cxp, test$strikes)
plot(simStrikes, test$strikes)
abline(a = 0, b = 1, col = "red", lty = 2, lwd = 2)

```

```{r}
# Extract the log-likelihood
log_lik <- extract_log_lik(m2, parameter_name = "log_lik")

# Calculate WAIC
waic_result_m2 <- waic(log_lik)
# Calculate LOO
loo_result_m2 <- loo(log_lik)

# Print WAIC and Loo result
print(waic_result_m2)
print(loo_result_m2)

```

# M3: Relative Humidity

S ~ gamma(alpha, beta);
alpha = a_alpha + b_alpha * RH;
beta = a_beta + b_beta * RH;

```{r, echo=FALSE, results='hide', include=FALSE}

#library(rstan)

dat <- list(
  N = nrow(train),
  S = train$strikes,
  RH = train$rh
)


# Compile and fit the Stan model
m3 <- stan(file = "gamma_rh.stan", data = dat, 
           iter = 2000, 
           chains = 4, 
           cores = min(parallel::detectCores(), 4))
```


```{r}

print(m3, pars = c("a_alpha", "b_alpha", "a_beta", "b_beta"))

traceplot(m3)

parameter_stats <- summary(m3)$summary[, c("mean", "sd")][1:4, 1:2]
print(parameter_stats)

summary_table <- data.frame(
  parameter = rownames(parameter_stats),
  mean = parameter_stats[, "mean"],
  sd = parameter_stats[, "sd"]
)

ggplot(summary_table, aes(x = parameter, y = mean)) +
  geom_point(color = "blue", size = 3) +  # Point for the mean
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), width = 0.2) +  # Error bars for SD
  labs(y = "Parameter Estimate", x = "Parameter") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Parameter Means and SDs from Stan Model")

#predictions 
a_alpha <- summary_table["a_alpha", "mean"]
b_alpha <- summary_table["b_alpha", "mean"]
a_beta <- summary_table["a_beta", "mean"]
b_beta  <- summary_table["b_beta ", "mean"]
a_beta <- summary_table[3,2]
b_beta  <- summary_table[4,2]

alpha <-  a_alpha + b_alpha * test$rh
beta <-  a_beta + b_beta * test$rh

simStrikes <- rgamma(length(test$strikes), alpha, beta)

hist(simStrikes, breaks=30, main = "predictions")

hist(test$strikes, breaks=30, main = "observations")

plot(test$rh, simStrikes)
plot(test$rh, test$strikes)
plot(simStrikes, test$strikes)
abline(a = 0, b = 1, col = "red", lty = 2, lwd = 2)

```

```{r}
# Extract the log-likelihood
log_lik <- extract_log_lik(m3, parameter_name = "log_lik")

# Calculate WAIC
waic_result_m3 <- waic(log_lik)
# Calculate LOO
loo_result_m3 <- loo(log_lik)

# Print WAIC and Loo result
print(waic_result_m3)
print(loo_result_m3)

```

# M4: Short Wave Radiation

S ~ gamma(alpha, beta);
alpha = a_alpha + b_alpha * SWR;
beta = a_beta + b_beta * SWR;

```{r, echo=FALSE, results='hide', include=FALSE}

#library(rstan)

dat <- list(
  N = nrow(train),
  S = train$strikes,
  SWR = train$swr
)


# Compile and fit the Stan model
m4 <- stan(file = "gamma_swr.stan", data = dat, 
           iter = 2000, 
           chains = 4, 
           cores = min(parallel::detectCores(), 4))
```


```{r}

print(m4, pars = c("a_alpha", "b_alpha", "a_beta", "b_beta"))

traceplot(m4)

parameter_stats <- summary(m4)$summary[, c("mean", "sd")][1:4, 1:2]
print(parameter_stats)

summary_table <- data.frame(
  parameter = rownames(parameter_stats),
  mean = parameter_stats[, "mean"],
  sd = parameter_stats[, "sd"]
)

ggplot(summary_table, aes(x = parameter, y = mean)) +
  geom_point(color = "blue", size = 3) +  # Point for the mean
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), width = 0.2) +  # Error bars for SD
  labs(y = "Parameter Estimate", x = "Parameter") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Parameter Means and SDs from Stan Model")

#predictions 
a_alpha <- summary_table["a_alpha", "mean"]
b_alpha <- summary_table["b_alpha", "mean"]
a_beta <- summary_table["a_beta", "mean"]
b_beta  <- summary_table["b_beta ", "mean"]
a_beta <- summary_table[3,2]
b_beta  <- summary_table[4,2]

alpha <-  a_alpha + b_alpha * test$swr
beta <-  a_beta + b_beta * test$swr

simStrikes <- rgamma(length(test$strikes), alpha, beta)

hist(simStrikes, breaks=30, main = "predictions")

hist(test$strikes, breaks=30, main = "observations")

plot(test$swr, simStrikes)
plot(test$swr, test$strikes)
plot(simStrikes, test$strikes)
abline(a = 0, b = 1, col = "red", lty = 2, lwd = 2)

```

```{r}
# Extract the log-likelihood
log_lik <- extract_log_lik(m4, parameter_name = "log_lik")

# Calculate WAIC
waic_result_m4 <- waic(log_lik)
# Calculate LOO
loo_result_m4 <- loo(log_lik)

# Print WAIC and Loo result
print(waic_result_m4)
print(loo_result_m4)

```

# M5: Temperature

S ~ gamma(alpha, beta);
alpha = a_alpha + b_alpha * T;
beta = a_beta + b_beta * T;

```{r, echo=FALSE, results='hide', include=FALSE}

#library(rstan)

dat <- list(
  N = nrow(train),
  S = train$strikes,
  T = train$tair
)


# Compile and fit the Stan model
m5 <- stan(file = "gamma_tair.stan", data = dat, 
           iter = 2000, 
           chains = 4, 
           cores = min(parallel::detectCores(), 4))
```


```{r}

print(m5, pars = c("a_alpha", "b_alpha", "a_beta", "b_beta"))

traceplot(m5)

parameter_stats <- summary(m5)$summary[, c("mean", "sd")][1:4, 1:2]
print(parameter_stats)

summary_table <- data.frame(
  parameter = rownames(parameter_stats),
  mean = parameter_stats[, "mean"],
  sd = parameter_stats[, "sd"]
)

ggplot(summary_table, aes(x = parameter, y = mean)) +
  geom_point(color = "blue", size = 3) +  # Point for the mean
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), width = 0.2) +  # Error bars for SD
  labs(y = "Parameter Estimate", x = "Parameter") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Parameter Means and SDs from Stan Model")

#predictions 
a_alpha <- summary_table["a_alpha", "mean"]
b_alpha <- summary_table["b_alpha", "mean"]
a_beta <- summary_table["a_beta", "mean"]
b_beta  <- summary_table["b_beta ", "mean"]
a_beta <- summary_table[3,2]
b_beta  <- summary_table[4,2]

alpha <-  a_alpha + b_alpha * test$tair
beta <-  a_beta + b_beta * test$tair

simStrikes <- rgamma(length(test$strikes), alpha, beta)

hist(simStrikes, breaks=30, main = "predictions")

hist(test$strikes, breaks=30, main = "observations")

plot(test$tair, simStrikes)
plot(test$tair, test$strikes)
plot(simStrikes, test$strikes)
abline(a = 0, b = 1, col = "red", lty = 2, lwd = 2)

```


```{r}
# Extract the log-likelihood
log_lik <- extract_log_lik(m5, parameter_name = "log_lik")

# Calculate WAIC
waic_result_m5 <- waic(log_lik)
# Calculate LOO
loo_result_m5 <- loo(log_lik)

# Print WAIC and Loo result
print(waic_result_m5)
print(loo_result_m5)

```

# M6: Surface Pressure

S ~ gamma(alpha, beta);
alpha = a_alpha + b_alpha * SP;
beta = a_beta + b_beta * SP;

```{r, echo=FALSE, results='hide', include=FALSE}

#library(rstan)

dat <- list(
  N = nrow(train),
  S = train$strikes,
  SP = train$sp
)


# Compile and fit the Stan model
m6 <- stan(file = "gamma_sp.stan", data = dat, 
           iter = 2000, 
           chains = 4, 
           cores = min(parallel::detectCores(), 4))
```


```{r}

print(m6, pars = c("a_alpha", "b_alpha", "a_beta", "b_beta"))

traceplot(m6)

parameter_stats <- summary(m6)$summary[, c("mean", "sd")][1:4, 1:2]
print(parameter_stats)

summary_table <- data.frame(
  parameter = rownames(parameter_stats),
  mean = parameter_stats[, "mean"],
  sd = parameter_stats[, "sd"]
)

ggplot(summary_table, aes(x = parameter, y = mean)) +
  geom_point(color = "blue", size = 3) +  # Point for the mean
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), width = 0.2) +  # Error bars for SD
  labs(y = "Parameter Estimate", x = "Parameter") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Parameter Means and SDs from Stan Model")

#predictions 
a_alpha <- summary_table["a_alpha", "mean"]
b_alpha <- summary_table["b_alpha", "mean"]
a_beta <- summary_table["a_beta", "mean"]
b_beta  <- summary_table["b_beta ", "mean"]
a_beta <- summary_table[3,2]
b_beta  <- summary_table[4,2]

alpha <-  a_alpha + b_alpha * test$sp
beta <-  a_beta + b_beta * test$sp

simStrikes <- rgamma(length(test$strikes), alpha, beta)

hist(simStrikes, breaks=30, main = "predictions")

hist(test$strikes, breaks=30, main = "observations")

plot(test$sp, simStrikes)
plot(test$sp, test$strikes)
plot(simStrikes, test$strikes)
abline(a = 0, b = 1, col = "red", lty = 2, lwd = 2)

```

```{r}
# Extract the log-likelihood
log_lik <- extract_log_lik(m6, parameter_name = "log_lik")

# Calculate WAIC
waic_result_m6 <- waic(log_lik)
# Calculate LOO
loo_result_m6 <- loo(log_lik)

# Print WAIC and Loo result
print(waic_result_m6)
print(loo_result_m6)

```


# M7: Precipitation

S ~ gamma(alpha, beta);
alpha = a_alpha + b_alpha * P;
beta = a_beta + b_beta * P;

```{r, echo=FALSE, results='hide', include=FALSE}

#library(rstan)

dat <- list(
  N = nrow(train),
  S = train$strikes,
  P = train$precip
)


# Compile and fit the Stan model
m7 <- stan(file = "gamma_precip.stan", data = dat, 
           iter = 2000, 
           chains = 4, 
           cores = min(parallel::detectCores(), 4))
```


```{r}

print(m7, pars = c("a_alpha", "b_alpha", "a_beta", "b_beta"))

traceplot(m7)

parameter_stats <- summary(m7)$summary[, c("mean", "sd")][1:4, 1:2]
print(parameter_stats)

summary_table <- data.frame(
  parameter = rownames(parameter_stats),
  mean = parameter_stats[, "mean"],
  sd = parameter_stats[, "sd"]
)

ggplot(summary_table, aes(x = parameter, y = mean)) +
  geom_point(color = "blue", size = 3) +  # Point for the mean
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), width = 0.2) +  # Error bars for SD
  labs(y = "Parameter Estimate", x = "Parameter") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Parameter Means and SDs from Stan Model")

#predictions 
a_alpha <- summary_table["a_alpha", "mean"]
b_alpha <- summary_table["b_alpha", "mean"]
a_beta <- summary_table["a_beta", "mean"]
b_beta  <- summary_table["b_beta ", "mean"]
a_beta <- summary_table[3,2]
b_beta  <- summary_table[4,2]

alpha <-  a_alpha + b_alpha * test$precip
beta <-  a_beta + b_beta * test$precip

simStrikes <- rgamma(length(test$strikes), alpha, beta)

hist(simStrikes, breaks=30, main = "predictions")

hist(test$strikes, breaks=30, main = "observations")

plot(test$precip, simStrikes)
plot(test$precip, test$strikes)
plot(simStrikes, test$strikes)
abline(a = 0, b = 1, col = "red", lty = 2, lwd = 2)

```

```{r}
# Extract the log-likelihood
log_lik <- extract_log_lik(m7, parameter_name = "log_lik")

# Calculate WAIC
waic_result_m7 <- waic(log_lik)
# Calculate LOO
loo_result_m7 <- loo(log_lik)

# Print WAIC and Loo result
print(waic_result_m7)
print(loo_result_m7)

```

# M8: Wind

S ~ gamma(alpha, beta);
alpha = a_alpha + b_alpha * W;
beta = a_beta + b_beta * W;

```{r, echo=FALSE, results='hide', include=FALSE}

#library(rstan)

dat <- list(
  N = nrow(train),
  S = train$strikes,
  W = train$wind
)


# Compile and fit the Stan model
m8 <- stan(file = "gamma_wind.stan", data = dat, 
           iter = 2000, 
           chains = 4, 
           cores = min(parallel::detectCores(), 4))
```


```{r}

print(m8, pars = c("a_alpha", "b_alpha", "a_beta", "b_beta"))

traceplot(m8)

parameter_stats <- summary(m8)$summary[, c("mean", "sd")][1:4, 1:2]
print(parameter_stats)

summary_table <- data.frame(
  parameter = rownames(parameter_stats),
  mean = parameter_stats[, "mean"],
  sd = parameter_stats[, "sd"]
)

ggplot(summary_table, aes(x = parameter, y = mean)) +
  geom_point(color = "blue", size = 3) +  # Point for the mean
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), width = 0.2) +  # Error bars for SD
  labs(y = "Parameter Estimate", x = "Parameter") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Parameter Means and SDs from Stan Model")

#predictions 
a_alpha <- summary_table["a_alpha", "mean"]
b_alpha <- summary_table["b_alpha", "mean"]
a_beta <- summary_table["a_beta", "mean"]
b_beta  <- summary_table["b_beta ", "mean"]
a_beta <- summary_table[3,2]
b_beta  <- summary_table[4,2]

alpha <-  a_alpha + b_alpha * test$wind
beta <-  a_beta + b_beta * test$wind

simStrikes <- rgamma(length(test$strikes), alpha, beta)

hist(simStrikes, breaks=30, main = "predictions")

hist(test$strikes, breaks=30, main = "observations")

plot(test$wind, simStrikes)
plot(test$wind, test$strikes)
plot(simStrikes, test$strikes)
abline(a = 0, b = 1, col = "red", lty = 2, lwd = 2)

```

```{r}
# Extract the log-likelihood
log_lik <- extract_log_lik(m8, parameter_name = "log_lik")

# Calculate WAIC
waic_result_m8 <- waic(log_lik)
# Calculate LOO
loo_result_m8 <- loo(log_lik)

# Print WAIC and Loo result
print(waic_result_m8)
print(loo_result_m8)

```

```{r}
colnames <- c("model", "climate_var", "elpd_loo", "p_loo", "looic", "elpd_waic", "p_waic", "waic")
model_comparison <- data.frame(rbind(c("m0.1", "normal", loo_result_m0.1$estimates[1,1], loo_result_m0.1$estimates[2,1], loo_result_m0.1$estimate[3,1], waic_result_m0.1$estimates[1,1], waic_result_m0.1$estimates[2,1], waic_result_m0.1$estimates[3,1]),
      c("m0.3", "gamma", loo_result_m0.3$estimates[1,1], loo_result_m0.3$estimates[2,1], loo_result_m0.3$estimate[3,1], waic_result_m0.3$estimates[1,1], waic_result_m2$estimates[2,1], waic_result_m2$estimates[3,1]),
  c("m1", "CAPE", loo_result_m1$estimates[1,1], loo_result_m1$estimates[2,1], loo_result_m1$estimate[3,1], waic_result_m1$estimates[1,1], waic_result_m1$estimates[2,1], waic_result_m1$estimates[3,1]),
  c("m2", "CXP", loo_result_m2$estimates[1,1], loo_result_m2$estimates[2,1], loo_result_m2$estimate[3,1], waic_result_m2$estimates[1,1], waic_result_m2$estimates[2,1], waic_result_m2$estimates[3,1]),
  c("m3", "rh", loo_result_m3$estimates[1,1], loo_result_m3$estimates[2,1], loo_result_m3$estimate[3,1], waic_result_m3$estimates[1,1], waic_result_m3$estimates[2,1], waic_result_m3$estimates[3,1]),
  c("m4", "swr", loo_result_m4$estimates[1,1], loo_result_m4$estimates[2,1], loo_result_m4$estimate[3,1], waic_result_m4$estimates[1,1], waic_result_m4$estimates[2,1], waic_result_m4$estimates[3,1]),
  c("m5", "tair", loo_result_m5$estimates[1,1], loo_result_m5$estimates[2,1], loo_result_m5$estimate[3,1], waic_result_m5$estimates[1,1], waic_result_m5$estimates[2,1], waic_result_m5$estimates[3,1]),
  c("m6", "sp", loo_result_m6$estimates[1,1], loo_result_m6$estimates[2,1], loo_result_m6$estimate[3,1], waic_result_m6$estimates[1,1], waic_result_m6$estimates[2,1], waic_result_m6$estimates[3,1]),
  c("m7", "precip", loo_result_m7$estimates[1,1], loo_result_m7$estimates[2,1], loo_result_m7$estimate[3,1], waic_result_m7$estimates[1,1], waic_result_m7$estimates[2,1], waic_result_m7$estimates[3,1]),
  c("m8", "wind", loo_result_m8$estimates[1,1], loo_result_m8$estimates[2,1], loo_result_m8$estimates[3,1], waic_result_m8$estimates[1,1], waic_result_m8$estimates[2,1], waic_result_m8$estimates[3,1])))

#model_comparison[,3:8] <- as.numeric(model_comparison[,3:8])

colnames(model_comparison) <- colnames

#print(arrange(model_comparison, waic))

model_comparison
```
The highest elpd_loo scores are cape and cape x precip. of the surface level climate variables, short wave radiation and temperature perform best, followed by relative humidity, wind, precipitation, and surface pressure. WAIC scores also indicate that short wave radiation and temperature are the best surface level climate predictors of lighting. Interestingly, the simple gamma model (m0.3) performs even better than models that include climate. Perhaps this is because WAIC takes number of parameters into cnsiteration. 

# M9: Short Wave Radiation + Temperature

Lightning strike rate modeled from a normal distribution where mu and sigma are dependent on climate: shortwave radiation and temperature, where SWR and temp have additive affects. They contribute independently to the mean lightning strike rate and standard deviation. swr and temp do not interact; ie. the effect of SWR on lighting is the same regardless of the temperature outside. 

    S ~ dnorm( mu , sigma )
    mu <- a_alpha + b_alpha * SWR + c_alpha * T
    sigma <- a_beta + b_beta * SWR + c_beta * T


```{r, echo=FALSE, results='hide', include=FALSE}

#library(rstan)

dat <- list(
  N = nrow(train),
  S = train$strikes,
  SWR = train$swr,
  T = train$tair
)


# Compile and fit the Stan model
m9 <- stan(file = "gamma_swr_tair.stan", data = dat, 
           iter = 2000, 
           chains = 4, 
           cores = min(parallel::detectCores(), 4))
```


```{r}

print(m9, pars = c("a_alpha", "b_alpha", "c_alpha", "a_beta", "b_beta", "c_beta"))

traceplot(m9)

parameter_stats <- summary(m9)$summary[, c("mean", "sd")][1:6, 1:2]
print(parameter_stats)

summary_table <- data.frame(
  parameter = rownames(parameter_stats),
  mean = parameter_stats[, "mean"],
  sd = parameter_stats[, "sd"]
)

ggplot(summary_table, aes(x = parameter, y = mean)) +
  geom_point(color = "blue", size = 3) +  # Point for the mean
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), width = 0.2) +  # Error bars for SD
  labs(y = "Parameter Estimate", x = "Parameter") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Parameter Means and SDs from Stan Model")

#predictions 
a_alpha <- summary_table["a_alpha", "mean"]
b_alpha <- summary_table["b_alpha", "mean"]
c_alpha <- summary_table["c_alpha", "mean"]
a_beta  <- summary_table["a_beta", "mean"]
b_beta <- summary_table["b_beta", "mean"]
c_beta  <- summary_table["c_beta", "mean"]

alpha <-  a_alpha + b_alpha * test$swr + c_alpha * test$tair
beta <-  a_beta + b_beta * test$swr + c_beta * test$tair

simStrikes <- rgamma(length(test$strikes), alpha, beta)

hist(simStrikes, breaks=30, main = "predictions")

hist(test$strikes, breaks=30, main = "observations")

ggplot(data = data.frame(observed = test$strikes, predicted = simStrikes, T=test$tair, SWR = test$swr, RH = test$rh, P = test$precip)) + geom_point(aes(x=SWR, y=predicted, color=T))
ggplot(data = data.frame(observed = test$strikes, predicted = simStrikes, T=test$tair, SWR = test$swr, RH = test$rh, P = test$precip)) + geom_point(aes(x=SWR, y=observed, color=T))

plot(simStrikes, test$strikes)
abline(a = 0, b = 1, col = "red", lty = 2, lwd = 2)

```

```{r}
# Extract the log-likelihood
log_lik <- extract_log_lik(m9, parameter_name = "log_lik")

# Calculate WAIC
waic_result_m9 <- waic(log_lik)
# Calculate LOO
loo_result_m9 <- loo(log_lik)

# Print WAIC and Loo result
print(waic_result_m9)
print(loo_result_m9)

```
This model performs a little bit better than the simple short wave radiation model (m4). 

# M12: Short Wave Radiation + Temperature + Precipitation

Lightning strike rate modeled from a normal distribution where mu and sigma are dependent on climate: shortwave radiation and temperature, where SWR and temp have additive affects. They contribute independently to the mean lightning strike rate and standard deviation. swr and temp do not interact; ie. the effect of SWR on lighting is the same regardless of the temperature outside. 

    S ~ dnorm( mu , sigma ),
    mu <- a_alpha + b_alpha * SWR + c_alpha * T + d_alpha * RH,
    sigma <- a_beta + b_beta * SWR + c_beta * T+ d_beta * RH,
    
```{r, echo=FALSE, results='hide', include=FALSE}

#library(rstan)

dat <- list(
  N = nrow(train),
  S = train$strikes,
  SWR = train$swr,
  T = train$tair,
  RH = train$rh
)


# Compile and fit the Stan model
m12 <- stan(file = "gamma_swr_tair_rh.stan", data = dat, 
           iter = 2000, 
           chains = 4, 
           cores = min(parallel::detectCores(), 4))
```


```{r}

print(m12, pars = c("a_alpha", "b_alpha", "c_alpha", "d_alpha", "a_beta", "b_beta", "c_beta", "d_beta"))

traceplot(m12)

parameter_stats <- summary(m12)$summary[, c("mean", "sd")][1:8, 1:2]
print(parameter_stats)

summary_table <- data.frame(
  parameter = rownames(parameter_stats),
  mean = parameter_stats[, "mean"],
  sd = parameter_stats[, "sd"]
)

ggplot(summary_table, aes(x = parameter, y = mean)) +
  geom_point(color = "blue", size = 3) +  # Point for the mean
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), width = 0.2) +  # Error bars for SD
  labs(y = "Parameter Estimate", x = "Parameter") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Parameter Means and SDs from Stan Model")

#predictions 
a_alpha <- summary_table["a_alpha", "mean"]
b_alpha <- summary_table["b_alpha", "mean"]
c_alpha <- summary_table["c_alpha", "mean"]
d_alpha <- summary_table["d_alpha", "mean"]
a_beta  <- summary_table["a_beta", "mean"]
b_beta <- summary_table["b_beta", "mean"]
c_beta  <- summary_table["c_beta", "mean"]
d_beta  <- summary_table["d_beta", "mean"]

alpha <-  a_alpha + b_alpha * test$swr + c_alpha * test$tair + d_alpha * test$rh
beta <-  a_beta + b_beta * test$swr + c_beta * test$tair + d_beta * test$rh

simStrikes <- rgamma(length(test$strikes), alpha, beta)

hist(simStrikes, breaks=30, main = "predictions")

hist(test$strikes, breaks=30, main = "observations")

ggplot(data = data.frame(observed = test$strikes, predicted = simStrikes, T=test$tair, SWR = test$swr, RH = test$rh, P = test$precip)) + geom_point(aes(x=SWR, y=predicted, color=RH))
ggplot(data = data.frame(observed = test$strikes, predicted = simStrikes, T=test$tair, SWR = test$swr, RH = test$rh, P = test$precip)) + geom_point(aes(x=SWR, y=observed, color=RH))

plot(simStrikes, test$strikes)
abline(a = 0, b = 1, col = "red", lty = 2, lwd = 2)

```


```{r}
# Extract the log-likelihood
log_lik <- extract_log_lik(m12, parameter_name = "log_lik")

# Calculate WAIC
waic_result_m12 <- waic(log_lik)
# Calculate LOO
loo_result_m12 <- loo(log_lik)

# Print WAIC and Loo result
print(waic_result_m12)
print(loo_result_m12)

```


# M13: Short Wave Radiation + Temperature + Precipitation + Wind

Lightning strike rate modeled from a normal distribution where mu and sigma are dependent on climate: shortwave radiation,  temperature, relative humidity, and wind. 
    S ~ dnorm( mu , sigma ),
    mu <- a_alpha + b_alpha * SWR + c_alpha * T + d_alpha * RH + e_alpha * W,
    sigma <- a_beta + b_beta * SWR + c_beta * T + d_beta * RH + e_beta * W,,
    
```{r, echo=FALSE, results='hide', include=FALSE}

#library(rstan)

dat <- list(
  N = nrow(train),
  S = train$strikes,
  SWR = train$swr,
  T = train$tair,
  RH = train$rh,
  W = train$wind
)


# Compile and fit the Stan model
m13 <- stan(file = "gamma_swr_tair_rh_wind.stan", data = dat, 
           iter = 2000, 
           chains = 4, 
           cores = min(parallel::detectCores(), 4))
```


```{r}

print(m13, pars = c("a_alpha", "b_alpha", "c_alpha", "d_alpha", "e_alpha", "a_beta", "b_beta", "c_beta", "d_beta", "e_beta"))

traceplot(m13)

parameter_stats <- summary(m13)$summary[, c("mean", "sd")][1:10, 1:2]
print(parameter_stats)

summary_table <- data.frame(
  parameter = rownames(parameter_stats),
  mean = parameter_stats[, "mean"],
  sd = parameter_stats[, "sd"]
)

ggplot(summary_table, aes(x = parameter, y = mean)) +
  geom_point(color = "blue", size = 3) +  # Point for the mean
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), width = 0.2) +  # Error bars for SD
  labs(y = "Parameter Estimate", x = "Parameter") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Parameter Means and SDs from Stan Model")

#predictions 
a_alpha <- summary_table["a_alpha", "mean"]
b_alpha <- summary_table["b_alpha", "mean"]
c_alpha <- summary_table["c_alpha", "mean"]
d_alpha <- summary_table["d_alpha", "mean"]
e_alpha <- summary_table["e_alpha", "mean"]
a_beta <- summary_table["a_beta", "mean"]
b_beta <- summary_table["b_beta", "mean"]
c_beta <- summary_table["c_beta", "mean"]
d_beta <- summary_table["d_beta", "mean"]
e_beta <- summary_table["e_beta", "mean"]

alpha <-  a_alpha + b_alpha * test$swr + c_alpha * test$tair + d_alpha * test$rh + e_alpha * test$wind
beta <-  a_beta + b_beta * test$swr + c_beta * test$tair + d_beta * test$rh + e_beta * test$wind

simStrikes <- rgamma(length(test$strikes), alpha, beta)

hist(simStrikes, breaks=30, main = "predictions")

hist(test$strikes, breaks=30, main = "observations")

ggplot(data = data.frame(observed = test$strikes, predicted = simStrikes, T=test$tair, SWR = test$swr, RH = test$rh, P = test$precip, W = test$wind)) + geom_point(aes(x=SWR, y=predicted, color=W))
ggplot(data = data.frame(observed = test$strikes, predicted = simStrikes, T=test$tair, SWR = test$swr, RH = test$rh, P = test$precip, W = test$wind)) + geom_point(aes(x=SWR, y=observed, color=W))

plot(simStrikes, test$strikes)
abline(a = 0, b = 1, col = "red", lty = 2, lwd = 2)

```


```{r}
# Extract the log-likelihood
log_lik <- extract_log_lik(m13, parameter_name = "log_lik")

# Calculate WAIC
waic_result_m13 <- waic(log_lik)
# Calculate LOO
loo_result_m13 <- loo(log_lik)

# Print WAIC and Loo result
print(waic_result_m13)
print(loo_result_m13)

```

# M14: Short Wave Radiation + Temperature + Precipitation + Wind + Precipitation

Lightning strike rate modeled from a normal distribution where mu and sigma are dependent on climate: shortwave radiation,  temperature, relative humidity, wind, and precipitation. 
    S ~ dnorm( mu , sigma ),
    mu <- a_alpha + b_alpha * SWR + c_alpha * T + d_alpha * RH + e_alpha * W + f_alpha * P,
    sigma <- a_beta + b_beta * SWR + c_beta * T + d_beta * RH + e_beta * W + f_beta * P,
    
```{r, echo=FALSE, results='hide', include=FALSE}

#library(rstan)

dat <- list(
  N = nrow(train),
  S = train$strikes,
  SWR = train$swr,
  T = train$tair,
  RH = train$rh,
  W = train$wind,
  P = train$precip
)


# Compile and fit the Stan model
init_list <- function() {
  list(a_alpha = 0.1, b_alpha = 0.1, c_alpha = 0.1, d_alpha = 0.1, e_alpha = 0.1, f_alpha = 0.1,
       a_beta = 0.1, b_beta = 0.1, c_beta = 0.1, d_beta = 0.1, e_beta = 0.1, f_beta = 0.1)
}

m14 <- stan(file = "gamma_swr_tair_rh_wind_precip.stan", 
            data = dat,
            iter = 2000, 
            chains = 4, 
            init = init_list,
            cores = min(parallel::detectCores(), 4))

```


```{r}

print(m14, pars = c("a_alpha", "b_alpha", "c_alpha", "d_alpha", "e_alpha", "f_alpha", "a_beta", "b_beta", "c_beta", "d_beta", "e_beta", "f_beta"))

traceplot(m14)

parameter_stats <- summary(m14)$summary[, c("mean", "sd")][1:12, 1:2]
print(parameter_stats)

summary_table <- data.frame(
  parameter = rownames(parameter_stats),
  mean = parameter_stats[, "mean"],
  sd = parameter_stats[, "sd"]
)

ggplot(summary_table, aes(x = parameter, y = mean)) +
  geom_point(color = "blue", size = 3) +  # Point for the mean
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), width = 0.2) +  # Error bars for SD
  labs(y = "Parameter Estimate", x = "Parameter") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Parameter Means and SDs from Stan Model")

#predictions 
a_alpha <- summary_table["a_alpha", "mean"]
b_alpha <- summary_table["b_alpha", "mean"]
c_alpha <- summary_table["c_alpha", "mean"]
d_alpha <- summary_table["d_alpha", "mean"]
e_alpha <- summary_table["e_alpha", "mean"]
f_alpha <- summary_table["f_alpha", "mean"]
a_beta <- summary_table["a_beta", "mean"]
b_beta <- summary_table["b_beta", "mean"]
c_beta <- summary_table["c_beta", "mean"]
d_beta <- summary_table["d_beta", "mean"]
e_beta <- summary_table["e_beta", "mean"]
f_beta <- summary_table["f_beta", "mean"]

alpha <-  exp(a_alpha + b_alpha * test$swr + c_alpha * test$tair + d_alpha * test$rh + e_alpha * test$wind + f_alpha * test$precip)
beta <-  exp(a_beta + b_beta * test$swr + c_beta * test$tair + d_beta * test$rh + e_beta * test$wind + f_beta * test$precip)

simStrikes <- rgamma(length(test$strikes), alpha, beta)

hist(simStrikes, breaks=30, main = "predictions")

hist(test$strikes, breaks=30, main = "observations")

ggplot(data = data.frame(observed = test$strikes, predicted = simStrikes, T=test$tair, SWR = test$swr, RH = test$rh, P = test$precip, W = test$wind)) + geom_point(aes(x=SWR, y=predicted, color=P))
ggplot(data = data.frame(observed = test$strikes, predicted = simStrikes, T=test$tair, SWR = test$swr, RH = test$rh, P = test$precip, W = test$wind)) + geom_point(aes(x=SWR, y=observed, color=P))

plot(simStrikes, test$strikes)
abline(a = 0, b = 1, col = "red", lty = 2, lwd = 2)

```


```{r}
# Extract the log-likelihood
log_lik <- extract_log_lik(m14, parameter_name = "log_lik")

# Calculate WAIC
waic_result_m14 <- waic(log_lik)
# Calculate LOO
loo_result_m14 <- loo(log_lik)

# Print WAIC and Loo result
print(waic_result_m14)
print(loo_result_m14)

```

# M15: Short Wave Radiation + Temperature + Precipitation + Wind + Precipitation + Surface Pressure

Lightning strike rate modeled from a normal distribution where mu and sigma are dependent on climate: shortwave radiation,  temperature, relative humidity, wind, precipitation, and surface pressure
. 
    S ~ dnorm( mu , sigma ),
    mu <- a_alpha + b_alpha * SWR + c_alpha * T + d_alpha * RH + e_alpha * W + f_alpha * P + g_alpha * SP,
    sigma <- a_beta + b_beta * SWR + c_beta * T + d_beta * RH + e_beta * W + f_beta * P + g_beta * SP,
    
```{r, echo=FALSE, results='hide', include=FALSE}

#library(rstan)

dat <- list(
  N = nrow(train),
  S = train$strikes,
  SWR = train$swr,
  T = train$tair,
  RH = train$rh,
  W = train$wind,
  P = train$precip,
  SP = train$sp
)


# Compile and fit the Stan model
init_list <- function() {
  list(a_alpha = 0.1, b_alpha = 0.1, c_alpha = 0.1, d_alpha = 0.1, e_alpha = 0.1, f_alpha = 0.1, g_alpha = 0.1,
       a_beta = 0.1, b_beta = 0.1, c_beta = 0.1, d_beta = 0.1, e_beta = 0.1, f_beta = 0.1, g_beta = 0.1)
}

m15 <- stan(file = "gamma_swr_tair_rh_wind_precip_sp.stan", 
            data = dat,
            iter = 2000, 
            chains = 4, 
            init = init_list,
            cores = min(parallel::detectCores(), 4))

```


```{r}

print(m15, pars = c("a_alpha", "b_alpha", "c_alpha", "d_alpha", "e_alpha", "f_alpha", "g_alpha", "a_beta", "b_beta", "c_beta", "d_beta", "e_beta", "f_beta", "g_beta"))

traceplot(m15)

parameter_stats <- summary(m15)$summary[, c("mean", "sd")][1:14, 1:2]
print(parameter_stats)

summary_table <- data.frame(
  parameter = rownames(parameter_stats),
  mean = parameter_stats[, "mean"],
  sd = parameter_stats[, "sd"]
)

ggplot(summary_table, aes(x = parameter, y = mean)) +
  geom_point(color = "blue", size = 3) +  # Point for the mean
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), width = 0.2) +  # Error bars for SD
  labs(y = "Parameter Estimate", x = "Parameter") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Parameter Means and SDs from Stan Model")

#predictions 
a_alpha <- summary_table["a_alpha", "mean"]
b_alpha <- summary_table["b_alpha", "mean"]
c_alpha <- summary_table["c_alpha", "mean"]
d_alpha <- summary_table["d_alpha", "mean"]
e_alpha <- summary_table["e_alpha", "mean"]
f_alpha <- summary_table["f_alpha", "mean"]
g_alpha <- summary_table["g_alpha", "mean"]
a_beta <- summary_table["a_beta", "mean"]
b_beta <- summary_table["b_beta", "mean"]
c_beta <- summary_table["c_beta", "mean"]
d_beta <- summary_table["d_beta", "mean"]
e_beta <- summary_table["e_beta", "mean"]
f_beta <- summary_table["f_beta", "mean"]
g_beta <- summary_table["g_beta", "mean"]

alpha <-  exp(a_alpha + b_alpha * test$swr + c_alpha * test$tair + d_alpha * test$rh + e_alpha * test$wind + f_alpha * test$precip + g_alpha * test$sp)
beta <-  exp(a_beta + b_beta * test$swr + c_beta * test$tair + d_beta * test$rh + e_beta * test$wind + f_beta * test$precip + g_beta * test$sp)

simStrikes <- rgamma(length(test$strikes), alpha, beta)

hist(simStrikes, breaks=30, main = "predictions")

hist(test$strikes, breaks=30, main = "observations")

ggplot(data = data.frame(observed = test$strikes, predicted = simStrikes, T=test$tair, SWR = test$swr, RH = test$rh, P = test$precip, W = test$wind, SP = test$sp)) + geom_point(aes(x=SWR, y=predicted, color=SP))
ggplot(data = data.frame(observed = test$strikes, predicted = simStrikes, T=test$tair, SWR = test$swr, RH = test$rh, P = test$precip, W = test$wind, SP = test$sp)) + geom_point(aes(x=SWR, y=observed, color=SP))

plot(simStrikes, test$strikes)
abline(a = 0, b = 1, col = "red", lty = 2, lwd = 2)

```

```{r}
# Extract the log-likelihood
log_lik <- extract_log_lik(m15, parameter_name = "log_lik")

# Calculate WAIC
waic_result_m15 <- waic(log_lik)
# Calculate LOO
loo_result_m15 <- loo(log_lik)

# Print WAIC and Loo result
print(waic_result_m15)
print(loo_result_m15)

```

```{r}
colnames <- c("model", "climate_var", "elpd_loo", "p_loo", "looic", "elpd_waic", "p_waic", "waic")
model_comparison <- data.frame(rbind(
  c("m0.1", "normal", loo_result_m0.1$estimates[1,1], loo_result_m0.1$estimates[2,1], loo_result_m0.1$estimate[3,1], waic_result_m0.1$estimates[1,1], waic_result_m0.1$estimates[2,1], waic_result_m0.1$estimates[3,1]),
  c("m0.3", "gamma", loo_result_m0.3$estimates[1,1], loo_result_m0.3$estimates[2,1], loo_result_m0.3$estimate[3,1], waic_result_m0.3$estimates[1,1], waic_result_m2$estimates[2,1], waic_result_m2$estimates[3,1]),
  c("m1", "CAPE", loo_result_m1$estimates[1,1], loo_result_m1$estimates[2,1], loo_result_m1$estimate[3,1], waic_result_m1$estimates[1,1], waic_result_m1$estimates[2,1], waic_result_m1$estimates[3,1]),
  c("m2", "CXP", loo_result_m2$estimates[1,1], loo_result_m2$estimates[2,1], loo_result_m2$estimate[3,1], waic_result_m2$estimates[1,1], waic_result_m2$estimates[2,1], waic_result_m2$estimates[3,1]),
  c("m3", "rh", loo_result_m3$estimates[1,1], loo_result_m3$estimates[2,1], loo_result_m3$estimate[3,1], waic_result_m3$estimates[1,1], waic_result_m3$estimates[2,1], waic_result_m3$estimates[3,1]),
  c("m4", "swr", loo_result_m4$estimates[1,1], loo_result_m4$estimates[2,1], loo_result_m4$estimate[3,1], waic_result_m4$estimates[1,1], waic_result_m4$estimates[2,1], waic_result_m4$estimates[3,1]),
  c("m5", "tair", loo_result_m5$estimates[1,1], loo_result_m5$estimates[2,1], loo_result_m5$estimate[3,1], waic_result_m5$estimates[1,1], waic_result_m5$estimates[2,1], waic_result_m5$estimates[3,1]),
  c("m6", "sp", loo_result_m6$estimates[1,1], loo_result_m6$estimates[2,1], loo_result_m6$estimate[3,1], waic_result_m6$estimates[1,1], waic_result_m6$estimates[2,1], waic_result_m6$estimates[3,1]),
  c("m7", "precip", loo_result_m7$estimates[1,1], loo_result_m7$estimates[2,1], loo_result_m7$estimate[3,1], waic_result_m7$estimates[1,1], waic_result_m7$estimates[2,1], waic_result_m7$estimates[3,1]),
  c("m8", "wind", loo_result_m8$estimates[1,1], loo_result_m8$estimates[2,1], loo_result_m8$estimates[3,1], waic_result_m8$estimates[1,1], waic_result_m8$estimates[2,1], waic_result_m8$estimates[3,1]),
  c("m9", "SWR+T", loo_result_m9$estimates[1,1], loo_result_m9$estimates[2,1], loo_result_m9$estimates[3,1], waic_result_m9$estimates[1,1], waic_result_m9$estimates[2,1], waic_result_m9$estimates[3,1]),
  c("m12", "SWR+T+RH", loo_result_m12$estimates[1,1], loo_result_m12$estimates[2,1], loo_result_m12$estimates[3,1], waic_result_m12$estimates[1,1], waic_result_m12$estimates[2,1], waic_result_m12$estimates[3,1]),
  c("m13", "SWR+T+RH+W", loo_result_m13$estimates[1,1], loo_result_m13$estimates[2,1], loo_result_m13$estimates[3,1], waic_result_m13$estimates[1,1], waic_result_m13$estimates[2,1], waic_result_m13$estimates[3,1]),
  c("m14", "SWR+T+RH+W+P", loo_result_m14$estimates[1,1], loo_result_m14$estimates[2,1], loo_result_m14$estimates[3,1], waic_result_m14$estimates[1,1], waic_result_m14$estimates[2,1], waic_result_m14$estimates[3,1]),
  c("m15", "SWR+T+RH+W+P+SP", loo_result_m15$estimates[1,1], loo_result_m15$estimates[2,1], loo_result_m15$estimates[3,1], waic_result_m15$estimates[1,1], waic_result_m15$estimates[2,1], waic_result_m15$estimates[3,1])))

#model_comparison[,3:8] <- as.numeric(model_comparison[,3:8])

colnames(model_comparison) <- colnames

#print(arrange(model_comparison, waic))

model_comparison
```
M15, which includes all 6 surface level climate variables, performs best, with the lowest WAIC score and highest loo score. 

```{r}

ggplot(data = data.frame(
  observed = test$strikes, predicted = simStrikes, 
  T=test$tair, SWR = test$swr, RH = test$rh, P = test$precip, W = test$wind, SP = test$sp)) + 
  geom_point(aes(x=T, y=predicted, color=SP)) + 
  scale_color_continuous(high="green", low="orange")

ggplot(data = data.frame(
  observed = test$strikes, predicted = simStrikes, 
  T=test$tair, SWR = test$swr, RH = test$rh, P = test$precip, W = test$wind, SP = test$sp)) + 
  geom_point(aes(x=T, y=observed, color=SP)) + 
  scale_color_continuous(high="green", low="orange")

```










