---
title: "Rstan Zero-Inflated Gamma Models: Weekly"
author: "Charlotte Uden"
date: "2024-11-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, include=FALSE}
library(rstan)
library(loo)
library(dplyr)
library(stats)
library(caret)
```
1. Read in the data


```{r, echo=FALSE}
data <- read.csv("/raid/cuden/data/era5WeeklySummaries_vaisalaLightningWeeklyRate_2005-2010_NEclip.csv") 
#select columns of interest - I'm going with strikes per day per kmm rather than strikes per hour per km 
data <- data[,c("lat","lon","strike_rate_daily","cape","cxp","precip","tair","wind","swr","sp","rh")]
colnames(data) <- c("lat","lon","strikes","cape","cxp","precip","tair","wind","swr","sp","rh")
hist(data$strikes, breaks = 30)
```

2. Split into training and testing data
```{r, echo=FALSE}
## 80% of the sample size
sample_size <- floor(0.8 * nrow(data))

## set the seed to make partition reproducible
set.seed(123)
train_index <- sample(seq_len(nrow(data)), size = sample_size)

train <- data[train_index, ]
test <- data[-train_index, ]

```


3. Standardize predictor variables 
```{r, echo=FALSE}
#save min and max (normalization parameters?):
#-----
mu_cxp <- mean(train$cxp)
sigma_cxp <- sd(train$cxp)
mu_cape <- mean(train$cape)
sigma_cape <- sd(train$cape)
mu_precip <- mean(train$precip)
sigma_precip <- sd(train$precip)
mu_tair <- mean(train$tair)
sigma_tair <- sd(train$tair)
mu_wind <- mean(train$wind)
sigma_wind <- sd(train$wind)
mu_swr <- mean(train$swr)
sigma_swr <- sd(train$swr)
mu_sp <- mean(train$sp)
sigma_sp <- sd(train$sp)
mu_rh <- mean(train$rh)
sigma_rh <- sd(train$rh)
#-----

train <- train %>%
  mutate(cape = (cape-mu_cape)/sigma_cape) %>%
  mutate(precip = (precip-mu_precip)/sigma_precip) %>%
  mutate(cxp = (cxp-mu_cxp)/sigma_cxp) %>%
  mutate(tair = (tair-mu_tair)/sigma_tair) %>%
  mutate(wind = (wind-mu_wind)/sigma_wind) %>%
  mutate(swr = (swr-mu_swr)/sigma_swr) %>%
  mutate(sp = (sp-mu_sp)/sigma_sp) %>%
  mutate(rh = (rh-mu_rh)/sigma_rh) 

train <- train[,c("strikes", "cape", "precip", "cxp", "tair", "wind", "swr", "sp", "rh")]
head(train)

#plot(train)

#use the min and max values from the train set to normalize the test set 
test <- test %>%
  mutate(cape = (cape-mu_cape)/sigma_cape) %>%
  mutate(precip = (precip-mu_precip)/sigma_precip) %>%
  mutate(cxp = (cxp-mu_cxp)/sigma_cxp) %>%
  mutate(tair = (tair-mu_tair)/sigma_tair) %>%
  mutate(wind = (wind-mu_wind)/sigma_wind) %>%
  mutate(swr = (swr-mu_swr)/sigma_swr) %>%
  mutate(sp = (sp-mu_sp)/sigma_sp) %>%
  mutate(rh = (rh-mu_rh)/sigma_rh) 

test <- test[,c("strikes", "cape", "precip", "cxp", "tair", "wind", "swr", "sp", "rh")]

```



```{r, echo=FALSE, results='hide', include=FALSE}

#library(rstan)

dat <- list(
  N = nrow(train),
  S = train$strikes
)

if (!file.exists("compiled_normal_weekly_m01.rds")) {
  m0.1 <- stan(file = "normal.stan", data = dat, 
           iter = 2000, 
           chains = 4, 
           cores = min(parallel::detectCores(), 4))
    saveRDS(m0.1, file = "compiled_normal_weekly_m01.rds")
    } else {
      m0.1 <- readRDS("compiled_normal_weekly_m01.rds")
    }

```


```{r}
print(m0.1, pars = c("mu", "sigma"))

traceplot(m0.1)

parameter_stats <- summary(m0.1)$summary[, c("mean", "sd")][1:2,1:2]

summary_table <- data.frame(
  parameter = rownames(parameter_stats),
  mean = parameter_stats[, "mean"],
  sd = parameter_stats[, "sd"]
)

ggplot(summary_table, aes(x = parameter, y = mean)) +
  geom_point(color = "blue", size = 3) +  # Point for the mean
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), width = 0.2) +  # Error bars for SD
  labs(y = "Parameter Estimate", x = "Parameter") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Parameter Means and SDs from Stan Model")

#predictions 
mean_mu <- summary_table["mu", "mean"]
mean_sigma <- summary_table["sigma", "mean"]

simStrikes <- rnorm(length(test$strikes), mean_mu, mean_sigma)

hist(simStrikes, breaks=30, main = "predictions")

hist(test$strikes, breaks=30, main = "observations")

```
Model Comparison in Rstan:
Loo: Leave one out cross validation
WAIC: widely applicable information criterion

```{r}
# Extract the log-likelihood
log_lik <- extract_log_lik(m0.1, parameter_name = "log_lik")

# Calculate WAIC
waic_result_m0.1 <- waic(log_lik)
# Calculate LOO
loo_result_m0.1 <- loo(log_lik)

# Print WAIC and Loo result
print(waic_result_m0.1)
print(loo_result_m0.1)

rmse_m0.1 <- RMSE(simStrikes, test$strikes)

param_table <- data.frame(scale = c("weekly"),
                          distribution = c("normal"),
                          model_name = c("m0.1"),
                          climate_var = NA,
                          alpha = mean_mu,
                          a_alpha = NA,
                          b_alpha = NA,
                          c_alpha = NA,
                          d_alpha = NA,
                          e_alpha = NA,
                          f_alpha = NA,
                          g_alpha = NA,
                          beta = mean_sigma,
                          a_beta = NA,
                          b_beta = NA,
                          c_beta = NA,
                          d_beta = NA,
                          e_beta = NA,
                          f_beta = NA,
                          g_beta = NA,
                          pi = NA,
                          a_pi = NA,
                          b_pi = NA,
                          c_pi = NA,
                          d_pi = NA,
                          e_pi = NA,
                          f_pi = NA,
                          g_pi = NA) 

```


```{r, echo=FALSE, results='hide', include=FALSE}

#library(rstan)

dat <- list(
  N = nrow(train),
  S = train$strikes
)

if (!file.exists("compiled_ZIG_weekly_m03.rds")) {
  m0.3 <- stan(file = "ZIG.stan", data = dat, 
           iter = 2000, 
           chains = 4, 
           cores = min(parallel::detectCores(), 4))
    saveRDS(m0.3, file = "compiled_ZIG_weekly_m03.rds")
    } else {
      m0.3 <- readRDS("compiled_ZIG_weekly_m03.rds")
    }


```


```{r}

print(m0.3, pars = c("alpha", "beta", "pi"))

traceplot(m0.3)

parameter_stats <- summary(m0.3)$summary[, c("mean", "sd")][1:3, 1:2]
print(parameter_stats)

summary_table <- data.frame(
  parameter = rownames(parameter_stats),
  mean = parameter_stats[, "mean"],
  sd = parameter_stats[, "sd"]
)

ggplot(summary_table, aes(x = parameter, y = mean)) +
  geom_point(color = "blue", size = 3) +  # Point for the mean
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), width = 0.2) +  # Error bars for SD
  labs(y = "Parameter Estimate", x = "Parameter") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Parameter Means and SDs from Stan Model")

#predictions 
mean_alpha <- summary_table["alpha", "mean"]
mean_beta <- summary_table["beta", "mean"]
mean_pi <- summary_table["pi", "mean"]

# Simulate zero-inflated gamma predictions
n_predictions <- length(test$strikes)

is_zero <- rbinom(n_predictions, size = 1, prob = mean_pi)  # 1 if zero, 0 if non-zero
simStrikes <- ifelse(
  is_zero == 1,  # If the binomial result is 1, it's a zero
  0,  # Zero value
  rgamma(n_predictions, mean_alpha, mean_beta)  # Non-zero values follow a gamma distribution
)

hist(simStrikes, breaks=30, main = "predictions")

hist(test$strikes, breaks=30, main = "observations")

```


```{r}
# Extract the log-likelihood
log_lik <- extract_log_lik(m0.3, parameter_name = "log_lik")

# Calculate WAIC
waic_result_m0.3 <- waic(log_lik)
# Calculate LOO
loo_result_m0.3 <- loo(log_lik)

# Print WAIC and Loo result
print(waic_result_m0.3)
print(loo_result_m0.3)

rmse_m0.3 <- RMSE(simStrikes, test$strikes)

#build table of parameters:
param_table <- rbind(param_table, data.frame(scale = c("weekly"),
                          distribution = c("ZIG"),
                          model_name = c("m0.3"),
                          climate_var = NA,
                          alpha = mean_alpha,
                          a_alpha = NA,
                          b_alpha = NA,
                          c_alpha = NA,
                          d_alpha = NA,
                          e_alpha = NA,
                          f_alpha = NA,
                          g_alpha = NA,
                          beta = mean_beta,
                          a_beta = NA,
                          b_beta = NA,
                          c_beta = NA,
                          d_beta = NA,
                          e_beta = NA,
                          f_beta = NA,
                          g_beta = NA,
                          pi = mean_pi,
                          a_pi = NA,
                          b_pi = NA,
                          c_pi = NA,
                          d_pi = NA,
                          e_pi = NA,
                          f_pi = NA,
                          g_pi = NA))

```


# M1: CAPE

S ~ gamma(alpha, beta);
alpha = a_alpha + b_alpha * CAPE;
beta = a_beta + b_beta * CAPE;

```{r, echo=FALSE, results='hide', include=FALSE}

#library(rstan)

dat <- list(
  N = nrow(train),
  S = train$strikes,
  CAPE = train$cape
)

if (!file.exists("compiled_ZIG_weekly_m1.rds")) {
  m1 <- stan(file = "ZIG_cape.stan", data = dat, 
           iter = 2000, 
           chains = 4, 
           cores = min(parallel::detectCores(), 4))
    saveRDS(m1, file = "compiled_ZIG_weekly_m1.rds")
    } else {
      m1 <- readRDS("compiled_ZIG_weekly_m1.rds")
    }

```


```{r}

print(m1, pars = c("a_alpha", "b_alpha", "a_beta", "b_beta", "a_pi", "b_pi"))

traceplot(m1)

parameter_stats <- summary(m1)$summary[, c("mean", "sd")][1:6, 1:2]
print(parameter_stats)

summary_table <- data.frame(
  parameter = rownames(parameter_stats),
  mean = parameter_stats[, "mean"],
  sd = parameter_stats[, "sd"]
)

ggplot(summary_table, aes(x = parameter, y = mean)) +
  geom_point(color = "blue", size = 3) +  # Point for the mean
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), width = 0.2) +  # Error bars for SD
  labs(y = "Parameter Estimate", x = "Parameter") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Parameter Means and SDs from Stan Model")

#predictions 
a_alpha <- summary_table["a_alpha", "mean"]
b_alpha <- summary_table["b_alpha", "mean"]
a_beta <- summary_table["a_beta", "mean"]
b_beta  <- summary_table["b_beta", "mean"]
a_pi <- summary_table["a_pi", "mean"]
b_pi  <- summary_table["b_pi", "mean"]

alpha <-  exp(a_alpha + b_alpha * test$cape)
beta <-  exp(a_beta + b_beta * test$cape)
pi <-  plogis(a_pi + b_pi * test$cape)

# Simulate zero-inflated gamma predictions
n_predictions <- length(test$strikes)

is_zero <- rbinom(n_predictions, size = 1, prob = pi)  
simStrikes <- ifelse(
  is_zero == 1,
  0,  
  rgamma(n_predictions, alpha, beta)  # Non-zero values follow a gamma distribution
)

hist(simStrikes, breaks=30, main = "predictions")

hist(test$strikes, breaks=30, main = "observations")

plot(test$cape, simStrikes)
plot(test$cape, test$strikes)
plot(simStrikes, test$strikes)
abline(a = 0, b = 1, col = "red", lty = 2, lwd = 2)

```


```{r}
# Extract the log-likelihood
log_lik <- extract_log_lik(m1, parameter_name = "log_lik")

# Calculate WAIC
waic_result_m1 <- waic(log_lik)
# Calculate LOO
loo_result_m1 <- loo(log_lik)

# Print WAIC and Loo result
print(waic_result_m1)
print(loo_result_m1)

rmse_m1 <- RMSE(simStrikes, test$strikes)

param_table <- rbind(param_table, data.frame(scale = c("weekly"),
                          distribution = c("ZIG"),
                          model_name = c("m1"),
                          climate_var = c("CAPE"),
                          alpha = c("a_alpha + b_alpha * CAPE"),
                          a_alpha = a_alpha,
                          b_alpha = b_alpha,
                          c_alpha = NA,
                          d_alpha = NA,
                          e_alpha = NA,
                          f_alpha = NA,
                          g_alpha = NA,
                          beta = c("a_beta + b_beta * CAPE"),
                          a_beta = a_beta,
                          b_beta = b_beta,
                          c_beta = NA,
                          d_beta = NA,
                          e_beta = NA,
                          f_beta = NA,
                          g_beta = NA,
                          pi = c("a_pi + b_pi * CAPE"),
                          a_pi = a_pi,
                          b_pi = b_pi,
                          c_pi = NA,
                          d_pi = NA,
                          e_pi = NA,
                          f_pi = NA,
                          g_pi = NA))

```

# M2: CAPE x Precipitation

S ~ gamma(alpha, beta);
alpha = a_alpha + b_alpha * CXP;
beta = a_beta + b_beta * CXP;

```{r, echo=FALSE, results='hide', include=FALSE}

#library(rstan)

dat <- list(
  N = nrow(train),
  S = train$strikes,
  CXP = train$cxp
)


# Compile and fit the Stan model
if (!file.exists("compiled_ZIG_weekly_m2.rds")) {
  m2 <- stan(file = "ZIG_cxp.stan", data = dat, 
           iter = 2000, 
           chains = 4, 
           cores = min(parallel::detectCores(), 4))
    saveRDS(m2, file = "compiled_ZIG_weekly_m2.rds")
    } else {
      m2 <- readRDS("compiled_ZIG_weekly_m2.rds")
    }
```


```{r}

print(m2, pars = c("a_alpha", "b_alpha", "a_beta", "b_beta", "a_pi", "b_pi"))

traceplot(m2)

parameter_stats <- summary(m2)$summary[, c("mean", "sd")][1:6, 1:2]
print(parameter_stats)

summary_table <- data.frame(
  parameter = rownames(parameter_stats),
  mean = parameter_stats[, "mean"],
  sd = parameter_stats[, "sd"]
)

ggplot(summary_table, aes(x = parameter, y = mean)) +
  geom_point(color = "blue", size = 3) +  # Point for the mean
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), width = 0.2) +  # Error bars for SD
  labs(y = "Parameter Estimate", x = "Parameter") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Parameter Means and SDs from Stan Model")

#predictions 
a_alpha <- summary_table["a_alpha", "mean"]
b_alpha <- summary_table["b_alpha", "mean"]
a_beta <- summary_table["a_beta", "mean"]
b_beta  <- summary_table["b_beta", "mean"]
a_pi <- summary_table["a_pi", "mean"]
b_pi  <- summary_table["b_pi", "mean"]

alpha <-  exp(a_alpha + b_alpha * test$cxp)
beta <-  exp(a_beta + b_beta * test$cxp)
pi <-  plogis(a_pi + b_pi * test$cxp)

# Simulate zero-inflated gamma predictions
n_predictions <- length(test$strikes)

is_zero <- rbinom(n_predictions, size = 1, prob = pi)  
simStrikes <- ifelse(
  is_zero == 1,
  0,  
  rgamma(n_predictions, alpha, beta)  # Non-zero values follow a gamma distribution
)

hist(simStrikes, breaks=30, main = "predictions")

hist(test$strikes, breaks=30, main = "observations")

plot(test$cxp, simStrikes)
plot(test$cxp, test$strikes)
plot(simStrikes, test$strikes)
abline(a = 0, b = 1, col = "red", lty = 2, lwd = 2)

```

```{r}
# Extract the log-likelihood
log_lik <- extract_log_lik(m2, parameter_name = "log_lik")

# Calculate WAIC
waic_result_m2 <- waic(log_lik)
# Calculate LOO
loo_result_m2 <- loo(log_lik)

# Print WAIC and Loo result
print(waic_result_m2)
print(loo_result_m2)

rmse_m2 <- RMSE(simStrikes, test$strikes)

param_table <- rbind(param_table, data.frame(scale = c("weekly"),
                          distribution = c("ZIG"),
                          model_name = c("m2"),
                          climate_var = c("CXP"),
                          alpha = c("a_alpha + b_alpha * CXP"),
                          a_alpha = a_alpha,
                          b_alpha = b_alpha,
                          c_alpha = NA,
                          d_alpha = NA,
                          e_alpha = NA,
                          f_alpha = NA,
                          g_alpha = NA,
                          beta = c("a_beta + b_beta * CXP"),
                          a_beta = a_beta,
                          b_beta = b_beta,
                          c_beta = NA,
                          d_beta = NA,
                          e_beta = NA,
                          f_beta = NA,
                          g_beta = NA,
                          pi = c("a_pi + b_pi * CXP"),
                          a_pi = a_pi,
                          b_pi = b_pi,
                          c_pi = NA,
                          d_pi = NA,
                          e_pi = NA,
                          f_pi = NA,
                          g_pi = NA))

```

# M3: Relative Humidity

S ~ gamma(alpha, beta);
alpha = a_alpha + b_alpha * RH;
beta = a_beta + b_beta * RH;

```{r, echo=FALSE, results='hide', include=FALSE}

dat <- list(
  N = nrow(train),
  S = train$strikes,
  RH = train$rh
)


# Compile and fit the Stan model
if (!file.exists("compiled_ZIG_weekly_m3.rds")) {
  m3 <- stan(file = "ZIG_rh.stan", data = dat, 
           iter = 2000, 
           chains = 4, 
           cores = min(parallel::detectCores(), 4))
    saveRDS(m3, file = "compiled_ZIG_weekly_m3.rds")
    } else {
      m3 <- readRDS("compiled_ZIG_weekly_m3.rds")
    }
```


```{r}

print(m3, pars = c("a_alpha", "b_alpha", "a_beta", "b_beta", "a_pi", "b_pi"))

traceplot(m3)

parameter_stats <- summary(m3)$summary[, c("mean", "sd")][1:6, 1:2]
print(parameter_stats)

summary_table <- data.frame(
  parameter = rownames(parameter_stats),
  mean = parameter_stats[, "mean"],
  sd = parameter_stats[, "sd"]
)

ggplot(summary_table, aes(x = parameter, y = mean)) +
  geom_point(color = "blue", size = 3) +  # Point for the mean
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), width = 0.2) +  # Error bars for SD
  labs(y = "Parameter Estimate", x = "Parameter") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Parameter Means and SDs from Stan Model")

#predictions 
a_alpha <- summary_table["a_alpha", "mean"]
b_alpha <- summary_table["b_alpha", "mean"]
a_beta <- summary_table["a_beta", "mean"]
b_beta  <- summary_table["b_beta", "mean"]
a_pi <- summary_table["a_pi", "mean"]
b_pi  <- summary_table["b_pi", "mean"]

alpha <-  exp(a_alpha + b_alpha * test$rh)
beta <-  exp(a_beta + b_beta * test$rh)
pi <-  plogis(a_pi + b_pi * test$rh)

plot(test$rh, pi)

# Simulate zero-inflated gamma predictions
n_predictions <- length(test$strikes)

is_zero <- rbinom(n_predictions, size = 1, prob = pi)  
simStrikes <- ifelse(
  is_zero == 1,
  0,  
  rgamma(n_predictions, alpha, beta)  # Non-zero values follow a gamma distribution
)

hist(simStrikes, breaks=30, main = "predictions")

hist(test$strikes, breaks=30, main = "observations")

plot(test$rh, simStrikes)
plot(test$rh, test$strikes)
plot(simStrikes, test$strikes)
abline(a = 0, b = 1, col = "red", lty = 2, lwd = 2)

```

```{r}
# Extract the log-likelihood
log_lik <- extract_log_lik(m3, parameter_name = "log_lik")

# Calculate WAIC
waic_result_m3 <- waic(log_lik)
# Calculate LOO
loo_result_m3 <- loo(log_lik)

# Print WAIC and Loo result
print(waic_result_m3)
print(loo_result_m3)

rmse_m3 <- RMSE(simStrikes, test$strikes)

param_table <- rbind(param_table, data.frame(scale = c("weekly"),
                          distribution = c("ZIG"),
                          model_name = c("m3"),
                          climate_var = c("RH"),
                          alpha = c("a_alpha + b_alpha * RH"),
                          a_alpha = a_alpha,
                          b_alpha = b_alpha,
                          c_alpha = NA,
                          d_alpha = NA,
                          e_alpha = NA,
                          f_alpha = NA,
                          g_alpha = NA,
                          beta = c("a_beta + b_beta * RH"),
                          a_beta = a_beta,
                          b_beta = b_beta,
                          c_beta = NA,
                          d_beta = NA,
                          e_beta = NA,
                          f_beta = NA,
                          g_beta = NA,
                          pi = c("a_pi + b_pi * RH"),
                          a_pi = a_pi,
                          b_pi = b_pi,
                          c_pi = NA,
                          d_pi = NA,
                          e_pi = NA,
                          f_pi = NA,
                          g_pi = NA))

```

# M4: Short Wave Radiation

S ~ gamma(alpha, beta);
alpha = a_alpha + b_alpha * SWR;
beta = a_beta + b_beta * SWR;

```{r, echo=FALSE, results='hide', include=FALSE}

#library(rstan)

dat <- list(
  N = nrow(train),
  S = train$strikes,
  SWR = train$swr
)

# init_function <- function() {
#   list(
#     a_alpha = 0.5,       # Initial guess for a_alpha
#     b_alpha = 0.1,     # Initial guess for b_alpha
#     a_beta = 1.5,        # Initial guess for a_beta
#     b_beta = 0.1,      # Initial guess for b_beta
#     a_pi = 0.5,        # Initial guess for a_pi (constrained to [0, 1])
#     b_pi = 0.1        # Initial guess for b_pi
#   )
# }

# Compile and fit the Stan model
if (!file.exists("compiled_ZIG_weekly_m4.rds")) {
  m4 <- stan(file = "ZIG_swr.stan", data = dat, 
           iter = 2000, 
           chains = 4, 
           cores = min(parallel::detectCores(), 4))
    saveRDS(m4, file = "compiled_ZIG_weekly_m4.rds")
    } else {
      m4 <- readRDS("compiled_ZIG_weekly_m4.rds")
    }

```


```{r}

print(m4, pars = c("a_alpha", "b_alpha", "a_beta", "b_beta"))

traceplot(m4)

parameter_stats <- summary(m4)$summary[, c("mean", "sd")][1:6, 1:2]
print(parameter_stats)

summary_table <- data.frame(
  parameter = rownames(parameter_stats),
  mean = parameter_stats[, "mean"],
  sd = parameter_stats[, "sd"]
)

ggplot(summary_table, aes(x = parameter, y = mean)) +
  geom_point(color = "blue", size = 3) +  # Point for the mean
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), width = 0.2) +  # Error bars for SD
  labs(y = "Parameter Estimate", x = "Parameter") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Parameter Means and SDs from Stan Model")

#predictions
a_alpha <- summary_table["a_alpha", "mean"]
b_alpha <- summary_table["b_alpha", "mean"]
a_beta <- summary_table["a_beta", "mean"]
b_beta  <- summary_table["b_beta", "mean"]
a_pi <- summary_table["a_pi", "mean"]
b_pi  <- summary_table["b_pi", "mean"]

alpha <-  exp(a_alpha + b_alpha * test$swr)
beta <-  exp(a_beta + b_beta * test$swr)
pi <-  plogis(a_pi + b_pi * test$swr)

# Simulate zero-inflated gamma predictions
n_predictions <- length(test$strikes)

is_zero <- rbinom(n_predictions, size = 1, prob = pi)  
simStrikes <- ifelse(
  is_zero == 1,
  0,  
  rgamma(n_predictions, alpha, beta)  # Non-zero values follow a gamma distribution
)


hist(simStrikes, breaks=30, main = "predictions")

hist(test$strikes, breaks=30, main = "observations")

plot(test$swr, simStrikes)
plot(test$swr, test$strikes)
plot(simStrikes, test$strikes)
abline(a = 0, b = 1, col = "red", lty = 2, lwd = 2)

```

```{r}
# Extract the log-likelihood
log_lik <- extract_log_lik(m4, parameter_name = "log_lik")

# Calculate WAIC
waic_result_m4 <- waic(log_lik)
# Calculate LOO
loo_result_m4 <- loo(log_lik)

# Print WAIC and Loo result
print(waic_result_m4)
print(loo_result_m4)

rmse_m4 <- RMSE(simStrikes, test$strikes)

param_table <- rbind(param_table, data.frame(scale = c("weekly"),
                          distribution = c("ZIG"),
                          model_name = c("m4"),
                          climate_var = c("SWR"),
                          alpha = c("a_alpha + b_alpha * SWR"),
                          a_alpha = a_alpha,
                          b_alpha = b_alpha,
                          c_alpha = NA,
                          d_alpha = NA,
                          e_alpha = NA,
                          f_alpha = NA,
                          g_alpha = NA,
                          beta = c("a_beta + b_beta * SWR"),
                          a_beta = a_beta,
                          b_beta = b_beta,
                          c_beta = NA,
                          d_beta = NA,
                          e_beta = NA,
                          f_beta = NA,
                          g_beta = NA,
                          pi = c("a_pi + b_pi * SWR"),
                          a_pi = a_pi,
                          b_pi = b_pi,
                          c_pi = NA,
                          d_pi = NA,
                          e_pi = NA,
                          f_pi = NA,
                          g_pi = NA))
```

# M5: Temperature

S ~ gamma(alpha, beta);
alpha = a_alpha + b_alpha * T;
beta = a_beta + b_beta * T;

```{r, echo=FALSE, results='hide', include=FALSE}

#library(rstan)

dat <- list(
  N = nrow(train),
  S = train$strikes,
  T = train$tair
)


# Compile and fit the Stan model
if (!file.exists("compiled_ZIG_weekly_m5.rds")) {
  m5 <- stan(file = "ZIG_tair.stan", data = dat, 
           iter = 2000, 
           chains = 4, 
           cores = min(parallel::detectCores(), 4))
    saveRDS(m5, file = "compiled_ZIG_weekly_m5.rds")
    } else {
      m5 <- readRDS("compiled_ZIG_weekly_m5.rds")
    }
```


```{r}

print(m5, pars = c("a_alpha", "b_alpha", "a_beta", "b_beta", "a_pi", "b_pi"))

traceplot(m5)

parameter_stats <- summary(m5)$summary[, c("mean", "sd")][1:6, 1:2]
print(parameter_stats)

summary_table <- data.frame(
  parameter = rownames(parameter_stats),
  mean = parameter_stats[, "mean"],
  sd = parameter_stats[, "sd"]
)

ggplot(summary_table, aes(x = parameter, y = mean)) +
  geom_point(color = "blue", size = 3) +  # Point for the mean
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), width = 0.2) +  # Error bars for SD
  labs(y = "Parameter Estimate", x = "Parameter") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Parameter Means and SDs from Stan Model")

#predictions 
a_alpha <- summary_table["a_alpha", "mean"]
b_alpha <- summary_table["b_alpha", "mean"]
a_beta <- summary_table["a_beta", "mean"]
b_beta  <- summary_table["b_beta", "mean"]
a_pi <- summary_table["a_pi", "mean"]
b_pi  <- summary_table["b_pi", "mean"]

alpha <-  exp(a_alpha + b_alpha * test$tair)
beta <-  exp(a_beta + b_beta * test$tair)
pi <-  plogis(a_pi + b_pi * test$tair)

# Simulate zero-inflated gamma predictions
n_predictions <- length(test$strikes)

is_zero <- rbinom(n_predictions, size = 1, prob = pi)  
simStrikes <- ifelse(
  is_zero == 1,
  0,  
  rgamma(n_predictions, alpha, beta)  # Non-zero values follow a gamma distribution
)


hist(simStrikes, breaks=30, main = "predictions")

hist(test$strikes, breaks=30, main = "observations")

plot(test$tair, simStrikes)
plot(test$tair, test$strikes)
plot(simStrikes, test$strikes)
abline(a = 0, b = 1, col = "red", lty = 2, lwd = 2)

```


```{r}
# Extract the log-likelihood
log_lik <- extract_log_lik(m5, parameter_name = "log_lik")

# Calculate WAIC
waic_result_m5 <- waic(log_lik)
# Calculate LOO
loo_result_m5 <- loo(log_lik)

# Print WAIC and Loo result
print(waic_result_m5)
print(loo_result_m5)

rmse_m5 <- RMSE(simStrikes, test$strikes)

param_table <- rbind(param_table, data.frame(scale = c("weekly"),
                          distribution = c("ZIG"),
                          model_name = c("m5"),
                          climate_var = c("T"),
                          alpha = c("a_alpha + b_alpha * T"),
                          a_alpha = a_alpha,
                          b_alpha = b_alpha,
                          c_alpha = NA,
                          d_alpha = NA,
                          e_alpha = NA,
                          f_alpha = NA,
                          g_alpha = NA,
                          beta = c("a_beta + b_beta * T"),
                          a_beta = a_beta,
                          b_beta = b_beta,
                          c_beta = NA,
                          d_beta = NA,
                          e_beta = NA,
                          f_beta = NA,
                          g_beta = NA,
                          pi = c("a_pi + b_pi * T"),
                          a_pi = a_pi,
                          b_pi = b_pi,
                          c_pi = NA,
                          d_pi = NA,
                          e_pi = NA,
                          f_pi = NA,
                          g_pi = NA))

```

# M6: Surface Pressure

S ~ gamma(alpha, beta);
alpha = a_alpha + b_alpha * SP;
beta = a_beta + b_beta * SP;

```{r, echo=FALSE, results='hide', include=FALSE}

#library(rstan)

dat <- list(
  N = nrow(train),
  S = train$strikes,
  SP = train$sp
)


# Compile and fit the Stan model
if (!file.exists("compiled_ZIG_weekly_m6.rds")) {
  m6 <- stan(file = "ZIG_sp.stan", data = dat, 
           iter = 2000, 
           chains = 4, 
           cores = min(parallel::detectCores(), 4))
    saveRDS(m6, file = "compiled_ZIG_weekly_m6.rds")
    } else {
      m6 <- readRDS("compiled_ZIG_weekly_m6.rds")
    }
```


```{r}

print(m6, pars = c("a_alpha", "b_alpha", "a_beta", "b_beta"))

traceplot(m6)

parameter_stats <- summary(m6)$summary[, c("mean", "sd")][1:6, 1:2]
print(parameter_stats)

summary_table <- data.frame(
  parameter = rownames(parameter_stats),
  mean = parameter_stats[, "mean"],
  sd = parameter_stats[, "sd"]
)

ggplot(summary_table, aes(x = parameter, y = mean)) +
  geom_point(color = "blue", size = 3) +  # Point for the mean
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), width = 0.2) +  # Error bars for SD
  labs(y = "Parameter Estimate", x = "Parameter") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Parameter Means and SDs from Stan Model")

#predictions 
a_alpha <- summary_table["a_alpha", "mean"]
b_alpha <- summary_table["b_alpha", "mean"]
a_beta <- summary_table["a_beta", "mean"]
b_beta  <- summary_table["b_beta", "mean"]
a_pi <- summary_table["a_pi", "mean"]
b_pi  <- summary_table["b_pi", "mean"]

alpha <-  exp(a_alpha + b_alpha * test$sp)
beta <-  exp(a_beta + b_beta * test$sp)
pi <-  plogis(a_pi + b_pi * test$sp)

# Simulate zero-inflated gamma predictions
n_predictions <- length(test$strikes)

is_zero <- rbinom(n_predictions, size = 1, prob = pi)  
simStrikes <- ifelse(
  is_zero == 1,
  0,  
  rgamma(n_predictions, alpha, beta)  # Non-zero values follow a gamma distribution
)


hist(simStrikes, breaks=30, main = "predictions")

hist(test$strikes, breaks=30, main = "observations")

plot(test$sp, simStrikes)
plot(test$sp, test$strikes)
plot(simStrikes, test$strikes)
abline(a = 0, b = 1, col = "red", lty = 2, lwd = 2)

```

```{r}
# Extract the log-likelihood
log_lik <- extract_log_lik(m6, parameter_name = "log_lik")

# Calculate WAIC
waic_result_m6 <- waic(log_lik)
# Calculate LOO
loo_result_m6 <- loo(log_lik)

# Print WAIC and Loo result
print(waic_result_m6)
print(loo_result_m6)

rmse_m6 <- RMSE(simStrikes, test$strikes)

param_table <- rbind(param_table, data.frame(scale = c("weekly"),
                          distribution = c("ZIG"),
                          model_name = c("m6"),
                          climate_var = c("SP"),
                          alpha = c("a_alpha + b_alpha * SP"),
                          a_alpha = a_alpha,
                          b_alpha = b_alpha,
                          c_alpha = NA,
                          d_alpha = NA,
                          e_alpha = NA,
                          f_alpha = NA,
                          g_alpha = NA,
                          beta = c("a_beta + b_beta * SP"),
                          a_beta = a_beta,
                          b_beta = b_beta,
                          c_beta = NA,
                          d_beta = NA,
                          e_beta = NA,
                          f_beta = NA,
                          g_beta = NA,
                          pi = c("a_pi + b_pi * SP"),
                          a_pi = a_pi,
                          b_pi = b_pi,
                          c_pi = NA,
                          d_pi = NA,
                          e_pi = NA,
                          f_pi = NA,
                          g_pi = NA))

```


# M7: Precipitation

S ~ gamma(alpha, beta);
alpha = a_alpha + b_alpha * P;
beta = a_beta + b_beta * P;

```{r, echo=FALSE, results='hide', include=FALSE}

#library(rstan)

dat <- list(
  N = nrow(train),
  S = train$strikes,
  P = train$precip
)


# Compile and fit the Stan model
if (!file.exists("compiled_ZIG_weekly_m7.rds")) {
  m7 <- stan(file = "ZIG_precip.stan", data = dat, 
           iter = 2000, 
           chains = 4, 
           cores = min(parallel::detectCores(), 4))
    saveRDS(m7, file = "compiled_ZIG_weekly_m7.rds")
    } else {
      m7 <- readRDS("compiled_ZIG_weekly_m7.rds")
    }
```


```{r}

print(m7, pars = c("a_alpha", "b_alpha", "a_beta", "b_beta"))

traceplot(m7)

parameter_stats <- summary(m7)$summary[, c("mean", "sd")][1:6, 1:2]
print(parameter_stats)

summary_table <- data.frame(
  parameter = rownames(parameter_stats),
  mean = parameter_stats[, "mean"],
  sd = parameter_stats[, "sd"]
)

ggplot(summary_table, aes(x = parameter, y = mean)) +
  geom_point(color = "blue", size = 3) +  # Point for the mean
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), width = 0.2) +  # Error bars for SD
  labs(y = "Parameter Estimate", x = "Parameter") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Parameter Means and SDs from Stan Model")

#predictions 
a_alpha <- summary_table["a_alpha", "mean"]
b_alpha <- summary_table["b_alpha", "mean"]
a_beta <- summary_table["a_beta", "mean"]
b_beta  <- summary_table["b_beta", "mean"]
a_pi <- summary_table["a_pi", "mean"]
b_pi  <- summary_table["b_pi", "mean"]

alpha <-  exp(a_alpha + b_alpha * test$precip)
beta <-  exp(a_beta + b_beta * test$precip)
pi <-  plogis(a_pi + b_pi * test$precip)

# Simulate zero-inflated gamma predictions
n_predictions <- length(test$strikes)

is_zero <- rbinom(n_predictions, size = 1, prob = pi)  
simStrikes <- ifelse(
  is_zero == 1,
  0,  
  rgamma(n_predictions, alpha, beta)  # Non-zero values follow a gamma distribution
)


hist(simStrikes, breaks=30, main = "predictions")

hist(test$strikes, breaks=30, main = "observations")

plot(test$precip, simStrikes)
plot(test$precip, test$strikes)
plot(simStrikes, test$strikes)
abline(a = 0, b = 1, col = "red", lty = 2, lwd = 2)

```

```{r}
# Extract the log-likelihood
log_lik <- extract_log_lik(m7, parameter_name = "log_lik")

# Calculate WAIC
waic_result_m7 <- waic(log_lik)
# Calculate LOO
loo_result_m7 <- loo(log_lik)

# Print WAIC and Loo result
print(waic_result_m7)
print(loo_result_m7)

rmse_m7 <- RMSE(simStrikes, test$strikes)

param_table <- rbind(param_table, data.frame(scale = c("weekly"),
                          distribution = c("ZIG"),
                          model_name = c("m7"),
                          climate_var = c("P"),
                          alpha = c("a_alpha + b_alpha * P"),
                          a_alpha = a_alpha,
                          b_alpha = b_alpha,
                          c_alpha = NA,
                          d_alpha = NA,
                          e_alpha = NA,
                          f_alpha = NA,
                          g_alpha = NA,
                          beta = c("a_beta + b_beta * P"),
                          a_beta = a_beta,
                          b_beta = b_beta,
                          c_beta = NA,
                          d_beta = NA,
                          e_beta = NA,
                          f_beta = NA,
                          g_beta = NA,
                          pi = c("a_pi + b_pi * P"),
                          a_pi = a_pi,
                          b_pi = b_pi,
                          c_pi = NA,
                          d_pi = NA,
                          e_pi = NA,
                          f_pi = NA,
                          g_pi = NA))

```

# M8: Wind

S ~ gamma(alpha, beta);
alpha = a_alpha + b_alpha * W;
beta = a_beta + b_beta * W;

```{r, echo=FALSE, results='hide', include=FALSE}

#library(rstan)

dat <- list(
  N = nrow(train),
  S = train$strikes,
  W = train$wind
)


# Compile and fit the Stan model
if (!file.exists("compiled_ZIG_weekly_m8.rds")) {
  m8 <- stan(file = "ZIG_wind.stan", data = dat, 
           iter = 2000, 
           chains = 4, 
           cores = min(parallel::detectCores(), 4))
    saveRDS(m8, file = "compiled_ZIG_weekly_m8.rds")
    } else {
      m8 <- readRDS("compiled_ZIG_weekly_m8.rds")
    }
```


```{r}

print(m8, pars = c("a_alpha", "b_alpha", "a_beta", "b_beta", "a_pi", "b_pi"))

traceplot(m8)

parameter_stats <- summary(m8)$summary[, c("mean", "sd")][1:6, 1:2]
print(parameter_stats)

summary_table <- data.frame(
  parameter = rownames(parameter_stats),
  mean = parameter_stats[, "mean"],
  sd = parameter_stats[, "sd"]
)

ggplot(summary_table, aes(x = parameter, y = mean)) +
  geom_point(color = "blue", size = 3) +  # Point for the mean
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), width = 0.2) +  # Error bars for SD
  labs(y = "Parameter Estimate", x = "Parameter") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Parameter Means and SDs from Stan Model")

#predictions 
a_alpha <- summary_table["a_alpha", "mean"]
b_alpha <- summary_table["b_alpha", "mean"]
a_beta <- summary_table["a_beta", "mean"]
b_beta  <- summary_table["b_beta", "mean"]
a_pi <- summary_table["b_pi", "mean"]
b_pi  <- summary_table["b_pi", "mean"]

alpha <-  exp(a_alpha + b_alpha * test$wind)
beta <-  exp(a_beta + b_beta * test$wind)
pi <-  plogis(a_pi + b_pi * test$wind)

# Simulate zero-inflated gamma predictions
n_predictions <- length(test$strikes)

is_zero <- rbinom(n_predictions, size = 1, prob = pi)  
simStrikes <- ifelse(
  is_zero == 1,
  0,  
  rgamma(n_predictions, alpha, beta)  # Non-zero values follow a gamma distribution
)


hist(simStrikes, breaks=30, main = "predictions")

hist(test$strikes, breaks=30, main = "observations")

plot(test$wind, simStrikes)
plot(test$wind, test$strikes)
plot(simStrikes, test$strikes)
abline(a = 0, b = 1, col = "red", lty = 2, lwd = 2)

```

```{r}
# Extract the log-likelihood
log_lik <- extract_log_lik(m8, parameter_name = "log_lik")

# Calculate WAIC
waic_result_m8 <- waic(log_lik)
# Calculate LOO
loo_result_m8 <- loo(log_lik)

# Print WAIC and Loo result
print(waic_result_m8)
print(loo_result_m8)

rmse_m8 <- RMSE(simStrikes, test$strikes)

param_table <- rbind(param_table, data.frame(scale = c("weekly"),
                          distribution = c("ZIG"),
                          model_name = c("m8"),
                          climate_var = c("W"),
                          alpha = c("a_alpha + b_alpha * W"),
                          a_alpha = a_alpha,
                          b_alpha = b_alpha,
                          c_alpha = NA,
                          d_alpha = NA,
                          e_alpha = NA,
                          f_alpha = NA,
                          g_alpha = NA,
                          beta = c("a_beta + b_beta * W"),
                          a_beta = a_beta,
                          b_beta = b_beta,
                          c_beta = NA,
                          d_beta = NA,
                          e_beta = NA,
                          f_beta = NA,
                          g_beta = NA,
                          pi = c("a_pi + b_pi * W"),
                          a_pi = a_pi,
                          b_pi = b_pi,
                          c_pi = NA,
                          d_pi = NA,
                          e_pi = NA,
                          f_pi = NA,
                          g_pi = NA))

```

```{r}
colnames <- c("model", "climate_var", "elpd_loo", "p_loo", "looic", "elpd_waic", "p_waic", "waic", "rmse")
model_comparison <- data.frame(rbind(c("m0.1", "normal", loo_result_m0.1$estimates[1,1], loo_result_m0.1$estimates[2,1], loo_result_m0.1$estimate[3,1], waic_result_m0.1$estimates[1,1], waic_result_m0.1$estimates[2,1], waic_result_m0.1$estimates[3,1], rmse_m0.1),
      c("m0.3", "gamma", loo_result_m0.3$estimates[1,1], loo_result_m0.3$estimates[2,1], loo_result_m0.3$estimate[3,1], waic_result_m0.3$estimates[1,1], waic_result_m2$estimates[2,1], waic_result_m2$estimates[3,1], rmse_m0.3),
  c("m1", "CAPE", loo_result_m1$estimates[1,1], loo_result_m1$estimates[2,1], loo_result_m1$estimate[3,1], waic_result_m1$estimates[1,1], waic_result_m1$estimates[2,1], waic_result_m1$estimates[3,1], rmse_m1),
  c("m2", "CXP", loo_result_m2$estimates[1,1], loo_result_m2$estimates[2,1], loo_result_m2$estimate[3,1], waic_result_m2$estimates[1,1], waic_result_m2$estimates[2,1], waic_result_m2$estimates[3,1], rmse_m2),
  c("m3", "rh", loo_result_m3$estimates[1,1], loo_result_m3$estimates[2,1], loo_result_m3$estimate[3,1], waic_result_m3$estimates[1,1], waic_result_m3$estimates[2,1], waic_result_m3$estimates[3,1], rmse_m3),
  c("m4", "swr", loo_result_m4$estimates[1,1], loo_result_m4$estimates[2,1], loo_result_m4$estimate[3,1], waic_result_m4$estimates[1,1], waic_result_m4$estimates[2,1], waic_result_m4$estimates[3,1], rmse_m4),
  c("m5", "tair", loo_result_m5$estimates[1,1], loo_result_m5$estimates[2,1], loo_result_m5$estimate[3,1], waic_result_m5$estimates[1,1], waic_result_m5$estimates[2,1], waic_result_m5$estimates[3,1], rmse_m5),
  c("m6", "sp", loo_result_m6$estimates[1,1], loo_result_m6$estimates[2,1], loo_result_m6$estimate[3,1], waic_result_m6$estimates[1,1], waic_result_m6$estimates[2,1], waic_result_m6$estimates[3,1], rmse_m6),
  c("m7", "precip", loo_result_m7$estimates[1,1], loo_result_m7$estimates[2,1], loo_result_m7$estimate[3,1], waic_result_m7$estimates[1,1], waic_result_m7$estimates[2,1], waic_result_m7$estimates[3,1], rmse_m7),
  c("m8", "wind", loo_result_m8$estimates[1,1], loo_result_m8$estimates[2,1], loo_result_m8$estimates[3,1], waic_result_m8$estimates[1,1], waic_result_m8$estimates[2,1], waic_result_m8$estimates[3,1], rmse_m8)))

#model_comparison[,3:8] <- as.numeric(model_comparison[,3:8])

colnames(model_comparison) <- colnames

#print(arrange(model_comparison, waic))

model_comparison
```
The highest elpd_loo scores are cape and cape x precip. of the surface level climate variables, short wave radiation and temperature perform best, followed by relative humidity, wind, precipitation, and surface pressure. WAIC scores also indicate that short wave radiation and temperature are the best surface level climate predictors of lighting. Interestingly, the simple gamma model (m0.3) performs even better than models that include climate. Perhaps this is because WAIC takes number of parameters into cnsiteration. 

# M9: Short Wave Radiation + Temperature

Lightning strike rate modeled from a normal distribution where mu and sigma are dependent on climate: shortwave radiation and temperature, where SWR and temp have additive affects. They contribute independently to the mean lightning strike rate and standard deviation. swr and temp do not interact; ie. the effect of SWR on lighting is the same regardless of the temperature outside. 

    S ~ dnorm( mu , sigma )
    mu <- a_alpha + b_alpha * SWR + c_alpha * T
    sigma <- a_beta + b_beta * SWR + c_beta * T


```{r, echo=FALSE, results='hide', include=FALSE}

#library(rstan)

dat <- list(
  N = nrow(train),
  S = train$strikes,
  SWR = train$swr,
  T = train$tair
)


# Compile and fit the Stan model
if (!file.exists("compiled_ZIG_weekly_m9.rds")) {
  m9 <- stan(file = "ZIG_swr_tair.stan", data = dat, 
           iter = 2000, 
           chains = 4, 
           cores = min(parallel::detectCores(), 4))
    saveRDS(m9, file = "compiled_ZIG_weekly_m9.rds")
    } else {
      m9 <- readRDS("compiled_ZIG_weekly_m9.rds")
    }
```


```{r}

print(m9, pars = c("a_alpha", "b_alpha", "c_alpha", "a_beta", "b_beta", "c_beta", "a_pi", "b_pi", "c_pi"))

traceplot(m9)

parameter_stats <- summary(m9)$summary[, c("mean", "sd")][1:9, 1:2]
print(parameter_stats)

summary_table <- data.frame(
  parameter = rownames(parameter_stats),
  mean = parameter_stats[, "mean"],
  sd = parameter_stats[, "sd"]
)

ggplot(summary_table, aes(x = parameter, y = mean)) +
  geom_point(color = "blue", size = 3) +  # Point for the mean
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), width = 0.2) +  # Error bars for SD
  labs(y = "Parameter Estimate", x = "Parameter") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Parameter Means and SDs from Stan Model")

#predictions 
a_alpha <- summary_table["a_alpha", "mean"]
b_alpha <- summary_table["b_alpha", "mean"]
c_alpha <- summary_table["c_alpha", "mean"]
a_beta  <- summary_table["a_beta", "mean"]
b_beta <- summary_table["b_beta", "mean"]
c_beta  <- summary_table["c_beta", "mean"]
a_pi  <- summary_table["a_pi", "mean"]
b_pi <- summary_table["b_pi", "mean"]
c_pi  <- summary_table["c_pi", "mean"]

alpha <-  exp(a_alpha + b_alpha * test$swr + c_alpha * test$tair)
beta <-  exp(a_beta + b_beta * test$swr + c_beta * test$tair)
pi <-  plogis(a_pi + b_pi * test$swr + c_pi * test$tair)

# Simulate zero-inflated gamma predictions
n_predictions <- length(test$strikes)

is_zero <- rbinom(n_predictions, size = 1, prob = pi)  
simStrikes <- ifelse(
  is_zero == 1,
  0,  
  rgamma(n_predictions, alpha, beta)  # Non-zero values follow a gamma distribution
)


hist(simStrikes, breaks=30, main = "predictions")

hist(test$strikes, breaks=30, main = "observations")

ggplot(data = data.frame(observed = test$strikes, predicted = simStrikes, T=test$tair, SWR = test$swr, RH = test$rh, P = test$precip)) + geom_point(aes(x=SWR, y=predicted, color=T))
ggplot(data = data.frame(observed = test$strikes, predicted = simStrikes, T=test$tair, SWR = test$swr, RH = test$rh, P = test$precip)) + geom_point(aes(x=SWR, y=observed, color=T))

plot(simStrikes, test$strikes)
abline(a = 0, b = 1, col = "red", lty = 2, lwd = 2)

```

```{r}
# Extract the log-likelihood
log_lik <- extract_log_lik(m9, parameter_name = "log_lik")

# Calculate WAIC
waic_result_m9 <- waic(log_lik)
# Calculate LOO
loo_result_m9 <- loo(log_lik)

# Print WAIC and Loo result
print(waic_result_m9)
print(loo_result_m9)

rmse_m9 <- RMSE(simStrikes, test$strikes)

param_table <- rbind(param_table, data.frame(scale = c("weekly"),
                          distribution = c("ZIG"),
                          model_name = c("m9"),
                          climate_var = c("SWR, T"),
                          alpha = c("a_alpha + b_alpha * SWR + c_alpha * T"),
                          a_alpha = a_alpha,
                          b_alpha = b_alpha,
                          c_alpha = c_alpha,
                          d_alpha = NA,
                          e_alpha = NA,
                          f_alpha = NA,
                          g_alpha = NA,
                          beta = c("a_beta + b_beta * SWR + c_beta * T"),
                          a_beta = a_beta,
                          b_beta = b_beta,
                          c_beta = c_beta,
                          d_beta = NA,
                          e_beta = NA,
                          f_beta = NA,
                          g_beta = NA,
                          pi = c("a_pi + b_pi * SWR + c_pi * T"),
                          a_pi = a_pi,
                          b_pi = b_pi,
                          c_pi = c_pi,
                          d_pi = NA,
                          e_pi = NA,
                          f_pi = NA,
                          g_pi = NA))

```
This model performs a little bit better than the simple short wave radiation model (m4). 

# M12: Short Wave Radiation + Temperature + Precipitation

Lightning strike rate modeled from a normal distribution where mu and sigma are dependent on climate: shortwave radiation and temperature, where SWR and temp have additive affects. They contribute independently to the mean lightning strike rate and standard deviation. swr and temp do not interact; ie. the effect of SWR on lighting is the same regardless of the temperature outside. 

    S ~ dnorm( mu , sigma ),
    mu <- a_alpha + b_alpha * SWR + c_alpha * T + d_alpha * RH,
    sigma <- a_beta + b_beta * SWR + c_beta * T+ d_beta * RH,
    
```{r, echo=FALSE, results='hide', include=FALSE}

#library(rstan)

dat <- list(
  N = nrow(train),
  S = train$strikes,
  SWR = train$swr,
  T = train$tair,
  RH = train$rh
)

# init_list <- function() {
#   list(a_alpha = 1, b_alpha = 0.1, c_alpha = 0.1, d_alpha = 0.1,
#        a_beta = 0.1, b_beta = 0.1, c_beta = 0.1, d_beta = 0.1, 
#        a_pi = 0.1, b_pi = 0.1, c_pi = 0.1, d_pi = 0.1)
# }

# Compile and fit the Stan model
if (!file.exists("compiled_ZIG_weekly_m12.rds")) {
  m12 <- stan(file = "ZIG_swr_tair_rh.stan", data = dat, 
           iter = 2000, 
           chains = 4,
           cores = min(parallel::detectCores(), 4))
    saveRDS(m12, file = "compiled_ZIG_weekly_m12.rds")
    } else {
      m12 <- readRDS("compiled_ZIG_weekly_m12.rds")
    }
```


```{r}

print(m12, pars = c("a_alpha", "b_alpha", "c_alpha", "d_alpha", "a_beta", "b_beta", "c_beta", "d_beta", "a_pi", "b_pi", "c_pi", "d_pi"))

traceplot(m12)

parameter_stats <- summary(m12)$summary[, c("mean", "sd")][1:12, 1:2]
print(parameter_stats)

summary_table <- data.frame(
  parameter = rownames(parameter_stats),
  mean = parameter_stats[, "mean"],
  sd = parameter_stats[, "sd"]
)

ggplot(summary_table, aes(x = parameter, y = mean)) +
  geom_point(color = "blue", size = 3) +  # Point for the mean
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), width = 0.2) +  # Error bars for SD
  labs(y = "Parameter Estimate", x = "Parameter") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Parameter Means and SDs from Stan Model")

#predictions 
a_alpha <- summary_table["a_alpha", "mean"]
b_alpha <- summary_table["b_alpha", "mean"]
c_alpha <- summary_table["c_alpha", "mean"]
d_alpha <- summary_table["d_alpha", "mean"]
a_beta  <- summary_table["a_beta", "mean"]
b_beta <- summary_table["b_beta", "mean"]
c_beta  <- summary_table["c_beta", "mean"]
d_beta  <- summary_table["d_beta", "mean"]
a_pi <- summary_table["a_pi", "mean"]
b_pi <- summary_table["b_pi", "mean"]
c_pi <- summary_table["c_pi", "mean"]
d_pi <- summary_table["d_pi", "mean"]

alpha <-  exp(a_alpha + b_alpha * test$swr + c_alpha * test$tair + d_alpha * test$rh)
beta <-  exp(a_beta + b_beta * test$swr + c_beta * test$tair + d_beta * test$rh)
pi <-  plogis(a_pi + b_pi * test$swr + c_pi * test$tair + d_pi * test$rh)

# Simulate zero-inflated gamma predictions
n_predictions <- length(test$strikes)

is_zero <- rbinom(n_predictions, size = 1, prob = pi)  
simStrikes <- ifelse(
  is_zero == 1,
  0,  
  rgamma(n_predictions, alpha, beta)  # Non-zero values follow a gamma distribution
)


hist(simStrikes, breaks=30, main = "predictions")

hist(test$strikes, breaks=30, main = "observations")

ggplot(data = data.frame(observed = test$strikes, predicted = simStrikes, T=test$tair, SWR = test$swr, RH = test$rh, P = test$precip)) + geom_point(aes(x=SWR, y=predicted, color=RH))
ggplot(data = data.frame(observed = test$strikes, predicted = simStrikes, T=test$tair, SWR = test$swr, RH = test$rh, P = test$precip)) + geom_point(aes(x=SWR, y=observed, color=RH))

plot(simStrikes, test$strikes)
abline(a = 0, b = 1, col = "red", lty = 2, lwd = 2)

```


```{r}
# Extract the log-likelihood
log_lik <- extract_log_lik(m12, parameter_name = "log_lik")

# Calculate WAIC
waic_result_m12 <- waic(log_lik)
# Calculate LOO
loo_result_m12 <- loo(log_lik)

# Print WAIC and Loo result
print(waic_result_m12)
print(loo_result_m12)

rmse_m12 <- RMSE(simStrikes, test$strikes)

param_table <- rbind(param_table, data.frame(scale = c("weekly"),
                          distribution = c("ZIG"),
                          model_name = c("m12"),
                          climate_var = c("SWR, T, RH"),
                          alpha = c("a_alpha + b_alpha * SWR + c_alpha * T + d_alpha * RH"),
                          a_alpha = a_alpha,
                          b_alpha = b_alpha,
                          c_alpha = c_alpha,
                          d_alpha = d_alpha,
                          e_alpha = NA,
                          f_alpha = NA,
                          g_alpha = NA,
                          beta = c("a_beta + b_beta * SWR + c_beta * T + d_beta * RH"),
                          a_beta = a_beta,
                          b_beta = b_beta,
                          c_beta = c_beta,
                          d_beta = d_beta,
                          e_beta = NA,
                          f_beta = NA,
                          g_beta = NA,
                          pi = ("a_pi + b_pi * SWR + c_pi * T + d_pi * RH"),
                          a_pi = a_pi,
                          b_pi = b_pi,
                          c_pi = c_pi,
                          d_pi = d_pi,
                          e_pi = NA,
                          f_pi = NA,
                          g_pi = NA))


```


# M13: Short Wave Radiation + Temperature + Precipitation + Wind

Lightning strike rate modeled from a normal distribution where mu and sigma are dependent on climate: shortwave radiation,  temperature, relative humidity, and wind. 
    S ~ dnorm( mu , sigma ),
    mu <- a_alpha + b_alpha * SWR + c_alpha * T + d_alpha * RH + e_alpha * W,
    sigma <- a_beta + b_beta * SWR + c_beta * T + d_beta * RH + e_beta * W,,
    
```{r, echo=FALSE, results='hide', include=FALSE}

#library(rstan)

dat <- list(
  N = nrow(train),
  S = train$strikes,
  SWR = train$swr,
  T = train$tair,
  RH = train$rh,
  W = train$wind
)


# Compile and fit the Stan model
if (!file.exists("compiled_ZIG_weekly_m13.rds")) {
  m13 <- stan(file = "ZIG_swr_tair_rh_wind.stan", data = dat, 
           iter = 2000, 
           chains = 4, 
           cores = min(parallel::detectCores(), 4))
    saveRDS(m13, file = "compiled_ZIG_weekly_m13.rds")
    } else {
      m13 <- readRDS("compiled_ZIG_weekly_m13.rds")
    }
```


```{r}

print(m13, pars = c("a_alpha", "b_alpha", "c_alpha", "d_alpha", "e_alpha", "a_beta", "b_beta", "c_beta", "d_beta", "e_beta", "a_pi", "b_pi", "c_pi", "d_pi", "e_pi"))

traceplot(m13)

parameter_stats <- summary(m13)$summary[, c("mean", "sd")][1:15, 1:2]
print(parameter_stats)

summary_table <- data.frame(
  parameter = rownames(parameter_stats),
  mean = parameter_stats[, "mean"],
  sd = parameter_stats[, "sd"]
)

ggplot(summary_table, aes(x = parameter, y = mean)) +
  geom_point(color = "blue", size = 3) +  # Point for the mean
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), width = 0.2) +  # Error bars for SD
  labs(y = "Parameter Estimate", x = "Parameter") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Parameter Means and SDs from Stan Model")

#predictions 
a_alpha <- summary_table["a_alpha", "mean"]
b_alpha <- summary_table["b_alpha", "mean"]
c_alpha <- summary_table["c_alpha", "mean"]
d_alpha <- summary_table["d_alpha", "mean"]
e_alpha <- summary_table["e_alpha", "mean"]
a_beta <- summary_table["a_beta", "mean"]
b_beta <- summary_table["b_beta", "mean"]
c_beta <- summary_table["c_beta", "mean"]
d_beta <- summary_table["d_beta", "mean"]
e_beta <- summary_table["e_beta", "mean"]
a_pi <- summary_table["a_pi", "mean"]
b_pi <- summary_table["b_pi", "mean"]
c_pi <- summary_table["c_pi", "mean"]
d_pi <- summary_table["d_pi", "mean"]
e_pi <- summary_table["e_pi", "mean"]

alpha <-  exp(a_alpha + b_alpha * test$swr + c_alpha * test$tair + d_alpha * test$rh + e_alpha * test$wind)
beta <-  exp(a_beta + b_beta * test$swr + c_beta * test$tair + d_beta * test$rh + e_beta * test$wind)
pi <-  plogis(a_pi + b_pi * test$swr + c_pi * test$tair + d_pi * test$rh + e_pi * test$wind)

# Simulate zero-inflated gamma predictions
n_predictions <- length(test$strikes)

is_zero <- rbinom(n_predictions, size = 1, prob = pi)  
simStrikes <- ifelse(
  is_zero == 1,
  0,  
  rgamma(n_predictions, alpha, beta)  # Non-zero values follow a gamma distribution
)


hist(simStrikes, breaks=30, main = "predictions")

hist(test$strikes, breaks=30, main = "observations")

ggplot(data = data.frame(observed = test$strikes, predicted = simStrikes, T=test$tair, SWR = test$swr, RH = test$rh, P = test$precip, W = test$wind)) + geom_point(aes(x=SWR, y=predicted, color=W))
ggplot(data = data.frame(observed = test$strikes, predicted = simStrikes, T=test$tair, SWR = test$swr, RH = test$rh, P = test$precip, W = test$wind)) + geom_point(aes(x=SWR, y=observed, color=W))

plot(simStrikes, test$strikes)
abline(a = 0, b = 1, col = "red", lty = 2, lwd = 2)

```


```{r}
# Extract the log-likelihood
log_lik <- extract_log_lik(m13, parameter_name = "log_lik")

# Calculate WAIC
waic_result_m13 <- waic(log_lik)
# Calculate LOO
loo_result_m13 <- loo(log_lik)

# Print WAIC and Loo result
print(waic_result_m13)
print(loo_result_m13)

rmse_m13 <- RMSE(simStrikes, test$strikes)

param_table <- rbind(param_table, data.frame(scale = c("weekly"),
                          distribution = c("ZIG"),
                          model_name = c("m13"),
                          climate_var = c("SWR, T, RH, W"),
                          alpha = c("a_alpha + b_alpha * SWR + c_alpha * T + d_alpha * RH + e_alpha * W"),
                          a_alpha = a_alpha,
                          b_alpha = b_alpha,
                          c_alpha = c_alpha,
                          d_alpha = d_alpha,
                          e_alpha = e_alpha,
                          f_alpha = NA,
                          g_alpha = NA,
                          beta = c("a_beta + b_beta * SWR + c_beta * T + d_beta * RH + e_beta * W"),
                          a_beta = a_beta,
                          b_beta = b_beta,
                          c_beta = c_beta,
                          d_beta = d_beta,
                          e_beta = e_beta,
                          f_beta = NA,
                          g_beta = NA,
                          pi = ("a_pi + b_pi * SWR + c_pi * T + d_pi * RH + e_pi * W"),
                          a_pi = a_pi,
                          b_pi = b_pi,
                          c_pi = c_pi,
                          d_pi = d_pi,
                          e_pi = e_pi,
                          f_pi = NA,
                          g_pi = NA))

```

# M14: Short Wave Radiation + Temperature + Precipitation + Wind + Precipitation

Lightning strike rate modeled from a normal distribution where mu and sigma are dependent on climate: shortwave radiation,  temperature, relative humidity, wind, and precipitation. 
    S ~ dnorm( mu , sigma ),
    mu <- a_alpha + b_alpha * SWR + c_alpha * T + d_alpha * RH + e_alpha * W + f_alpha * P,
    sigma <- a_beta + b_beta * SWR + c_beta * T + d_beta * RH + e_beta * W + f_beta * P,
    
```{r, echo=FALSE, results='hide', include=FALSE}

#library(rstan)

dat <- list(
  N = nrow(train),
  S = train$strikes,
  SWR = train$swr,
  T = train$tair,
  RH = train$rh,
  W = train$wind,
  P = train$precip
)


# Compile and fit the Stan model
# init_list <- function() {
#   list(a_alpha = 0.1, b_alpha = 0.1, c_alpha = 0.1, d_alpha = 0.1, e_alpha = 0.1, f_alpha = 0.1,
#        a_beta = 0.1, b_beta = 0.1, c_beta = 0.1, d_beta = 0.1, e_beta = 0.1, f_beta = 0.1,
#        a_pi = 0.1, b_pi = 0.1, c_pi = 0.1, d_pi = 0.1, e_pi = 0.1, f_pi = 0.1)
# }

if (!file.exists("compiled_ZIG_weekly_m14.rds")) {
  m14 <- stan(file = "ZIG_swr_tair_rh_wind_precip.stan", 
            data = dat,
            iter = 2000, 
            chains = 4, 
            cores = min(parallel::detectCores(), 4))
    saveRDS(m14, file = "compiled_ZIG_weekly_m14.rds")
    } else {
      m14 <- readRDS("compiled_ZIG_weekly_m14.rds")
    }

```


```{r}

print(m14, pars = c("a_alpha", "b_alpha", "c_alpha", "d_alpha", "e_alpha", "f_alpha", "a_beta", "b_beta", "c_beta", "d_beta", "e_beta", "f_beta", "a_pi", "b_pi", "c_pi", "d_pi", "e_pi", "f_pi"))

traceplot(m14)

parameter_stats <- summary(m14)$summary[, c("mean", "sd")][1:18, 1:2]
print(parameter_stats)

summary_table <- data.frame(
  parameter = rownames(parameter_stats),
  mean = parameter_stats[, "mean"],
  sd = parameter_stats[, "sd"]
)

ggplot(summary_table, aes(x = parameter, y = mean)) +
  geom_point(color = "blue", size = 3) +  # Point for the mean
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), width = 0.2) +  # Error bars for SD
  labs(y = "Parameter Estimate", x = "Parameter") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Parameter Means and SDs from Stan Model")

#predictions 
a_alpha <- summary_table["a_alpha", "mean"]
b_alpha <- summary_table["b_alpha", "mean"]
c_alpha <- summary_table["c_alpha", "mean"]
d_alpha <- summary_table["d_alpha", "mean"]
e_alpha <- summary_table["e_alpha", "mean"]
f_alpha <- summary_table["f_alpha", "mean"]
a_beta <- summary_table["a_beta", "mean"]
b_beta <- summary_table["b_beta", "mean"]
c_beta <- summary_table["c_beta", "mean"]
d_beta <- summary_table["d_beta", "mean"]
e_beta <- summary_table["e_beta", "mean"]
f_beta <- summary_table["f_beta", "mean"]
a_pi <- summary_table["a_pi", "mean"]
b_pi <- summary_table["b_pi", "mean"]
c_pi <- summary_table["c_pi", "mean"]
d_pi <- summary_table["d_pi", "mean"]
e_pi <- summary_table["e_pi", "mean"]
f_pi <- summary_table["f_pi", "mean"]

alpha <-  exp(a_alpha + b_alpha * test$swr + c_alpha * test$tair + d_alpha * test$rh + e_alpha * test$wind + f_alpha * test$precip)
beta <-  exp(a_beta + b_beta * test$swr + c_beta * test$tair + d_beta * test$rh + e_beta * test$wind + f_beta * test$precip)
pi <-  plogis(a_pi + b_pi * test$swr + c_pi * test$tair + d_pi * test$rh + e_pi * test$wind + f_pi * test$precip)

# Simulate zero-inflated gamma predictions
n_predictions <- length(test$strikes)

is_zero <- rbinom(n_predictions, size = 1, prob = pi)  
simStrikes <- ifelse(
  is_zero == 1,
  0,  
  rgamma(n_predictions, alpha, beta)  # Non-zero values follow a gamma distribution
)


hist(simStrikes, breaks=30, main = "predictions")

hist(test$strikes, breaks=30, main = "observations")

ggplot(data = data.frame(observed = test$strikes, predicted = simStrikes, T=test$tair, SWR = test$swr, RH = test$rh, P = test$precip, W = test$wind)) + geom_point(aes(x=SWR, y=predicted, color=P))
ggplot(data = data.frame(observed = test$strikes, predicted = simStrikes, T=test$tair, SWR = test$swr, RH = test$rh, P = test$precip, W = test$wind)) + geom_point(aes(x=SWR, y=observed, color=P))

plot(simStrikes, test$strikes)
abline(a = 0, b = 1, col = "red", lty = 2, lwd = 2)

```


```{r}
# Extract the log-likelihood
log_lik <- extract_log_lik(m14, parameter_name = "log_lik")

# Calculate WAIC
waic_result_m14 <- waic(log_lik)
# Calculate LOO
loo_result_m14 <- loo(log_lik)

# Print WAIC and Loo result
print(waic_result_m14)
print(loo_result_m14)

rmse_m14 <- RMSE(simStrikes, test$strikes)

param_table <- rbind(param_table, data.frame(scale = c("weekly"),
                          distribution = c("ZIG"),
                          model_name = c("m14"),
                          climate_var = c("SWR, T, RH, W, P"),
                          alpha = c("a_alpha + b_alpha * SWR + c_alpha * T + d_alpha * RH + e_alpha * W + f_alpha * P"),
                          a_alpha = a_alpha,
                          b_alpha = b_alpha,
                          c_alpha = c_alpha,
                          d_alpha = d_alpha,
                          e_alpha = e_alpha,
                          f_alpha = f_alpha,
                          g_alpha = NA,
                          beta = c("a_beta + b_beta * SWR + c_beta * T + d_beta * RH + e_beta * W + f_beta * P"),
                          a_beta = a_beta,
                          b_beta = b_beta,
                          c_beta = c_beta,
                          d_beta = d_beta,
                          e_beta = e_beta,
                          f_beta = f_beta,
                          g_beta = NA,
                          pi = c("a_pi + b_pi * SWR + c_pi * T + d_pi * RH + e_pi * W + f_pi * P"),
                          a_pi = a_pi,
                          b_pi = b_pi,
                          c_pi = c_pi,
                          d_pi = d_pi,
                          e_pi = e_pi,
                          f_pi = f_pi,
                          g_pi = NA))
```

# M15: Short Wave Radiation + Temperature + Precipitation + Wind + Precipitation + Surface Pressure

Lightning strike rate modeled from a normal distribution where mu and sigma are dependent on climate: shortwave radiation,  temperature, relative humidity, wind, precipitation, and surface pressure
. 
    S ~ dnorm( mu , sigma ),
    mu <- a_alpha + b_alpha * SWR + c_alpha * T + d_alpha * RH + e_alpha * W + f_alpha * P + g_alpha * SP,
    sigma <- a_beta + b_beta * SWR + c_beta * T + d_beta * RH + e_beta * W + f_beta * P + g_beta * SP,
    
```{r, echo=FALSE, results='hide', include=FALSE}

#library(rstan)

dat <- list(
  N = nrow(train),
  S = train$strikes,
  SWR = train$swr,
  T = train$tair,
  RH = train$rh,
  W = train$wind,
  P = train$precip,
  SP = train$sp
)


# Compile and fit the Stan model
# init_list <- function() {
#   list(a_alpha = 0.1, b_alpha = 0.1, c_alpha = 0.1, d_alpha = 0.1, e_alpha = 0.1, f_alpha = 0.1, g_alpha = 0.1,
#        a_beta = 0.1, b_beta = 0.1, c_beta = 0.1, d_beta = 0.1, e_beta = 0.1, f_beta = 0.1, g_beta = 0.1,
#        a_pi = 0.1, b_pi = 0.1, c_pi = 0.1, d_pi = 0.1, e_pi = 0.1, f_pi = 0.1, g_pi = 0.1)
# }

if (!file.exists("compiled_ZIG_weekly_m15.rds")) {
  m15 <- stan(file = "ZIG_swr_tair_rh_wind_precip_sp.stan", 
            data = dat,
            iter = 2000, 
            chains = 4, 
            cores = min(parallel::detectCores(), 4))
    saveRDS(m15, file = "compiled_ZIG_weekly_m15.rds")
    } else {
      m15 <- readRDS("compiled_ZIG_weekly_m15.rds")
    }

```


```{r}

print(m15, pars = c("a_alpha", "b_alpha", "c_alpha", "d_alpha", "e_alpha", "f_alpha", "g_alpha", "a_beta", "b_beta", "c_beta", "d_beta", "e_beta", "f_beta", "g_beta", "a_pi", "b_pi", "c_pi", "d_pi", "e_pi", "f_pi", "g_pi"))

traceplot(m15)

parameter_stats <- summary(m15)$summary[, c("mean", "sd")][1:21, 1:2]
print(parameter_stats)

summary_table <- data.frame(
  parameter = rownames(parameter_stats),
  mean = parameter_stats[, "mean"],
  sd = parameter_stats[, "sd"]
)

ggplot(summary_table, aes(x = parameter, y = mean)) +
  geom_point(color = "blue", size = 3) +  # Point for the mean
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), width = 0.2) +  # Error bars for SD
  labs(y = "Parameter Estimate", x = "Parameter") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Parameter Means and SDs from Stan Model")

#predictions 
a_alpha <- summary_table["a_alpha", "mean"]
b_alpha <- summary_table["b_alpha", "mean"]
c_alpha <- summary_table["c_alpha", "mean"]
d_alpha <- summary_table["d_alpha", "mean"]
e_alpha <- summary_table["e_alpha", "mean"]
f_alpha <- summary_table["f_alpha", "mean"]
g_alpha <- summary_table["g_alpha", "mean"]
a_beta <- summary_table["a_beta", "mean"]
b_beta <- summary_table["b_beta", "mean"]
c_beta <- summary_table["c_beta", "mean"]
d_beta <- summary_table["d_beta", "mean"]
e_beta <- summary_table["e_beta", "mean"]
f_beta <- summary_table["f_beta", "mean"]
g_beta <- summary_table["g_beta", "mean"]
a_pi<- summary_table["a_pi", "mean"]
b_pi <- summary_table["b_pi", "mean"]
c_pi <- summary_table["c_pi", "mean"]
d_pi <- summary_table["d_pi", "mean"]
e_pi <- summary_table["e_pi", "mean"]
f_pi <- summary_table["f_pi", "mean"]
g_pi <- summary_table["g_pi", "mean"]

alpha <-  exp(a_alpha + b_alpha * test$swr + c_alpha * test$tair + d_alpha * test$rh + e_alpha * test$wind + f_alpha * test$precip + g_alpha * test$sp)
beta <-  exp(a_beta + b_beta * test$swr + c_beta * test$tair + d_beta * test$rh + e_beta * test$wind + f_beta * test$precip + g_beta * test$sp)
pi <-  plogis(a_pi + b_pi * test$swr + c_pi * test$tair + d_pi * test$rh + e_pi * test$wind + f_pi * test$precip + g_pi * test$sp)

# Simulate zero-inflated gamma predictions
n_predictions <- length(test$strikes)

is_zero <- rbinom(n_predictions, size = 1, prob = pi)  
simStrikes <- ifelse(
  is_zero == 1,
  0,  
  rgamma(n_predictions, alpha, beta)  # Non-zero values follow a gamma distribution
)


hist(simStrikes, breaks=30, main = "predictions")

hist(test$strikes, breaks=30, main = "observations")

ggplot(data = data.frame(observed = test$strikes, predicted = simStrikes, T=test$tair, SWR = test$swr, RH = test$rh, P = test$precip, W = test$wind, SP = test$sp)) + geom_point(aes(x=SWR, y=predicted, color=SP))
ggplot(data = data.frame(observed = test$strikes, predicted = simStrikes, T=test$tair, SWR = test$swr, RH = test$rh, P = test$precip, W = test$wind, SP = test$sp)) + geom_point(aes(x=SWR, y=observed, color=SP))

plot(simStrikes, test$strikes)
abline(a = 0, b = 1, col = "red", lty = 2, lwd = 2)

```

```{r}
# Extract the log-likelihood
log_lik <- extract_log_lik(m15, parameter_name = "log_lik")

# Calculate WAIC
waic_result_m15 <- waic(log_lik)
# Calculate LOO
loo_result_m15 <- loo(log_lik)

# Print WAIC and Loo result
print(waic_result_m15)
print(loo_result_m15)

rmse_m15 <- RMSE(simStrikes, test$strikes)

param_table <- rbind(param_table, data.frame(scale = c("weekly"),
                          distribution = c("ZIG"),
                          model_name = c("m15"),
                          climate_var = c("SWR, T, RH, W, P, SP"),
                          alpha = c("a_alpha + b_alpha * SWR + c_alpha * T + d_alpha * RH + e_alpha * W + f_alpha * P + g_alpha * SP"),
                          a_alpha = a_alpha,
                          b_alpha = b_alpha,
                          c_alpha = c_alpha,
                          d_alpha = d_alpha,
                          e_alpha = e_alpha,
                          f_alpha = f_alpha,
                          g_alpha = g_alpha,
                          beta = c("a_beta + b_beta * SWR + c_beta * T + d_beta * RH + e_beta * W + f_beta * P + g_beta * SP"),
                          a_beta = a_beta,
                          b_beta = b_beta,
                          c_beta = c_beta,
                          d_beta = d_beta,
                          e_beta = e_beta,
                          f_beta = f_beta,
                          g_beta = g_beta,
                          pi = c("a_pi + b_pi * SWR + c_pi * T + d_pi * RH + e_pi * W + f_pi * P + g_pi * SP"),
                          a_pi = a_pi,
                          b_pi = b_pi,
                          c_pi = c_pi,
                          d_pi = d_pi,
                          e_pi = e_pi,
                          f_pi = f_pi,
                          g_pi = g_pi))

write.csv(param_table, "/raid/cuden/data/gamma_weekly_model_parameters.csv")

```

```{r}
colnames <- c("model", "climate_var", "elpd_loo", "p_loo", "looic", "elpd_waic", "p_waic", "waic")
model_comparison <- data.frame(rbind(c("m0.1", "normal", loo_result_m0.1$estimates[1,1], loo_result_m0.1$estimates[2,1], loo_result_m0.1$estimate[3,1], waic_result_m0.1$estimates[1,1], waic_result_m0.1$estimates[2,1], waic_result_m0.1$estimates[3,1], rmse_m0.1),
      c("m0.3", "ZIG", loo_result_m0.3$estimates[1,1], loo_result_m0.3$estimates[2,1], loo_result_m0.3$estimate[3,1], waic_result_m0.3$estimates[1,1], waic_result_m2$estimates[2,1], waic_result_m2$estimates[3,1], rmse_m0.3),
  c("m1", "CAPE", loo_result_m1$estimates[1,1], loo_result_m1$estimates[2,1], loo_result_m1$estimate[3,1], waic_result_m1$estimates[1,1], waic_result_m1$estimates[2,1], waic_result_m1$estimates[3,1], rmse_m1),
  c("m2", "CXP", loo_result_m2$estimates[1,1], loo_result_m2$estimates[2,1], loo_result_m2$estimate[3,1], waic_result_m2$estimates[1,1], waic_result_m2$estimates[2,1], waic_result_m2$estimates[3,1], rmse_m2),
  c("m3", "rh", loo_result_m3$estimates[1,1], loo_result_m3$estimates[2,1], loo_result_m3$estimate[3,1], waic_result_m3$estimates[1,1], waic_result_m3$estimates[2,1], waic_result_m3$estimates[3,1], rmse_m3),
  c("m4", "swr", loo_result_m4$estimates[1,1], loo_result_m4$estimates[2,1], loo_result_m4$estimate[3,1], waic_result_m4$estimates[1,1], waic_result_m4$estimates[2,1], waic_result_m4$estimates[3,1], rmse_m4),
  c("m5", "tair", loo_result_m5$estimates[1,1], loo_result_m5$estimates[2,1], loo_result_m5$estimate[3,1], waic_result_m5$estimates[1,1], waic_result_m5$estimates[2,1], waic_result_m5$estimates[3,1], rmse_m5),
  c("m6", "sp", loo_result_m6$estimates[1,1], loo_result_m6$estimates[2,1], loo_result_m6$estimate[3,1], waic_result_m6$estimates[1,1], waic_result_m6$estimates[2,1], waic_result_m6$estimates[3,1], rmse_m6),
  c("m7", "precip", loo_result_m7$estimates[1,1], loo_result_m7$estimates[2,1], loo_result_m7$estimate[3,1], waic_result_m7$estimates[1,1], waic_result_m7$estimates[2,1], waic_result_m7$estimates[3,1], rmse_m7),
  c("m8", "wind", loo_result_m8$estimates[1,1], loo_result_m8$estimates[2,1], loo_result_m8$estimates[3,1], waic_result_m8$estimates[1,1], waic_result_m8$estimates[2,1], waic_result_m8$estimates[3,1], rmse_m8), 
  c("m9", "SWR+T", loo_result_m9$estimates[1,1], loo_result_m9$estimates[2,1], loo_result_m9$estimates[3,1], waic_result_m9$estimates[1,1], waic_result_m9$estimates[2,1], waic_result_m9$estimates[3,1], rmse_m9),
  c("m12", "SWR+T+RH", loo_result_m12$estimates[1,1], loo_result_m12$estimates[2,1], loo_result_m12$estimates[3,1], waic_result_m12$estimates[1,1], waic_result_m12$estimates[2,1], waic_result_m12$estimates[3,1], rmse_m12),
  c("m13", "SWR+T+RH+W", loo_result_m13$estimates[1,1], loo_result_m13$estimates[2,1], loo_result_m13$estimates[3,1], waic_result_m13$estimates[1,1], waic_result_m13$estimates[2,1], waic_result_m13$estimates[3,1], rmse_m13),
  c("m14", "SWR+T+RH+W+P", loo_result_m14$estimates[1,1], loo_result_m14$estimates[2,1], loo_result_m14$estimates[3,1], waic_result_m14$estimates[1,1], waic_result_m14$estimates[2,1], waic_result_m14$estimates[3,1], rmse_m14),
  c("m15", "SWR+T+RH+W+P+SP", loo_result_m15$estimates[1,1], loo_result_m15$estimates[2,1], loo_result_m15$estimates[3,1], waic_result_m15$estimates[1,1], waic_result_m15$estimates[2,1], waic_result_m15$estimates[3,1], rmse_m15)))

#model_comparison[,3:8] <- as.numeric(model_comparison[,3:8])

colnames(model_comparison) <- colnames

#print(arrange(model_comparison, waic))

write.csv(model_comparison, "/raid/cuden/data/gamma_weekly_model_comparison.csv")

model_comparison
```
M15, which includes all 6 surface level climate variables, performs best, with the lowest WAIC score and highest loo score. 

```{r}

ggplot(data = data.frame(
  observed = test$strikes, predicted = simStrikes, 
  T=test$tair, SWR = test$swr, RH = test$rh, P = test$precip, W = test$wind, SP = test$sp)) + 
  geom_point(aes(x=T, y=predicted, color=SP)) + 
  scale_color_continuous(high="green", low="orange")

ggplot(data = data.frame(
  observed = test$strikes, predicted = simStrikes, 
  T=test$tair, SWR = test$swr, RH = test$rh, P = test$precip, W = test$wind, SP = test$sp)) + 
  geom_point(aes(x=T, y=observed, color=SP)) + 
  scale_color_continuous(high="green", low="orange")

```










