---
title: "Chen et al. 2021 NE Parameterizations"
author: "Charlotte Uden"
date: "2025-02-09"
output: html_document
---

#Applying models from Chen 2021 to the Northeast. This script parameterizes four models from this paper:

#-- Power law: power law regression: FR = a(CAPE × Precip)^b, with the parameters a and b derived from the 
#optimization at the log-log scale (that is, a linear regression between logarithmic values of flash 
#rate and CAPE × Precip were performed in practice). Similar to Romps et al.36, this model assumed that 
#the lightning flash rate depends on the product of CAPE and Precip. The coefficient b allows nonlinear 
#impacts to be represented in this form. 
#-- Power law (linear opt): is a variation of the power law model, but with the least squares optimization 
#obtained without a logarithmic transformation. 
#-- scale: the original scaling approach proposed by Romps et al.36: FR = a(CAPE × Precip). 
#-- linear: we slightly modified the scaling approach by allowing a negative intercept: FR = a(CAPE × Precip) + b 
#-that is, positive flash rate was simulated only when CAPE × Precip was greater than a threshold. 
#-- a non-parametric: a non-parametric regression model that includes a lookup table of flash rate as 
#a function of CAPE × Precip within the range of contemporary observations. The modelled flash 
#rate value in each bin was derived using the arithmetic means of the observed flash rate in all 
#grid cells associated with the corresponding ranges of CAPE × Precip. A linear extrapolation was used to 
#extend the model to cover CAPE × Precip values outside of the observational range.

```{r, echo=FALSE, include=FALSE}
library(rstan)
library(loo)
library(dplyr)
library(stats)
library(caret)
library(tidyr)
library(gridExtra)
library(mapdata)
```
1. Read in the data


```{r, echo=FALSE}
df <- read.csv("/raid/cuden/data/era5_vaisalaLightning_monthlySummerSummaries_2005-2010_NEclip.csv")[,2:16]
colnames(df) <- c("lon", "lat", "year", "strikes", "cape", "precip", "cxp", "d2m", "tair", "wind", "swr", "sp", "rh", "tmin", "tmax")
hist(df$strikes, breaks = 30)
```

2. Split into training and testing data
```{r, echo=FALSE}
## 80% of the sample size
sample_size <- floor(0.8 * nrow(df))

## set the seed to make partition reproducible
set.seed(123)
train_index <- sample(seq_len(nrow(df)), size = sample_size)

train <- df[train_index, ]
test <- df[-train_index, ]

```


3. Standardize predictor variables - chen et al do not do this! skip this code chunk 
```{r, echo=FALSE}
# #save mean and sd:
# #-----
# mu_cxp <- mean(train$cxp)
# sigma_cxp <- sd(train$cxp)
# mu_cape <- mean(train$cape)
# sigma_cape <- sd(train$cape)
# mu_precip <- mean(train$precip)
# sigma_precip <- sd(train$precip)
# mu_tair <- mean(train$tair)
# sigma_tair <- sd(train$tair)
# mu_wind <- mean(train$wind)
# sigma_wind <- sd(train$wind)
# mu_swr <- mean(train$swr)
# sigma_swr <- sd(train$swr)
# mu_sp <- mean(train$sp)
# sigma_sp <- sd(train$sp)
# mu_rh <- mean(train$rh)
# sigma_rh <- sd(train$rh)
# #-----
# 
# train <- train %>%
#   mutate(cape = (cape-mu_cape)/sigma_cape) %>%
#   mutate(precip = (precip-mu_precip)/sigma_precip) %>%
#   mutate(cxp = (cxp-mu_cxp)/sigma_cxp) %>%
#   mutate(tair = (tair-mu_tair)/sigma_tair) %>%
#   mutate(wind = (wind-mu_wind)/sigma_wind) %>%
#   mutate(swr = (swr-mu_swr)/sigma_swr) %>%
#   mutate(sp = (sp-mu_sp)/sigma_sp) %>%
#   mutate(rh = (rh-mu_rh)/sigma_rh) 
# 
# train <- train[,c("strikes", "cape", "precip", "cxp", "tair", "wind", "swr", "sp", "rh")]
# head(train)
# 
# plot(train)
# 
# #use the min and max values from the train set to normalize the test set 
# test <- test %>%
#   mutate(cape = (cape-mu_cape)/sigma_cape) %>%
#   mutate(precip = (precip-mu_precip)/sigma_precip) %>%
#   mutate(cxp = (cxp-mu_cxp)/sigma_cxp) %>%
#   mutate(tair = (tair-mu_tair)/sigma_tair) %>%
#   mutate(wind = (wind-mu_wind)/sigma_wind) %>%
#   mutate(swr = (swr-mu_swr)/sigma_swr) %>%
#   mutate(sp = (sp-mu_sp)/sigma_sp) %>%
#   mutate(rh = (rh-mu_rh)/sigma_rh) %>%
#   mutate(lon = lon) %>%
#   mutate(lat = lat)
# 
# test <- test[,c("lon", "lat", "strikes", "cape", "precip", "cxp", "tair", "wind", "swr", "sp", "rh")]

```

```{r}
train <- train[,c("lon", "lat", "strikes", "cxp")]
head(train)
test <- test[,c("lon", "lat", "strikes", "cxp")]
head(test)
```


Look at the data
```{r}
x <- train$cxp
y <- train$strikes

x_test <- test$cxp
y_test <- test$strikes

plot(x, y)
```

#-- Power Law: FR = a(CAPE × Precip)^b -- 

```{r}

# Transform data
log_x <- log(x) # standardize CAPE * precipitation
log_y <- log(y + 0.0001) # lightning strike rate - add a small consgtant so that log(y) doesn not produce - inf.

# Fit the power law model in log-log space
pl <- nls(log_y ~ log(a) + b * log_x, 
          start = list(a = 1, b = 1))

```

### Predicitons
```{r}

# Predict flash rate using the fitted power law
y_pred_pl <- coef(pl)["a"] * x_test^coef(pl)["b"]

# Plot results
df_results <- as.data.frame(cbind(x_test=x_test, y_test=y_test, y_pred_pl = y_pred_pl))
ggplot(df_results) + geom_point(aes(x=x_test, y=y_test)) + geom_line(aes(x=x_test, y = y_pred_pl, col="red"))

```





#-- Power law (linear opt) --
```{r}
pl_op <- nls(y ~ a * x^b, start = list(a = 1, b = 1))
summary(pl_op)

a_pl_op <- summary(pl_op)$coefficients[1,1]
b_pl_op <- summary(pl_op)$coefficients[2,1]
```

### Predicitons
```{r}
# Predict flash rate using the fitted power law
y_pred_pl_op <- a_pl_op * x_test^b_pl_op

# Plot results
df_results <- as.data.frame(cbind(df_results, y_pred_pl_op = y_pred_pl_op))
ggplot(df_results) + geom_point(aes(x=x_test, y=y_test)) + geom_line(aes(x=x_test, y = y_pred_pl_op, col="red"))
```


#-- scale: FR = a(CAPE × Precip) --
```{r}
# Fit the scale model (y = a * x)
sc <- nls(y ~ a * x, 
                   start = list(a = 1))  # Set reasonable starting value
summary(sc)

# Extract parameter
a_sc <- coef(sc)["a"]
```

```{r}
# Make predictions on test set
y_pred_sc <- a_sc * x_test  # Since b is fixed at 1, we use y = a * x

df_results <- data.frame(cbind(df_results, y_pred_sc = y_pred_sc))
ggplot(df_results) + geom_point(aes(x=x_test, y=y_test)) + geom_line(aes(x=x_test, y = y_pred_sc, col="red"))
```

#-- linear: FR = a(CAPE × Precip) + b -- 
```{r}
# Fit the scale model (y = a * x + b)
linear <- nls(y ~ a * x + b, 
                   start = list(a = 1, b = 0))  # Set reasonable starting value
summary(linear)

# Extract parameter
a_linear <- coef(linear)["a"]
b_linear<- coef(linear)["b"]
```

#predictions
```{r}
y_pred_linear <- a_linear * x_test + b_linear  # Apply the fitted equation

df_results <- data.frame(cbind(df_results, y_pred_linear = y_pred_linear))
ggplot(df_results) + geom_point(aes(x=x_test, y=y_test)) + geom_line(aes(x=x_test, y = y_pred_linear, col="red"))

```
#-- Non-parametric --
#modelled flash rate value in each bin was derived using the arithmetic means of the observed flash rate in all 
#grid cells associated with the corresponding ranges of CAPE × Precip.
#based on the script Chen shared, there are 50 bins. 

```{r}

# Define binning parameters from the TRAINING SET ONLY
max_cxp <- max(train$cxp)
min_cxp <- min(train$cxp)
nbins <- 50
bin_size <- (max_cxp - min_cxp) / nbins
cxp_bin_min <- seq(min_cxp, max_cxp, by = bin_size)

# Bin the TRAINING data
train <- train %>%
  mutate(bin_index = cut(cxp, breaks = cxp_bin_min, include.lowest = TRUE, labels = FALSE))

#Compute mean strike rate per bin using TRAINING data
np_table <- train %>%
  group_by(bin_index) %>%
  summarise(strike_bin_means = mean(strikes, na.rm = TRUE)) %>%
  mutate(cxp_bin_min = cxp_bin_min[bin_index],
         cxp_bin_max = cxp_bin_min + bin_size)

head(np_table)
```

```{r}
# Make predictions for the TEST SET
df_results <- df_results %>%
  mutate(bin_index = cut(x_test, breaks = cxp_bin_min, include.lowest = TRUE, labels = FALSE)) %>%
  left_join(np_table, by = "bin_index")

# Ensure missing bins are handled (e.g., if a test sample falls in an unseen bin)
df_results$strike_bin_means[is.na(df_results$strike_bin_means)] <- mean(train$strikes, na.rm = TRUE)  # Fill missing bins with global mean

head(df_results)

# Final predictions
y_pred_np <- df_results$strike_bin_means
```
```{r}
df_results <- data.frame(cbind(df_results, y_pred_np = y_pred_np))
ggplot(df_results) + geom_point(aes(x=x_test, y=y_test)) + geom_line(aes(x=x_test, y = y_pred_np, col="red"))

```


# Next get mean of all predictions and calculate RMSE for comparison with bayes models
Chen eta l. 2021 exclude the linear model. possible reasons include: it does not properly enforce the physical requirement that lightning should be near-zero when CAPE × Precip is near zero. The power law models and the non-parametric approach provide better statistical and physical performance.The power law and non-parametric models adapt well to future climate scenarios because they capture how lightning frequency changes non-linearly with warming.
A linear model may overestimate lightning at low CAPE × Precip and underestimate it at high CAPE × Precip, leading to systematic errors.

```{r}
# Compute mean prediction across all models
y_pred_mean <- (y_pred_pl + y_pred_pl_op + y_pred_sc + y_pred_np) / 4

df_results <- data.frame(cbind(df_results, y_pred_mean = y_pred_mean))
ggplot(df_results) + geom_point(aes(x=x_test, y=y_test)) + geom_line(aes(x=x_test, y = y_pred_mean, col="red"))

```

```{r}
library(Metrics)  # For RMSE calculation

# Compute RMSE for each model
rmse_pl <- rmse(test$strikes, y_pred_pl)  # Power Law
rmse_pl_linear_opt <- rmse(test$strikes, y_pred_pl_op)  # Power Law (Linear Opt)
rmse_scale <- rmse(test$strikes, y_pred_sc)  # Scale Model
rmse_np <- rmse(test$strikes, y_pred_np)  # Non-Parametric Model
rmse_mean <- rmse(test$strikes, y_pred_mean)  # Mean of all models

# Print results
cat("RMSE for Power Law Model:", rmse_pl, "\n")
cat("RMSE for Power Law (Linear Opt) Model:", rmse_pl_linear_opt, "\n")
cat("RMSE for Scale Model:", rmse_scale, "\n")
cat("RMSE for Non-Parametric Model:", rmse_np, "\n")
cat("RMSE for Mean Model (Ensemble):", rmse_mean, "\n")


```





